### 2022-10-04 paper/paper.tex
Basic paper outline and structure

%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall, review, screen]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2022}
\acmYear{2022}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%% Packages
\usepackage{xcolor}
\usepackage{color}
\usepackage{listings}

%% CUSTOM PACKAGES - NON-ACCEPTED
\usepackage{todonotes}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{The Name of the Title Is Hope}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ben Trovato}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \country{USA}
  \postcode{43017-6221}
}

\author{Lars Th{\o}rv{\"a}ld}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Hekla}
  \country{Iceland}}
\email{larst@affiliation.org}

\author{Valerie B\'eranger}
\affiliation{%
  \institution{Inria Paris-Rocquencourt}
  \city{Rocquencourt}
  \country{France}
}

\author{Aparna Patel}
\affiliation{%
 \institution{Rajiv Gandhi University}
 \streetaddress{Rono-Hills}
 \city{Doimukh}
 \state{Arunachal Pradesh}
 \country{India}}

\author{Huifen Chan}
\affiliation{%
  \institution{Tsinghua University}
  \streetaddress{30 Shuangqing Rd}
  \city{Haidian Qu}
  \state{Beijing Shi}
  \country{China}}

\author{Charles Palmer}
\affiliation{%
  \institution{Palmer Research Laboratories}
  \streetaddress{8600 Datapoint Drive}
  \city{San Antonio}
  \state{Texas}
  \country{USA}
  \postcode{78229}}
\email{cpalmer@prl.com}

\author{John Smith}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Hekla}
  \country{Iceland}}
\email{jsmith@affiliation.org}

\author{Julius P. Kumquat}
\affiliation{%
  \institution{The Kumquat Consortium}
  \city{New York}
  \country{USA}}
\email{jpkumquat@consortium.net}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
	\todo[inline]{Write useful abstract}
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{datasets, neural networks, gaze detection, text tagging}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
\begin{teaserfigure}
  \includegraphics[width=\textwidth]{./samples/sampleteaser}
  \caption{Seattle Mariners at Spring Training, 2010.}
  \Description{Enjoying the baseball game from the third-base
  seats. Ichiro Suzuki preparing to bat.}
  \label{fig:teaser}
\end{teaserfigure}

\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
\subsection{Context}
Using Refinement Types allows us to define contracts for function types that define pre-conditions on the function arguments and post-conditions on it's return values. These contracts can then be checked statically at compile time. One implementation of this is LiquidHaskell. When combining function calls, it can be verified that the post-condition of the inner function will also fulfill all pre-conditions of the outher function. Listing \ref{lst:liquidhaskell} shows an example of this.

\begin{lstlisting}[language=Haskell,caption={Contracts in LiquidHaskell}, label={lst:liquidhaskell}]
-- Refinement type that describes a non-empty list
{-@ type NonEmpty a = {v:[a] | len v > 0 } @-}

-- Define a head function that only works on non-emtpy lists
{-@ head :: NonEmpty Int -> Int @-}
head :: [Int] -> Int
head (x:_) = x

{-@ ensureNonEmpty :: [Int] -> NonEmpty Int -}
ensureNonEmpty :: [Int] -> [Int]
ensureNonEmpty [] = [0]
ensureNonEmpty x = x

i1 :: Int
i1 = head []			-- Not allowed
i1 = head (ensureNonEmpty []) 	-- Allowed
\end{lstlisting}

Using contracts, the programmer can clearly define the expectations and assurances that are made to the functions they write. Ensuring their fulfillment before code execution can have a great impact on a programs correctness and stability.

\subsection{Motivation}

\subsection{Problem Definition}
\subsection{Impact}
\subsection{Approach}
\subsection{Contributions}
\section{Related Work}
\section{Background}
\section{Implementation}
\section{Evaluation}
\section{Conclusion}
\section{Acknowledgments}

\end{document}
\endinput
%%
%% End of file `sample-acmsmall-conf.tex'.

### 2022-10-06 paper/paper.tex
Worked on motivating example
Setup repository structure with Haskell code

@@
 
+%% Custom Definitions
+\graphicspath{{graph/}}
+\definecolor{ForestGreen}{HTML}{009B55}
+
+\lstset{inputpath=code_examples/src/Examples}
+\lstset{
+  frame=none,
+  xleftmargin=2pt,
+  stepnumber=1,
+  numbers=left,
+  numbersep=5pt,
+  numberstyle=\ttfamily\tiny\color[gray]{0.3},
+  belowcaptionskip=\bigskipamount,
+  escapeinside={*'}{'*},
+  language=haskell,
+  tabsize=2,
+  emphstyle={\bf},
+  commentstyle=\color{ForestGreen},
+  stringstyle=\mdseries\rmfamily,
+  showspaces=false,
+  keywordstyle=\bfseries\rmfamily,
+  columns=flexible,
+  basicstyle=\small\sffamily,
+  showstringspaces=false,
+  morecomment=[l]\%,
+}
+
 %% CUSTOM PACKAGES - NON-ACCEPTED
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{The Name of the Title Is Hope}
+\title{LayeredTypes}
+
+%\title[LayeredTypes]{Layered Types -- An approach to arbitrarily combine LiquidTypes in a structured Manner}
 
@@
 %% used to denote shared contribution to the research.
-\author{Ben Trovato}
-\authornote{Both authors contributed equally to this research.}
-\email{trovato@corporation.com}
+\author{Lukas Abelt}
+\email{l.abelt@luabelt.de}
 \orcid{1234-5678-9012}
-\author{G.K.M. Tobin}
-\authornotemark[1]
-\email{webmaster@marysville-ohio.com}
 \affiliation{%
-  \institution{Institute for Clarity in Documentation}
-  \streetaddress{P.O. Box 1212}
-  \city{Dublin}
-  \state{Ohio}
-  \country{USA}
-  \postcode{43017-6221}
+  \institution{Saarland University}
+  \streetaddress{XXX}
+  \city{Saarbrücken}
+  \state{Saarland}
+  \country{Germany}
+  \postcode{66113}
 }
-
-\author{Lars Th{\o}rv{\"a}ld}
-\affiliation{%
-  \institution{The Th{\o}rv{\"a}ld Group}
-  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-  \city{Hekla}
-  \country{Iceland}}
-\email{larst@affiliation.org}
-
-\author{Valerie B\'eranger}
-\affiliation{%
-  \institution{Inria Paris-Rocquencourt}
-  \city{Rocquencourt}
-  \country{France}
-}
-
-\author{Aparna Patel}
-\affiliation{%
- \institution{Rajiv Gandhi University}
- \streetaddress{Rono-Hills}
- \city{Doimukh}
- \state{Arunachal Pradesh}
- \country{India}}
-
-\author{Huifen Chan}
-\affiliation{%
-  \institution{Tsinghua University}
-  \streetaddress{30 Shuangqing Rd}
-  \city{Haidian Qu}
-  \state{Beijing Shi}
-  \country{China}}
-
-\author{Charles Palmer}
-\affiliation{%
-  \institution{Palmer Research Laboratories}
-  \streetaddress{8600 Datapoint Drive}
-  \city{San Antonio}
-  \state{Texas}
-  \country{USA}
-  \postcode{78229}}
-\email{cpalmer@prl.com}
-
-\author{John Smith}
-\affiliation{%
-  \institution{The Th{\o}rv{\"a}ld Group}
-  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-  \city{Hekla}
-  \country{Iceland}}
-\email{jsmith@affiliation.org}
-
-\author{Julius P. Kumquat}
-\affiliation{%
-  \institution{The Kumquat Consortium}
-  \city{New York}
-  \country{USA}}
-\email{jpkumquat@consortium.net}
 
@@
 %% of authors' names for this purpose.
-\renewcommand{\shortauthors}{Trovato et al.}
 
@@
 %% Please copy and paste the code instead of the example below.
-%%
+%% TODO
 \begin{CCSXML}
@@
 \end{CCSXML}
-
-\ccsdesc[500]{Computer systems organization~Embedded systems}
-\ccsdesc[300]{Computer systems organization~Redundancy}
-\ccsdesc{Computer systems organization~Robotics}
-\ccsdesc[100]{Networks~Network reliability}
+%TODO
+%\ccsdesc[500]{Computer systems organization~Embedded systems}
+%\ccsdesc[300]{Computer systems organization~Redundancy}
+%\ccsdesc{Computer systems organization~Robotics}
+%\ccsdesc[100]{Networks~Network reliability}
 
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{datasets, neural networks, gaze detection, text tagging}
+%TODO: Keywords
+\keywords{}
 %% A "teaser" image appears between the author and affiliation
@@
 \begin{teaserfigure}
-  \includegraphics[width=\textwidth]{./samples/sampleteaser}
-  \caption{Seattle Mariners at Spring Training, 2010.}
-  \Description{Enjoying the baseball game from the third-base
-  seats. Ichiro Suzuki preparing to bat.}
+  \includegraphics[width=\textwidth]{layered-shots.jpg}
   \label{fig:teaser}
@@
 
-\begin{lstlisting}[language=Haskell,caption={Contracts in LiquidHaskell}, label={lst:liquidhaskell}]
--- Refinement type that describes a non-empty list
-{-@ type NonEmpty a = {v:[a] | len v > 0 } @-}
+\lstinputlisting[language=Haskell,caption={Contracts in LiquidHaskell}, label={lst:liquidhaskell}]{BasicLT.hs}
 
--- Define a head function that only works on non-emtpy lists
-{-@ head :: NonEmpty Int -> Int @-}
-head :: [Int] -> Int
-head (x:_) = x
-
-{-@ ensureNonEmpty :: [Int] -> NonEmpty Int -}
-ensureNonEmpty :: [Int] -> [Int]
-ensureNonEmpty [] = [0]
-ensureNonEmpty x = x
-
-i1 :: Int
-i1 = head []			-- Not allowed
-i1 = head (ensureNonEmpty []) 	-- Allowed
-\end{lstlisting}
-
-Using contracts, the programmer can clearly define the expectations and assurances that are made to the functions they write.
+Using these contracts, the programmer can clearly define the expectations and assurances that are made to the functions they write.
 Ensuring their fulfillment before code execution can have a great impact on a programs correctness and stability.
@@
 \subsection{Motivation}
+One challenge current implementations of LiquidTypes face is that is is not always possible to trivially combine different constraints.
+Assume that in LiquidHaskell we have two refinement types for non-empty and ordered lists respectively.
+Freely combining these two types is not possible, so we rather have to create a new, separate, refinement type for this as shown in Listing \ref{lst:combined_refs}.
+
+\lstinputlisting[language=Haskell,caption={Combining Refinement Types},label={lst:combined_refs}]{CombinedLT.hs}
+
+As we have to combine the two constraints into one refinement type they are also always checked together.
+While this seems like a non-issue for simple functions as the ones presented so far, this may hinder the programmers ability to identify which pre-conditions are not ensured when writing functions that interface with already existing code.
+Assume that some developer writes a function whose result he wants to pass to the \texttt{maxVal} function defined in Listing \ref{lst:combined_refs}.
 

### 2022-10-06 paper/code_examples/src/Main.hs

@@ -1,0 +1,29 @@
+main :: IO ()
+
+-- Refinement type that describes a non-empty list
+{-@ type NonEmpty a = {v:[a] | len v > 0 } @-}
+
+-- Define a head function that only works on non-emtpy lists
+{-@ myHead :: NonEmpty Int -> Int @-}
+myHead :: [Int] -> Int
+myHead (x:_) = x
+
+{-@ ensureNonEmpty :: [Int] -> NonEmpty Int @-}
+ensureNonEmpty :: [Int] -> [Int]
+ensureNonEmpty [] = [0]
+ensureNonEmpty x = x
+
+i1 :: Int
+-- i1 = myHead []                    -- Not allowed
+i1 = myHead (ensureNonEmpty [])   -- Allowed
+
+{-@ type Ordered a = [a]<{\x v -> x <= v}> @-}
+-- Not possible to arbitrarly combine NonEmpty and Ordered into one refinement type
+{-@ type NonEmptyOrdered a = {v:[a]<{\x v -> x >= v}> | len v > 0 }@-}
+
+-- Extracting the maximum value from a non-empty and ordered list is equivalent to head
+{-@ maxVal :: NonEmptyOrdered Int -> Int @-}
+maxVal :: [Int] -> Int
+maxVal (x:_) = x
+
+main = putStrLn "Hello, Worldi!"

### 2022-10-07 paper/paper.tex

@@
 
-\lstinputlisting[language=Haskell,caption={Contracts in LiquidHaskell}, label={lst:liquidhaskell}]{BasicLT.hs}
+\lstinputlisting[caption={Contracts in LiquidHaskell}, label={lst:liquidhaskell}]{BasicLT.hs}
 
@@
 
-\lstinputlisting[language=Haskell,caption={Combining Refinement Types},label={lst:combined_refs}]{CombinedLT.hs}
+\lstinputlisting[caption={Combining Refinement Types},label={lst:combined_refs}]{CombinedLT.hs}
 

### 2022-10-07 paper/paper.tex
Rephrased context in preparation for motivation

@@
 \title{LayeredTypes}
-
-%\title[LayeredTypes]{Layered Types -- An approach to arbitrarily combine LiquidTypes in a structured Manner}
+%TODO: Proper Title
+%\title[LayeredTypes]{Layered Types -- A structured approach to arbitrarily combine Liquid Types}
 
@@
 \affiliation{%
-  \institution{Saarland University}
+  \institution{LASIGE}
   \streetaddress{XXX}
-  \city{Saarbrücken}
-  \state{Saarland}
-  \country{Germany}
-  \postcode{66113}
+  \city{Lisbon}
+  \state{}
+  \country{Portugal}
+  \postcode{}
 }
@@
 \begin{abstract}
-	\todo[inline]{Write useful abstract}
+	\todo[inline]{TODO: Abstract}
 \end{abstract}
@@
   \label{fig:teaser}
+  \caption{Layers in Liquids}
 \end{teaserfigure}
@@
 \subsection{Context}
-Using Refinement Types allows us to define contracts for function types that define pre-conditions on the function arguments and post-conditions on it's return values.
-These contracts can then be checked statically at compile time.
-One implementation of this is LiquidHaskell.
-When combining function calls, it can be verified that the post-condition of the inner function will also fulfill all pre-conditions of the outher function.
-Listing \ref{lst:liquidhaskell} shows an example of this.
+Modern program languages such as C++, Haskell, Rust or Java use type systems to ensure certain promises about the correct behavior of an implementation.
+In essence, the type system allows the programer to specify what inputs to accept for a program or function and what outputs it will produce.
+While these guarantees already prevent a multitude of runtime errors, they alone cannot prevent e.g.
+divisions by yero or out of range accesses in arrays.
 
-\lstinputlisting[caption={Contracts in LiquidHaskell}, label={lst:liquidhaskell}]{BasicLT.hs}
+One approach to bridge this gap is the introduction of \textit{Refinement Types}, also referred to as \textit{Liquid Types}.
+These allow us to annotate the traditional type system with additional logical predicates in order to ensure semantic properties, which can be checked statically at compile time.
+Using these techniques also allows us to define pre- and post-conditions on functions' arguments and return values.
+By properly defining these contracts a developer can clearly communicate and moreover enforce how other developers' code is allowed to interface with their implementations.
+Examples for constraints might include any of the following:
 
-Using these contracts, the programmer can clearly define the expectations and assurances that are made to the functions they write.
-Ensuring their fulfillment before code execution can have a great impact on a programs correctness and stability.
+\begin{itemize}
+	\item Ensuring values are in certain ranges
+	\item Ensuring that a list is non-empty
+	\item Ensuring that data structures certain to specific semantic properties e.g. that they are sorted
+\end{itemize}
+
+All in all, refinement types are valuable building blocks to creating reliable software systems.
+Using them can minimize the risk of any unhandled run-time errors occuring which might lead to undesired or even undefined behavior.
 
 \subsection{Motivation}
-One challenge current implementations of LiquidTypes face is that is is not always possible to trivially combine different constraints.
-Assume that in LiquidHaskell we have two refinement types for non-empty and ordered lists respectively.
-Freely combining these two types is not possible, so we rather have to create a new, separate, refinement type for this as shown in Listing \ref{lst:combined_refs}.
-
-\lstinputlisting[caption={Combining Refinement Types},label={lst:combined_refs}]{CombinedLT.hs}
-
-As we have to combine the two constraints into one refinement type they are also always checked together.
-While this seems like a non-issue for simple functions as the ones presented so far, this may hinder the programmers ability to identify which pre-conditions are not ensured when writing functions that interface with already existing code.
-Assume that some developer writes a function whose result he wants to pass to the \texttt{maxVal} function defined in Listing \ref{lst:combined_refs}.
 

### 2022-10-10 paper/paper.tex
Finished motivation subsection

@@
 
+While Refinement Types already vastly imrpove traditional type systems, current implementations still expose some drawbacks in their practical usage.
+One of these challenges stems from the fact that during the static analysis, each refinement is treated as a single monolithic requirement that needs to be fulfilled.
+However in reality, a refinement might combine different properties of a type that may benefit from being treated individually.
+
+One example for such a refinement in LiquidHaskell is shown in \ref{lst:combined_lt}.
+There, we define a function that extracts the maximum value of a (descendingly) ordered, non-empty list.
+In this example, this operation is equivalent to simply extracting the \texttt{head} of a list.
+While we only define one refinement, it actually operates on two different properties of a list.
+First, it imposes a restriction on it's \texttt{size} by requiring it to be non-empty, while the second restriction imposes an order on the list elements.
+
+\lstinputlisting[firstline=6,caption={Example for a function with a combined Refinement Type in LiquidHaskell},label={lst:combined_lt}]{CombinedLT.hs}
+
+While the refinement concerns two different aspects of the list that are independent from each other, current static analysis implementations of Liquid Types will treat it only as a single combined requirement to be fulfilled.
+However especially developers could benefit from a system that treats them independently: When writing code that interfaces with already existing codes such a separated approach may give a developer more fine-grained information on which requirements they still need to fulfill in their implementation.
+
+Effectively one could treat such different requirements as separate \textit{Layers} that can be verified independently.
+Additionally with a proper implementation using such layers may also result in an implementation benefit, that different Layers can be defined individually and combined arbitrarily, as opposed to the example in Listing \ref{lst:combined_lt} where the combination of the two traits had to be explicitly defined as an refinement on it's own.
+
 \subsection{Problem Definition}

### 2022-10-14 paper/paper.tex
Fix lstinput path
Added first idea for problem definition

@@
 
-\lstset{inputpath=code_examples/src/Examples}
+\lstset{inputpath=code_examples/lib/Examples}
 \lstset{
@@
 
+One other possible usage of such an iterative approach is that it could allow to more flexibly add different static analysis approaches on top of each other.
+For example \ref{handley2020} have shown how to use liquid types in Haskell to statically perform cost analysis of different resources such as memory consumption or computational complexity.
+In their approach they showed that liquid types can be used to reason about their upper and lower bounds.
+However, in their work they only analysed one resource at a time.
+However, using a layered approach would allow to easily combine these different resource analyses to eg.
+track and reason about both memory usage and computational complexity.
+
 \subsection{Problem Definition}
+
+Current Liquid Type implementations limit us to only defining one liquid type for variables, arguments or return types.
+This forces us to combine all properties we want to statically check and verify into one big predicate.
+Using such an approach has the drawback that all properties are checked simultaneously which might make it difficult for a developer to identify due to which property a type error occurs.
+Additionally previously declared liquid types may not be reused when wanting to refine them further.
+
+We therefore propose an implementation of \textit{Layered Liquid\todo{Just Layered or Layered Liquid?} Types}, that allows to define types as a conjunction of previously defined Liquid Types.
+These types are then checked iteratively to ensure compability between functions and arguments.
+
+Assume we have a function \texttt{f} that takes as input a single liquid type $T^{in}$ refined by the predicates $p^{in}_1,\cdots,p^{in}_n$ and result type \texttt{out} and a variable \texttt{arg} of liquid type $T^{arg}$ refined by the predicates $p^{arg}_1,\cdots,p^{arg}_m$:
+
+\begin{lstlisting}[mathescape=true]
+	-- Liquid Type Definitions
+	{-@ type $T^{in}$ t = { t | $p^{in}_1 \wedge ... \wedge p^{in}_n$  } @-}
+	{-@ type $T^{arg}$ t = { t | $p^{arg}_1 \wedge ... \wedge p^{arg}_m$ } @-}
+
+	f : $T^{in}$ -> out 
+	arg : $T^{arg}$
+\end{lstlisting}
+
+When now calling \texttt{f} with \texttt{arg} as an argument, current implementations of liquid types will check for compability of the arguments in a manner similar to:
+
+$$ \left( \bigwedge^{m}_{i=1} p^{arg}_i \right) \implies \left( \bigwedge^{n}_{j=1} p^{in}_j \right) $$
+
+In our layered approach we rather propose that we can define separate layered types for each predicate and easily use their conjunction as the specific refined type for arguments and values:
+
+\begin{lstlisting}[mathescape=true]
+	-- Liquid Type Definitions for inputs to f
+	{-@ type $T^{in}_1$ t = { t | $p^{in}_1$ } @-}
+	{-@ type ... @-}
+	{-@ type $T^{in}_n$ t = { t | $p^{in}_n$ } @-}
+
+	-- Liquid Type Definitions for arg
+	{-@ type $T^{arg}_1$ t = { t | $p^{arg}_1$ } @-}
+	{-@ type ...  @-}
+	{-@ type $T^{arg}_m$ t = { t | $p^{arg}_m$ } @-}
+
+	f : {$T^{in}_1 \wedge ... \wedge T^{in}_n$} -> out 
+	arg : {$T^{arg}_1 \wedge ... \wedge T^{arg}_m$}
+\end{lstlisting}
+
+In this layered approach predicates are not checked simultaneously but rather in an iterative way, which could intuitively rather be expressed as the (which is semantically equivalent to the version above):
+
+$$ \bigwedge^{n}_{j=1} \left( \bigwedge^{m}_{i=1} p^{arg}_i \implies  p^{in}_j \right) $$
+
+Using this, we can specifically identify for which predicate the implications fail, giving the user a more accurate description of why their program may contain a type error.
+
+While on first glance this approach could be described as an approach to \textit{Gradual Verification}, we will explicitly refer to our approach as \textit{Layered} to distinguish it from \textit{Gradual Verification} techniques.
+While the latter deal with the gradient between incomplete and complete information (in that case on types), our approach is rather concerned not with what information is available on the typing but rather how it is processed.
+While gradual approaches might still type-check with incomplete typing information our approach would fail to typecheck in such a scenario as it still requires complete information.
+
 \subsection{Impact}

### 2022-10-17 paper/paper.tex
Minor
Added first version of Impact and short description of approach

@@
 \begin{itemize}
-	\item Ensuring values are in certain ranges
+	\item Ensuring values are in certain bounds
 	\item Ensuring that a list is non-empty
@@
 
-Using this, we can specifically identify for which predicate the implications fail, giving the user a more accurate description of why their program may contain a type error.
-
 While on first glance this approach could be described as an approach to \textit{Gradual Verification}, we will explicitly refer to our approach as \textit{Layered} to distinguish it from \textit{Gradual Verification} techniques.
@@
 \subsection{Impact}
+
+We expect our approach to have the following impacts on Liquid Types in practice:
+
+\paragraph{Improved Error Messages}
+
+One current challenge when using liquid types is that all predicates for a type are checked simultaneously.
+This can make it difficult to identify which predicate caused a type liquid error.
+By separating the predicates into separate \textit{Layers} and verifying them individually, it is possible to precisely identify the predicate causing the type error.
+We expect this to help developers to more easily identify and fix these type errors in their code.
+
+\paragraph{Improved Usability and Reusability}
+
+In current Liquid Type implementations, types that require multiple predicates are defined as a single type.
+This may make it difficult for a developer to directly identify which properties a specific liquid type ensures.
+By separating different properties into a more fine-grained layered approach, this may make it easier to identify these properties.
+Additionally, by separating the predicates into different layers, it is possible to reuse previously defined liquid types in a more flexible way.
+For example, if a developer wants to refine a previously defined liquid type, they can now do so by adding a new layer to it, instead of having to redefine the entire type.
+
+\paragraph{Improved Composability}
+
+In addition to the improved reusability of our approach, we also expect it to improve the composability of different liquid types.
+With our suggested approach, new refinements can simply be created by composing already existing layers.
+This vastly increases modularity of the liquid type approach.
+It also may allow to re-use layers that have already been defined by other developers for ones own implementation.
+
 \subsection{Approach}
+We will start by giving a brief overview of the general principles of refinement types and liquid types in general in section \ref{sec:background}.
+Additionally, we will discuss the current state of the art in liquid types and gradual verification in section \ref{sec:related}.
+
+In section \ref{sec:implementation} we will first give a more detailed description of our approach and define a simple functional language that we will use for further demonstration.
+We will then implement a prototype type checker for our layered approach.
+Using this simple prototype we will showcase how our approach can be used on various simple examples.
+
+In section \ref{sec:evaluation} we will discuss the limitations of our approach and possible extensions.
+Additionally, we will evaluate how our approach compares to using a pure, non-layered, verification approach.
+
+Test
+
 \subsection{Contributions}
+\section{Background}
+\label{sec:background}
 \section{Related Work}
-\section{Background}
+\label{sec:related}
 \section{Implementation}
+\label{sec:implementation}
 \section{Evaluation}
+\label{sec:evaluation}
 \section{Conclusion}
+\label{sec:conclusion}
 \section{Acknowledgments}
+\label{sec:acknowledgments}
 

### 2022-10-18 paper/paper.tex
Added brief description of language including syntax

@@
 \label{sec:implementation}
+
+In this section we will describe our implementation of a layered typing approach in detail.
+We will start by describing the simple functional language we will use for our examples.
+We will then describe how our layered type checker is implemented.
+Finally, we will showcase how our approach can be used on various simple examples.
+
+\subsection{Language}
+\label{ssec:language}
+
+To demonstrate our layered approach we implemented a simple functional language.
+The syntax of said language is described in \ref{fig:lang}.
+The language is similar to most simple functional languages and generally follows the reduction semantics of a standard Hindley-Milner type system.
+The only data types available in the language are the \texttt{Int} and the \texttt{Bool} type.
+However, the language does not strictly require the programmer to explicitly specify the type of a variable.
+
+\begin{figure}[ht!]
+\begin{align*}
+	id \in \texttt{Ident}\ 	&::=\ [a...zA...Z]+ \\
+	v \in \texttt{Var}\ 	&::=\ id \\
+	f \in \texttt{Fun}\ 	&::=\ id_1\ id_2\ \{\ s\ \}\\
+	e \in \texttt{Expr}\ 	&::=\ v\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ c \\
+	c \in \texttt{Const}\ 	&::=\ \texttt{True}\ |\ \texttt{False}\ |\ \mathbb{Z} \\
+	s \in \texttt{Stmt}\ 	&::=\ v\ :=\ e\ |\ f(v)\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ l \\
+	l \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
+	\circledplus\ 		&::=\ \texttt{+}\ |\ \texttt{-} \\
+	\circleddot\ 		&::=\ \texttt{!=}\ |\ \texttt{==}\ |\ \texttt{<} \ |\ \texttt{>} \ |\ \texttt{<=} \ |\ \texttt{>=} \\
+\end{align*}
+\caption{Syntax of the simple functional language}
+\label{fig:lang}
+\end{figure}
+
+Additionally to standard operations, the syntax of the language supports creating layers that can be applied onto each identifier. The syntax of a layer consists of the following:
+\begin{itemize}
+	\item $id_1$ -- The name of the identifier to be refined
+	\item $id_2$ -- The name of the layer
+	\item An arbitrary \textit{String} that represents the refinement.
+Note that the syntax does not restrict the format of this string, as this information will be parsed later by the corresponding type checking layer.
+\end{itemize}
+
+These layers can then be used to refine an identifier.
+In a very basic and minimal approach these layers can be used to annotate types to variables and identifiers.
+Additionally, one might add additional layers to a variable to ensure more properties.
+Each layer will be checked individually and independently.
+A more detailled approach of the architecture behind our layered approach will be described in section \ref{ssec:architecture}.
+
+An example of a program written in our language can be seen in \ref{lst:fac_example}.
+The program defines a factorial function that is refined in two layers.
+The first layer \textit{base} is a simple type layer that defines that \texttt{fac} takes an \texttt{Int} and returns an \texttt{Int}.
+The second layer \textit{liquid} is a refinement layer that refines that the input should be greater or equal to zero and that the output will be greater than zero.
+
+\begin{lstlisting}[caption={Example of a factorial function in our simple language}, label={lst:fac_example}]
+fac::base:: Int -> Int
+fac::liquid:: {v:Int | v >= 0} -> {v:Int | v > 0}
+fac x { 
+let cond := (x == 0) in
+	if cond then 
+		1 
+	else 
+		x * fac (x - 1) 
+}
+\end{lstlisting}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cl} 
+      \\
+      \(\frac
+         {x \ : \ \sigma \ \in \ \Gamma}
+        {\Gamma \  \vdash \ x \ : \ \sigma} \) 
+      &  [\texttt{Var}]  \\
+      \\
+
+      \\
+      \(\frac
+        {\Gamma \  \vdash \ {e}_{0} \ : \ \tau \rightarrow {\tau}' \;\;\;\; \Gamma \  \vdash \  {e}_{1} \ : \ \tau}
+        {\Gamma \vdash {e}_{0} \; {e}_{1} \ : \ {\tau}'} \)
+      & [\texttt{App}] \\
+      \\
+
+      \\
+      \(\frac
+        {\Gamma , \ x \ : \ \tau \ \vdash \ e \ : \ {\tau}'}
+        {\Gamma \ \vdash \ \lambda x \ . \ e \ : \ \tau \rightarrow{\tau}'}\)
+      & [\texttt{Abs}] \\
+      \\
+
+      \\
+      \(\frac
+        {\Gamma \ \vdash \ {e}_{0} \ : \ \sigma \;\;\;\; 
+         \Gamma, \ x \ : \ \sigma \ \vdash \ {e}_{1} \ : \ \tau}
+        {\Gamma \ \vdash \ \texttt{let} \ x \ = \ {e}_{0} \ \texttt{in} \ {e}_{1} \ : \ \tau}\)
+      & [\texttt{Let}] \\ 
+      \\
+
+      \\
+      \(\frac
+        {
+            \newline
+            \Gamma \ \vdash \ e \ : \ {\sigma}' \;\;\;\;
+            {\sigma}' \ \sqsubseteq  \ \sigma
+        }
+        {
+            \Gamma \ \vdash \ e \ : \ \sigma
+        }\)
+      & [\texttt{Inst}] \\
+      \\
+
+      \\
+      \(
+       \frac
+       {
+        \Gamma \ \vdash \ e \ : \ \sigma \;\;\;\;
+        \alpha \ \notin \text{free}(\Gamma)
+       }
+       {
+        \Gamma \ \vdash \ e \ : \ \forall \ \alpha \ . \ \sigma
+       }
+      \)
+      & [\texttt{Gen}] \\
+      \\
+    \end{tabular}
+  \end{center}
+\end{table}
+
+\subsection{Language Implementation Architecture}
+\label{ssec:architecture}
+
+The implementation of our language will be done in the \texttt{Python} programming language.
+Roughly speaking, the implementation will consist of three parts, which we will describe throughout this section:
+
+\begin{enumerate}
+	\item A parser that parses the input program and creates an abstract syntax tree (AST) of the program (\ref{sssec:parser}).
+	\item A layered type checker that checks the defined type layers for the tree incrementally (\ref{sssec:layered_type_checker}).
+	\item A code generator that generates the corresponding code for the input program (\ref{sssec:code_generator}).
+\end{enumerate}
+
+\subsubsection{Parser}
+\label{sssec:parser}
+
+\subsubsection{Layered Type Checker}
+\label{sssec:layered_type_checker}
+
+\subsubsection{Code Generator}
+\label{sssec:code_generator}
+
+\subsection{Examples}
+\label{ssec:examples}
+
 \section{Evaluation}

### 2022-10-24 paper/paper.tex
* First version of prototype compiler (Parser+Interpreter)

@@
 	f \in \texttt{Fun}\ 	&::=\ id_1\ id_2\ \{\ s\ \}\\
-	e \in \texttt{Expr}\ 	&::=\ v\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ c \\
+	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e) \\
 	c \in \texttt{Const}\ 	&::=\ \texttt{True}\ |\ \texttt{False}\ |\ \mathbb{Z} \\
-	s \in \texttt{Stmt}\ 	&::=\ v\ :=\ e\ |\ f(v)\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ l \\
+	s \in \texttt{Stmt}\ 	&::=\ v\ :=\ e\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ l \\
 	l \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
@@
 These layers can then be used to refine an identifier.
-In a very basic and minimal approach these layers can be used to annotate types to variables and identifiers.
+In a very basic and minimal approach these layers can be used to annotate types to variables and functions.
 Additionally, one might add additional layers to a variable to ensure more properties.
@@
 
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cl} 
-      \\
-      \(\frac
-         {x \ : \ \sigma \ \in \ \Gamma}
-        {\Gamma \  \vdash \ x \ : \ \sigma} \) 
-      &  [\texttt{Var}]  \\
-      \\
-
-      \\
-      \(\frac
-        {\Gamma \  \vdash \ {e}_{0} \ : \ \tau \rightarrow {\tau}' \;\;\;\; \Gamma \  \vdash \  {e}_{1} \ : \ \tau}
-        {\Gamma \vdash {e}_{0} \; {e}_{1} \ : \ {\tau}'} \)
-      & [\texttt{App}] \\
-      \\
-
-      \\
-      \(\frac
-        {\Gamma , \ x \ : \ \tau \ \vdash \ e \ : \ {\tau}'}
-        {\Gamma \ \vdash \ \lambda x \ . \ e \ : \ \tau \rightarrow{\tau}'}\)
-      & [\texttt{Abs}] \\
-      \\
-
-      \\
-      \(\frac
-        {\Gamma \ \vdash \ {e}_{0} \ : \ \sigma \;\;\;\; 
-         \Gamma, \ x \ : \ \sigma \ \vdash \ {e}_{1} \ : \ \tau}
-        {\Gamma \ \vdash \ \texttt{let} \ x \ = \ {e}_{0} \ \texttt{in} \ {e}_{1} \ : \ \tau}\)
-      & [\texttt{Let}] \\ 
-      \\
-
-      \\
-      \(\frac
-        {
-            \newline
-            \Gamma \ \vdash \ e \ : \ {\sigma}' \;\;\;\;
-            {\sigma}' \ \sqsubseteq  \ \sigma
-        }
-        {
-            \Gamma \ \vdash \ e \ : \ \sigma
-        }\)
-      & [\texttt{Inst}] \\
-      \\
-
-      \\
-      \(
-       \frac
-       {
-        \Gamma \ \vdash \ e \ : \ \sigma \;\;\;\;
-        \alpha \ \notin \text{free}(\Gamma)
-       }
-       {
-        \Gamma \ \vdash \ e \ : \ \forall \ \alpha \ . \ \sigma
-       }
-      \)
-      & [\texttt{Gen}] \\
-      \\
-    \end{tabular}
-  \end{center}
-\end{table}
-
 \subsection{Language Implementation Architecture}
@@
 \begin{enumerate}
-	\item A parser that parses the input program and creates an abstract syntax tree (AST) of the program (\ref{sssec:parser}).
-	\item A layered type checker that checks the defined type layers for the tree incrementally (\ref{sssec:layered_type_checker}).
-	\item A code generator that generates the corresponding code for the input program (\ref{sssec:code_generator}).
+	\item A parser that parses the input program and creates an abstract syntax tree (AST) of the program.
+	\item A layered type checker that checks the defined type layers for the tree incrementally.
+	\item A code generator that generates the corresponding code for the input program.
 \end{enumerate}
-
-\subsubsection{Parser}
-\label{sssec:parser}
-
-\subsubsection{Layered Type Checker}
-\label{sssec:layered_type_checker}
-
-\subsubsection{Code Generator}
-\label{sssec:code_generator}
 

### 2022-10-24 code/LayeredTypes/parser/implementations.py

@@ -1,0 +1,4 @@
+from builtins import *
+
+def test(x):
+    print(f"Called test with value {x}!")

### 2022-10-24 code/LayeredTypes/parser/layers/base.py

@@ -7,3 +7,6 @@ def typecheck(tree):
 
 def parse_type(type_str):
     print(f"parse_type called with type_str {type_str}")
+
+def test(x):
+    print(f"Called test with value {x}!")

### 2022-10-24 code/LayeredTypes/parser/parser.py

@@ -45,8 +45,8 @@ class CollectLayers(lark.Transformer):
 
 
     def layer(self, tree):
-        ident = tree[0].value
-        layer_ident = tree[1].value
+        ident = tree[0].children[0].value
+        layer_ident = tree[1].children[0].value
         refinement = tree[2].value
 
         if not layer_ident in self.layers.keys():
@@ -55,16 +55,163 @@ class CollectLayers(lark.Transformer):
         self.layers[layer_ident].add_refinement(ident,refinement)
         return lark.Discard
 
+class SimpleInterpreter(lark.visitors.Interpreter):
+    def __init__(self, implementation_file = "implementations"):
+        # Variables and functions are stored in dictionaries for access by our interpreter
+        self.variables = dict()
+        self.functions = dict()
+        self.external_functions = import_module(implementation_file)
+        super().__init__()
+
+    def run(self, tree):
+        # Empty variable and function context before running a new program
+        self.variables = dict()
+        self.functions = dict()
+
+        self.visit(tree)
+    def num(self, tree):
+        return int(tree.children[0])
+
+    def true(self, tree):
+        return True
+
+    def false(self, tree):
+        return False
+
+    def assign(self, tree):
+        identifier = tree.children[0].children[0].value
+        value = self.visit(tree.children[1])
+
+        self.variables[identifier] = value
+    def if_stmt(self, tree):
+
+        condition = self.visit(tree.children[0])
+
+        if condition:
+            return self.visit(tree.children[1])
+
+        return self.visit(tree.children[2])
+
+    def let_stmt(self, tree):
+
+        identifier = tree.children[0].children[0].value
+        value = self.visit(tree.children[1])
+
+        # Store the current value of ident and restore it after the let construct
+        restore_val = False
+        old_val = None
+        if identifier in self.variables.keys():
+            old_val = self.variables[identifier]
+            restore_val = True
+
+        self.variables[identifier] = value
+
+        result = self.visit(tree.children[2])
+
+        self.variables.pop(identifier)
+
+        if restore_val:
+            self.variables[identifier] = old_val
+
+        return result
+
+    def bin_op(self, tree):
+        lhs = self.visit(tree.children[0])
+        rhs = self.visit(tree.children[2])
+
+        op = tree.children[1].value
+
+        if op == "+":
+            return lhs + rhs
+
[…]

### 2022-10-24 code/LayeredTypes/parser/parser.py
* Added custom expression for variable assignments to grammar

@@ -202,6 +202,9 @@ class SimpleInterpreter(lark.visitors.Interpreter):
 
         return self.variables[identifier]
 
+    def custom_expr(self, tree):
+        print("Custom expression: {}".format(tree.children[0].value))
+
 
 Parser = lark.Lark.open("grammar_file.lark", parser= "lalr", debug=True)
 

### 2022-10-25 paper/paper.tex
Slight adaptions to grammar
First brief architecture explanation

@@
 }
+
+% Shorthand command for Lark
+\newcommand{\Lark}[0]{\textsc{Lark}}
 
@@
 	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e) \\
+	\texttt{CustomExpr}\ 	&::=\ \texttt{String}\\
 	c \in \texttt{Const}\ 	&::=\ \texttt{True}\ |\ \texttt{False}\ |\ \mathbb{Z} \\
-	s \in \texttt{Stmt}\ 	&::=\ v\ :=\ e\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ l \\
-	l \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
+	\texttt{Assign}\ 	&::=\ v\ :=\ e\ |\ v\ :=\ \texttt{CustomExpr}\ \\
+	s \in \texttt{Stmt}\ 	&::=\ \texttt{Assign}\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ \ell\ |\ s_1\ s_2\ \\
+	\ell \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
 	\circledplus\ 		&::=\ \texttt{+}\ |\ \texttt{-} \\
@@
 \end{figure}
+
+While our language allows to define functions on its own, we will further allow the usage of function names that may have not been defined in the language itself.
+To allow this, the user can create a separate Python file that contains the definitions of additional functions.
+Our interpreter will then forward the calls to these functions to the Python interpreter.
+This allows us to easily access more complicated functionality such as the \texttt{print} function.
 
@@
 	\item A layered type checker that checks the defined type layers for the tree incrementally.
-	\item A code generator that generates the corresponding code for the input program.
+	\item An interpreter that executes the program.
 \end{enumerate}
+
+To parse a input program, we will use the Python library \Lark{} \cite{Lark}.
+This library allows us to define a grammar for the language and then parse a program according to this grammar.
+All further processing will be performed on the resulting AST.
+To process and execute the AST we will use the Visitors and Transformers provided by the \Lark{} library.
+
+After initially parsing the AST, it will contain explicit nodes for each of the defined layers.
+As a first processing step we use a Transformer to collect these layers and remove the corresponding nodes from the AST.
+The implementation of each node has to be provided by the user as a separate Python module.
+Our language Interpreter will then dynamically load the corresponding module and use it to check the layer.
+To interact with our layered type checker correctly the module has to implement the following functions:
+
+\begin{itemize}
+	\item \texttt{depends\_on} -- This function returns a list of all layers that this layer depends on. This is used to ensure that the layers are checked in the correct order.
+	\item \texttt{parse\_type} -- This function may be used to define and parse custom data types that are defined usinf the $CustomExpr$ rule in the grammar.
+\todo{Implementation details -- When and how is this function called?}
+	\item \texttt{check\_layer} -- This function is used to check the layer.
+It is passed the AST of the program with all the annotated information from the previous layers.
+It should return a boolean value that indicates whether the layer is valid or not.
+\end{itemize}
+
+Using the \texttt{depends\_on} function, the layered type checker will first build a dependency graph of all layers.
+It will then topologically sort the layers and check them in the correct order.
+Once all layers have been checked successfully, the interpreter will execute the program.
 

### 2022-10-25 code/LayeredTypes/compiler/Compiler.py
* Refactored compiler into a module structure

@@ -1,0 +1,64 @@
+
+import lark
+from pathlib import Path
+
+from layers.Layer import Layer
+from layers.LayerImplWrapper import LayerImplWrapper
+from compiler.transformers.CollectLayers import CollectLayers
+from compiler.Interpreters import SimpleInterpreter
+
+class LayeredCompiler:
+    def __init__(self
+                 , implementations_file
+                 , layer_base_dir):
+
+        self.implementations_file = Path(implementations_file)
+        self.layer_base_dir = Path(layer_base_dir)
+
+        if not self.implementations_file.is_file():
+            raise FileNotFoundError(f"Could not find implementations file at {self.implementations_file}")
+
+        if not self.layer_base_dir.is_dir():
+            raise FileNotFoundError(f"Could not find layer base directory at {self.layer_base_dir}")
+
+        self.parser = lark.Lark.open("grammar_file.lark", parser= "lalr", debug=True)
+
+    def typecheck(self, input_file):
+        tree = self.__parse(input_file)
+
+        self.__typecheck(tree)
+
+        return True
+
+    def __typecheck(self, tree):
+        # TODO: Implement proper typechecking
+        lv = CollectLayers()
+
+        ReducedTree = lv.transform(tree)
+
+        layer_wrapper = LayerImplWrapper(lv.layers["base"])
+        pass
+
+    def __parse(self, input_file):
+        with open(input_file) as f:
+            tree = self.parser.parse(f.read())
+
+        return tree
+
+    def compile(self, program):
+        tree = self.__parse(program)
+
+        if not self.__typecheck(tree):
+            raise RuntimeError("Typechecking failed")
+
+        return tree
+    def run(self, program):
+        tree = self.compile(program)
+
+        interpreter = SimpleInterpreter(self.implementations_file)
+
+        interpreter.run(tree)
+        pass
+
+
+

### 2022-10-25 code/LayeredTypes/compiler/Interpreters.py

@@ -1,65 +1,17 @@
 from importlib import import_module
 import lark
-
-class Layer:
-    def __init__(self,
-                 layer_ident : str):
-        self.name = layer_ident
-        self.refinements = dict()
-
-    def add_refinement(self,
-                       identifier,
-                       refinement_str):
-        if not identifier in self.refinements.keys():
-            self.refinements[identifier] = []
-        self.refinements[identifier].append(refinement_str)
-
-    def __str__(self):
-
-        res = "{}:\n".format(self.name)
-        for ident in self.refinements:
-            refinements = self.refinements[ident]
-            res += "\t{}\n".format(ident)
-
-            for refinement in refinements:
-                res += "\t\t{}\n".format(refinement)
-
-        return res
-
-class LayerImplWrapper:
-    def __init__(self,
-                 layer: Layer):
-        self.layer = layer
-        try:
-            self.module = import_module("layers.{}".format(layer.name))
-        except ModuleNotFoundError as e:
-            raise FileNotFoundError(f"No implementation file (tried to load file './layers/{layer.name}.py') found for layer '{layer.name}'")
-        self.depends_on = getattr(self.module, "depends_on")
-        self.typecheck = getattr(self.module, "typecheck")
-        self.parse_type = getattr(self.module, "parse_type")
-
-class CollectLayers(lark.Transformer):
-    def __init__(self):
-        super().__init__()
-        self.layers = dict()
-
-
-    def layer(self, tree):
-        ident = tree[0].children[0].value
-        layer_ident = tree[1].children[0].value
-        refinement = tree[2].value
-
-        if not layer_ident in self.layers.keys():
-            self.layers[layer_ident] = Layer(layer_ident)
-        print("Processing layer {} for identifier {} with rule '{}'".format(layer_ident, ident, refinement))
-        self.layers[layer_ident].add_refinement(ident,refinement)
-        return lark.Discard
+from pathlib import Path
 
 class SimpleInterpreter(lark.visitors.Interpreter):
     def __init__(self, implementation_file = "implementations"):
         # Variables and functions are stored in dictionaries for access by our interpreter
         self.variables = dict()
         self.functions = dict()
+
+        impl_path = Path(implementation_file)
+        if not impl_path.is_file():
[…]

### 2022-10-25 code/LayeredTypes/compiler/transformers/CollectLayers.py

@@ -1,0 +1,19 @@
+import lark
+from layers.Layer import Layer
+
+class CollectLayers(lark.Transformer):
+    def __init__(self):
+        super().__init__()
+        self.layers = dict()
+
+
+    def layer(self, tree):
+        ident = tree[0].children[0].value
+        layer_ident = tree[1].children[0].value
+        refinement = tree[2].value
+
+        if not layer_ident in self.layers.keys():
+            self.layers[layer_ident] = Layer(layer_ident)
+        print("Processing layer {} for identifier {} with rule '{}'".format(layer_ident, ident, refinement))
+        self.layers[layer_ident].add_refinement(ident,refinement)
+        return lark.Discard

### 2022-10-25 code/LayeredTypes/layers/Layer.py

@@ -1,0 +1,24 @@
+class Layer:
+    def __init__(self,
+                 layer_ident : str):
+        self.name = layer_ident
+        self.refinements = dict()
+
+    def add_refinement(self,
+                       identifier,
+                       refinement_str):
+        if not identifier in self.refinements.keys():
+            self.refinements[identifier] = []
+        self.refinements[identifier].append(refinement_str)
+
+    def __str__(self):
+
+        res = "{}:\n".format(self.name)
+        for ident in self.refinements:
+            refinements = self.refinements[ident]
+            res += "\t{}\n".format(ident)
+
+            for refinement in refinements:
+                res += "\t\t{}\n".format(refinement)
+
+        return res

### 2022-10-25 code/LayeredTypes/layers/LayerImplWrapper.py

@@ -1,0 +1,30 @@
+import importlib.util
+import sys
+from importlib import import_module
+import Layer
+from pathlib import Path
+
+class LayerImplWrapper:
+    def __init__(self,
+                 layer_base_dir,
+                 layer: Layer):
+        self.layer = layer
+        self.layer_base_dir = Path(layer_base_dir)
+
+        if not self.layer_base_dir.is_dir():
+            raise FileNotFoundError(f"Could not find layer base directory at {self.layer_base_dir}")
+
+        full_path = Path(self.layer_base_dir, f"{layer.name}.py")
+
+        if not full_path.is_file():
+            raise FileNotFoundError(f"No implementation file (tried to load file '{full_path}') found for layer '{layer.name}'")
+
+        module_name = f"layers.{self.layer.name}"
+        spec = importlib.util.spec_from_file_location(module_name, full_path)
+        self.module = importlib.util.module_from_spec(spec)
+        sys.modules[module_name] = self.module
+        spec.loader.exec_module(self.module)
+
+        self.depends_on = getattr(self.module, "depends_on")
+        self.typecheck = getattr(self.module, "typecheck")
+        self.parse_type = getattr(self.module, "parse_type")

### 2022-10-25 code/LayeredTypes/test/simple.py

@@ -1,3 +1,3 @@
-from parser.parser import Parser
+from compiler.Compiler import Parser
 
 print(Parser.parse("myvar := 20"))

### 2022-10-25 code/LayeredTypes/compiler/Compiler.py
* Added __init__.py

@@ -2,7 +2,6 @@
 import lark
 from pathlib import Path
 
-from layers.Layer import Layer
 from layers.LayerImplWrapper import LayerImplWrapper
 from compiler.transformers.CollectLayers import CollectLayers
 from compiler.Interpreters import SimpleInterpreter
@@ -37,6 +36,8 @@ class LayeredCompiler:
         ReducedTree = lv.transform(tree)
 
         layer_wrapper = LayerImplWrapper(lv.layers["base"])
+
+        return ReducedTree
         pass
 
     def __parse(self, input_file):

### 2022-10-25 code/LayeredTypes/compiler/Compiler.py
* Added unit tests for parser

@@ -1,3 +1,4 @@
+import os.path
 
 import lark
 from pathlib import Path
@@ -20,10 +21,12 @@ class LayeredCompiler:
         if not self.layer_base_dir.is_dir():
             raise FileNotFoundError(f"Could not find layer base directory at {self.layer_base_dir}")
 
-        self.parser = lark.Lark.open("grammar_file.lark", parser= "lalr", debug=True)
+        # Build path for grammar file
+        grammar_file = os.path.dirname(os.path.realpath(__file__)) + "/grammar_file.lark"
+        self.parser = lark.Lark.open(grammar_file, parser= "lalr", debug=True)
 
     def typecheck(self, input_file):
-        tree = self.__parse(input_file)
+        tree = self.parse(input_file)
 
         self.__typecheck(tree)
 
@@ -33,21 +36,21 @@ class LayeredCompiler:
         # TODO: Implement proper typechecking
         lv = CollectLayers()
 
-        ReducedTree = lv.transform(tree)
+        reducedTree = lv.transform(tree)
 
-        layer_wrapper = LayerImplWrapper(lv.layers["base"])
+        # layer_wrapper = LayerImplWrapper(lv.layers["base"])
 
-        return ReducedTree
+        return reducedTree
         pass
 
-    def __parse(self, input_file):
+    def parse(self, input_file):
         with open(input_file) as f:
             tree = self.parser.parse(f.read())
 
         return tree
 
     def compile(self, program):
-        tree = self.__parse(program)
+        tree = self.parse(program)
 
         if not self.__typecheck(tree):
             raise RuntimeError("Typechecking failed")
@@ -58,8 +61,7 @@ class LayeredCompiler:
 
         interpreter = SimpleInterpreter(self.implementations_file)
 
-        interpreter.run(tree)
-        pass
+        return interpreter.run(tree)
 
 
 

### 2022-10-25 code/LayeredTypes/compiler/Interpreters.py

@@ -1,6 +1,7 @@
-from importlib import import_module
+import importlib
 import lark
 from pathlib import Path
+import sys
 
 class SimpleInterpreter(lark.visitors.Interpreter):
     def __init__(self, implementation_file = "implementations"):
@@ -12,7 +13,12 @@ class SimpleInterpreter(lark.visitors.Interpreter):
         if not impl_path.is_file():
             raise FileNotFoundError(f"No implementation file (tried to load file '{implementation_file}') found")
 
-        self.external_functions = import_module(implementation_file)
+        module_name = "implementations.test"
+        spec = importlib.util.spec_from_file_location(module_name, impl_path)
+        self.external_functions = importlib.util.module_from_spec(spec)
+        sys.modules[module_name] = self.external_functions
+        spec.loader.exec_module(self.external_functions)
+
         super().__init__()
 
     def run(self, tree):
@@ -20,7 +26,7 @@ class SimpleInterpreter(lark.visitors.Interpreter):
         self.variables = dict()
         self.functions = dict()
 
-        self.visit(tree)
+        return self.visit(tree)[-1]
     def num(self, tree):
         return int(tree.children[0])
 

### 2022-10-25 code/LayeredTypes/layers/LayerImplWrapper.py

@@ -1,7 +1,7 @@
 import importlib.util
 import sys
-from importlib import import_module
-import Layer
+
+from layers.Layer import Layer
 from pathlib import Path
 
 class LayerImplWrapper:

### 2022-10-25 code/LayeredTypes/test/implementations.py

@@ -2,3 +2,4 @@ from builtins import *
 
 def test(x):
     print(f"Called test with value {x}!")
+    return x

### 2022-10-25 code/LayeredTypes/test/simple.py

@@ -1,3 +1,0 @@
-from compiler.Compiler import Parser
-
-print(Parser.parse("myvar := 20"))

### 2022-10-25 code/LayeredTypes/compiler/Interpreters.py
Slightly restructuring for unittests
Setup Github Workflow for automated unit testing

@@ -13,7 +13,7 @@ class SimpleInterpreter(lark.visitors.Interpreter):
         if not impl_path.is_file():
             raise FileNotFoundError(f"No implementation file (tried to load file '{implementation_file}') found")
 
-        module_name = "implementations.test"
+        module_name = "implementations"
         spec = importlib.util.spec_from_file_location(module_name, impl_path)
         self.external_functions = importlib.util.module_from_spec(spec)
         sys.modules[module_name] = self.external_functions

### 2022-10-25 code/LayeredTypes/tests/test_syntax.py

@@ -1,0 +1,131 @@
+import os.path
+import unittest
+
+from compiler.Compiler import LayeredCompiler
+
+class TestParser(unittest.TestCase):
+
+    def __parse_file(self, file_path):
+        impl_path = os.path.dirname(os.path.realpath(__file__)) + "/implementations.py"
+        layer_path = os.path.dirname(os.path.realpath(__file__)) + "/layer_implementations"
+        compiler = LayeredCompiler(impl_path, layer_path)
+
+        src_path = os.path.dirname(os.path.realpath(__file__)) + f"{file_path}"
+        tree = compiler.parse(src_path)
+
+        return tree
+
+
+    def test_layer_def(self):
+        tree = self.__parse_file("/test_code/syntax/layer.fl")
+
+        self.assertEqual(tree.data, "start")
+        self.assertEqual(tree.children[0].data, "layer")
+        self.assertEqual(tree.children[0].children[0].data, "ident")
+        self.assertEqual(tree.children[0].children[0].children[0].value, "id")
+        self.assertEqual(tree.children[0].children[1].data, "ident")
+        self.assertEqual(tree.children[0].children[1].children[0].value, "layer")
+        self.assertEqual(tree.children[0].children[2].value, "{Refinement}")
+        pass
+
+    def test_assign(self):
+        tree = self.__parse_file("/test_code/syntax/assign.fl")
+
+        self.assertEqual(tree.data, "start")
+        self.assertEqual(tree.children[0].data, "assign")
+        self.assertEqual(tree.children[0].children[0].data, "ident")
+        self.assertEqual(tree.children[0].children[0].children[0], "x")
+        self.assertEqual(tree.children[0].children[1].data, "true")
+
+        self.assertEqual(tree.children[1].data, "assign")
+        self.assertEqual(tree.children[1].children[0].data, "ident")
+        self.assertEqual(tree.children[1].children[0].children[0], "y")
+        self.assertEqual(tree.children[1].children[1].data, "ident")
+        self.assertEqual(tree.children[1].children[1].children[0], "x")
+
+        tree = self.__parse_file("/test_code/syntax/assign_custom.fl")
+
+        self.assertEqual(tree.data, "start")
+        self.assertEqual(tree.children[0].data, "assign")
+        self.assertEqual(tree.children[0].children[0].data, "ident")
+        self.assertEqual(tree.children[0].children[0].children[0], "x")
[…]

### 2022-10-26 code/LayeredTypes/tests/syntax.py
* Added unit tests for binary operations * Adapted grammar for function calls to explicitly require braces

@@ -1,131 +1,0 @@
-import os.path
-import unittest
-
-from compiler.Compiler import LayeredCompiler
-
-class TestParser(unittest.TestCase):
-
-    def __parse_file(self, file_path):
-        impl_path = os.path.dirname(os.path.realpath(__file__)) + "/implementations.py"
-        layer_path = os.path.dirname(os.path.realpath(__file__)) + "/layer_implementations"
-        compiler = LayeredCompiler(impl_path, layer_path)
-
-        src_path = os.path.dirname(os.path.realpath(__file__)) + f"{file_path}"
-        tree = compiler.parse(src_path)
-
-        return tree
-
-
-    def test_layer_def(self):
-        tree = self.__parse_file("/test_code/syntax/layer.fl")
-
-        self.assertEqual(tree.data, "start")
-        self.assertEqual(tree.children[0].data, "layer")
-        self.assertEqual(tree.children[0].children[0].data, "ident")
-        self.assertEqual(tree.children[0].children[0].children[0].value, "id")
-        self.assertEqual(tree.children[0].children[1].data, "ident")
-        self.assertEqual(tree.children[0].children[1].children[0].value, "layer")
-        self.assertEqual(tree.children[0].children[2].value, "{Refinement}")
-        pass
-
-    def test_assign(self):
-        tree = self.__parse_file("/test_code/syntax/assign.fl")
-
-        self.assertEqual(tree.data, "start")
-        self.assertEqual(tree.children[0].data, "assign")
-        self.assertEqual(tree.children[0].children[0].data, "ident")
-        self.assertEqual(tree.children[0].children[0].children[0], "x")
-        self.assertEqual(tree.children[0].children[1].data, "true")
-
-        self.assertEqual(tree.children[1].data, "assign")
-        self.assertEqual(tree.children[1].children[0].data, "ident")
-        self.assertEqual(tree.children[1].children[0].children[0], "y")
-        self.assertEqual(tree.children[1].children[1].data, "ident")
-        self.assertEqual(tree.children[1].children[1].children[0], "x")
-
-        tree = self.__parse_file("/test_code/syntax/assign_custom.fl")
-
-        self.assertEqual(tree.data, "start")
-        self.assertEqual(tree.children[0].data, "assign")
-        self.assertEqual(tree.children[0].children[0].data, "ident")
-        self.assertEqual(tree.children[0].children[0].children[0], "x")
[…]

### 2022-10-26 code/LayeredTypes/tests/test_syntax.py

@@ -79,10 +79,29 @@ class TestParser(unittest.TestCase):
         self.assertEqual(tree.children[0].children[1].children[0], "42")
         self.assertEqual(tree.children[0].children[2].data, "ident")
         self.assertEqual(tree.children[0].children[2].children[0], "x")
-        pass
 
     def test_bin_op(self):
-        pass
+        tree = self.__parse_file("/test_code/syntax/bin_op.fl")
+
+        def check_bin_op(tree, op):
+            self.assertEqual(tree.data, "bin_op")
+            self.assertEqual(tree.children[0].data, "ident")
+            self.assertEqual(tree.children[0].children[0], "a")
+            self.assertEqual(tree.children[1], op)
+            self.assertEqual(tree.children[2].data, "ident")
+            self.assertEqual(tree.children[2].children[0], "b")
+
+        self.assertEqual(tree.data, "start")
+        check_bin_op(tree.children[0], "+")
+        check_bin_op(tree.children[1], "-")
+        check_bin_op(tree.children[2], "*")
+        check_bin_op(tree.children[3], ">")
+        check_bin_op(tree.children[4], "<")
+        check_bin_op(tree.children[5], ">=")
+        check_bin_op(tree.children[6], "<=")
+        check_bin_op(tree.children[7], "==")
+        check_bin_op(tree.children[8], "!=")
+
 
     def test_call(self):
         tree = self.__parse_file("/test_code/syntax/fun_call.fl")

### 2022-10-26 code/LayeredTypes/compiler/Interpreters.py
* Adapted language to support functions with an arbitrary number of arguments

@@ -26,6 +26,7 @@ class SimpleInterpreter(lark.visitors.Interpreter):
         self.variables = dict()
         self.functions = dict()
 
+        # Return the value of the last expression in the program
         return self.visit(tree)[-1]
     def num(self, tree):
         return int(tree.children[0])
@@ -111,45 +112,50 @@ class SimpleInterpreter(lark.visitors.Interpreter):
 
     def fun_call(self, tree):
         fun_id = tree.children[0].value
-        fun_arg = self.visit(tree.children[1])
+
+        fun_args = [self.visit(child) for child in tree.children[1:-1]]
 
         if not fun_id in self.functions.keys():
             if getattr(self.external_functions, fun_id, None):
-                return getattr(self.external_functions, fun_id)(fun_arg)
+                return getattr(self.external_functions, fun_id)(*fun_args)
             raise RuntimeError(f"Function '{fun_id}' not defined")
 
         fun_def = self.functions[fun_id]
-        arg_ident = fun_def[0]
+        arg_identifiers = fun_def[0]
 
         # In order to allow recursive functions we might need to restore the old value of the argument
-        restore_val = False
-        old_val = None
-        if arg_ident in self.variables.keys():
-            old_val = self.variables[arg_ident]
-            restore_val = True
+        restore_vals = dict()
+
+        for arg_ident,fun_arg in zip(arg_identifiers, fun_args):
+            if arg_ident in self.variables.keys():
+                restore_vals[arg_ident] = self.variables[arg_ident]
+
+            self.variables[arg_ident] = fun_arg
 
         # Add the function argument to the variable context
-        self.variables[arg_ident] = fun_arg
 
         result = self.visit(fun_def[1])
 
-        # Remove the function argument from the variable context
-        self.variables.pop(arg_ident)
+        # Remove the function arguments from the variable context
+        for arg_ident in arg_identifiers:
+            self.variables.pop(arg_ident)
 
-        if restore_val:
-            self.variables[arg_ident] = old_val
+            # Restore the old values of the arguments if necessary
+            if arg_ident in restore_vals.keys():
+                self.variables[arg_ident] = restore_vals[arg_ident]
 
         return result
     def fun_def(self, tree):
-
         fun_id = tree.children[0].value
-        arg_ident = tree.children[1].value
-        fun_def = tree.children[2]
+
+        arg_identifiers = [child.value for child in tree.children[1:-1]]
+
[…]

### 2022-10-26 code/LayeredTypes/tests/test_syntax.py

@@ -106,24 +106,26 @@ class TestParser(unittest.TestCase):
     def test_call(self):
         tree = self.__parse_file("/test_code/syntax/fun_call.fl")
 
+
+        def check_call(tree, function_identifier, arg_identifiers):
+            self.assertEqual(tree.data, "fun_call")
+            self.assertEqual(len(tree.children), len(arg_identifiers) + 1)
+            self.assertEqual(tree.children[0], function_identifier)
+
+            for i in range(len(arg_identifiers)):
+                self.assertEqual(tree.children[i+1].data, arg_identifiers[i][0])
+                if arg_identifiers[i][1] is not None:
+                    self.assertEqual(tree.children[i+1].children[0], arg_identifiers[i][1])
+
+
         self.assertEqual(tree.data, "start")
-        self.assertEqual(tree.children[0].data, "fun_call")
-        self.assertEqual(tree.children[0].children[0], "funCall")
-        self.assertEqual(tree.children[0].children[1].data, "ident")
-        self.assertEqual(tree.children[0].children[1].children[0], "argument")
-
-        self.assertEqual(tree.children[1].data, "fun_call")
-        self.assertEqual(tree.children[1].children[0], "funCall")
-        self.assertEqual(tree.children[1].children[1].data, "num")
-        self.assertEqual(tree.children[1].children[1].children[0], "42")
-
-        self.assertEqual(tree.children[2].data, "fun_call")
-        self.assertEqual(tree.children[2].children[0], "funCall")
-        self.assertEqual(tree.children[2].children[1].data, "true")
-
-        self.assertEqual(tree.children[3].data, "fun_call")
-        self.assertEqual(tree.children[3].children[0], "funCall")
-        self.assertEqual(tree.children[3].children[1].data, "false")
+        check_call(tree.children[0], "funCall", [("ident","argument")])
+        check_call(tree.children[1], "funCall", [("num","42")])
+        check_call(tree.children[2], "funCall", [("true",None)])
+        check_call(tree.children[3], "funCall", [("false",None)])
+        check_call(tree.children[4], "emptyCall", [])
+        check_call(tree.children[5], "twoArgs", [("ident","first"), ("ident","second")])
+        check_call(tree.children[6], "threeArgs", [("ident","first"), ("ident","second"), ("ident","third")])
 
     def test_func_def(self):
         tree = self.__parse_file("/test_code/syntax/fun_def.fl")

### 2022-11-14 paper/paper.tex
* Added tests for layer processing to test suite

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{LayeredTypes}
+% \title{LayeredTypes}
 %TODO: Proper Title
-%\title[LayeredTypes]{Layered Types -- A structured approach to arbitrarily combine Liquid Types}
+\title[Layered Types]{Layered Types -- A structured approach to arbitrarily combine Liquid Types}
 
@@
 By separating different properties into a more fine-grained layered approach, this may make it easier to identify these properties.
-Additionally, by separating the predicates into different layers, it is possible to reuse previously defined liquid types in a more flexible way.
+Additionally, by separating the predicates into different layers, it is possible to reuse liquid types in a more flexible way.
 For example, if a developer wants to refine a previously defined liquid type, they can now do so by adding a new layer to it, instead of having to redefine the entire type.
@@
 It also may allow to re-use layers that have already been defined by other developers for ones own implementation.
+
+\paragraph{Improved Modularity}
+
+By separating the predicates into different layers, it is possible to reuse previously defined liquid types in a more flexible way.
+For example, if a developer wants to refine a previously defined liquid type, they can now do so by adding a new layer to it, instead of having to redefine the entire type.
+Additionally, this also supports the seperation of concerns when collaborating with other developers.
+In current implementations, when interfacing with other developers' code, one needs to be aware of all the refinements required by such code in order to interact with it.
+When different properties are divided into proper layers, this is no longer required.
 
@@
 Additionally, we will evaluate how our approach compares to using a pure, non-layered, verification approach.
-
-Test
 
@@
 
+Using this architecture, we provide a flexible framework that allows to easily extend our toy language with various type checking layers.
+Similar to passes in a compiler, the layers can be added and removed easily.
+Additionally, the layers can be easily combined to create more complex type checking layers.
+
 \subsection{Examples}
@@
 \label{sec:acknowledgments}
+This work was supported by \textit{Fundação para a Ciência e Tecnologia} (FCT) in the LASIGE Research Unit under the ref.
+UIDB/00408/2020 and UIDP/00408/2020, by the CMU--Portugal project CAMELOT (LISBOA-01-0247-FEDER-045915), and the RAP project under the reference (EXPL/CCI-COM/1306/2021).
 

### 2022-11-14 code/LayeredTypes/compiler/Compiler.py

@@ -31,6 +31,7 @@ class LayeredCompiler:
                                      , debug=True
                                      , propagate_positions=True
                                      , transformer=RemoveTokens(["newline"]))
+        self.layers = dict()
 
     def typecheck(self, input_file):
         tree = self.parse(input_file)
@@ -43,14 +44,14 @@ class LayeredCompiler:
         lv = CollectLayers()
         cf_check = CheckCF()
 
-        reducedTree = lv.transform(tree)
-        cf_check.visit(reducedTree)
+        tree = lv.transform(tree)
+        cf_check.visit(tree)
 
         layer_graph = {}
-        layers = {}
+        self.layers = {}
         for layer_id in lv.layers:
-            layers[layer_id] = LayerImplWrapper(self.layer_base_dir, lv.layers[layer_id])
-            layer_graph[layer_id] = layers[layer_id].depends_on()
+            self.layers[layer_id] = LayerImplWrapper(self.layer_base_dir, lv.layers[layer_id])
+            layer_graph[layer_id] = self.layers[layer_id].depends_on()
 
         # Check for cycles
         try:
@@ -60,9 +61,9 @@ class LayeredCompiler:
 
         # Incrementally typecheck each layer based on their topological order
         for layer_id in topo_order:
-            layers[layer_id].typecheck()
+            tree = self.layers[layer_id].typecheck(tree)
 
-        return reducedTree
+        return tree
 
     def parse(self, input_file):
         with open(input_file) as f:

### 2022-11-14 code/LayeredTypes/compiler/transformers/CollectLayers.py

@@ -14,6 +14,6 @@ class CollectLayers(lark.Transformer):
 
         if not layer_ident in self.layers.keys():
             self.layers[layer_ident] = Layer(layer_ident)
-        print("Processing layer {} for identifier {} with rule '{}'".format(layer_ident, ident, refinement))
+        print("Collecting layer {} for identifier {} with rule '{}'".format(layer_ident, ident, refinement))
         self.layers[layer_ident].add_refinement(ident,refinement)
         return lark.Discard

### 2022-11-14 code/LayeredTypes/tests/layer_implementations/A.py

@@ -1,8 +1,10 @@
+from tests.utils import call_order
+
 def depends_on():
-    print("depends_on called")
     return set()
 
 def typecheck(tree):
+    call_order.append("A")
     return tree
 
 def parse_type(type_str):

### 2022-11-14 code/LayeredTypes/tests/layer_implementations/B.py

@@ -1,0 +1,11 @@
+from tests.utils import call_order
+
+def depends_on():
+    return {"A"}
+
+def typecheck(tree):
+    call_order.append("B")
+    return tree
+
+def parse_type(type_str):
+    return None

### 2022-11-14 code/LayeredTypes/tests/layer_implementations/C.py

@@ -1,0 +1,11 @@
+from tests.utils import call_order
+
+def depends_on():
+    return {"A"}
+
+def typecheck(tree):
+    call_order.append("C")
+    return tree
+
+def parse_type(type_str):
+    return None

### 2022-11-14 code/LayeredTypes/tests/layer_implementations/D.py

@@ -1,0 +1,11 @@
+from tests.utils import call_order
+
+def depends_on():
+    return { "B", "C"}
+
+def typecheck(tree):
+    call_order.append("D")
+    return tree
+
+def parse_type(type_str):
+    return None

### 2022-11-14 code/LayeredTypes/tests/layer_implementations/cycleA.py

@@ -1,6 +1,5 @@
 def depends_on():
-    print("depends_on called")
-    return set()
+    return {"cycleB"}
 
 def typecheck(tree):
     return tree

### 2022-11-14 code/LayeredTypes/tests/layer_implementations/cycleB.py

@@ -1,6 +1,5 @@
 def depends_on():
-    print("depends_on called")
-    return set()
+    return {"cycleA"}
 
 def typecheck(tree):
     return tree

### 2022-11-14 code/LayeredTypes/tests/test_code/test_layers.py

@@ -1,0 +1,81 @@
+import unittest
+from graphlib import CycleError
+
+from tests.utils import parse_file, get_compiler, full_path, call_order
+
+class TestLayerDefinitions(unittest.TestCase):
+
+    def test_single_definition(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/single_layer_def.fl")
+        compiler.typecheck(src_file)
+
+        self.assertEqual(len(compiler.layers), 1)
+        self.assertTrue("base" in compiler.layers)
+
+    def test_multiple_definitions(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/multiple_layers.fl")
+        compiler.typecheck(src_file)
+
+        self.assertEqual(len(compiler.layers), 2)
+        self.assertTrue("base" in compiler.layers)
+        self.assertTrue("other" in compiler.layers)
+
+    def test_missing_layer(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/non_existant_layer.fl")
+        with self.assertRaises(RuntimeError):
+            compiler.typecheck(src_file)
+        pass
+
+
+class TestLayerDependencies(unittest.TestCase):
+
+    def test_simple_dependency(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/simple_dependency.fl")
+
+        call_order.clear()
+
+        compiler.typecheck(src_file)
+
+        self.assertEqual(len(compiler.layers), 2)
+        self.assertTrue("A" in compiler.layers)
+        self.assertTrue("B" in compiler.layers)
+
+        self.assertEqual(call_order, ["A", "B"])
+        pass
+
+    def test_multiple_dependencies(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/proper_dependency.fl")
+
+        call_order.clear()
+
+        compiler.typecheck(src_file)
+
+        self.assertEqual(len(compiler.layers), 4)
+        self.assertTrue("A" in compiler.layers)
+        self.assertTrue("B" in compiler.layers)
+        self.assertTrue("C" in compiler.layers)
+        self.assertTrue("D" in compiler.layers)
+
+        self.assertIn(call_order, [["A", "B", "C", "D"],["A", "C", "B", "D"]])
+
+    def test_circular_dependency(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/create_cycle.fl")
+
[…]

### 2022-11-14 code/LayeredTypes/tests/utils.py

@@ -2,18 +2,22 @@ import os
 
 from compiler.Compiler import LayeredCompiler
 
+call_order = []
+
+def full_path(file_path):
+    return os.path.dirname(os.path.realpath(__file__)) + f"{file_path}"
 
 def parse_file(file_path):
     compiler = get_compiler()
 
-    src_path = os.path.dirname(os.path.realpath(__file__)) + f"{file_path}"
+    src_path = full_path(file_path)
     tree = compiler.parse(src_path)
 
     return tree
 
 def get_compiler():
-    impl_path = os.path.dirname(os.path.realpath(__file__)) + "/implementations.py"
-    layer_path = os.path.dirname(os.path.realpath(__file__)) + "/layer_implementations"
+    impl_path = full_path("/implementations.py")
+    layer_path = full_path("/layer_implementations")
     compiler = LayeredCompiler(impl_path, layer_path)
 
     return compiler

### 2022-11-14 paper/code_examples/lib/Main.hs

@@ -13,17 +13,19 @@ ensureNonEmpty :: [Int] -> [Int]
 ensureNonEmpty [] = [0]
 ensureNonEmpty x = x
 
-i1 :: Int
--- i1 = myHead []                    -- Not allowed
-i1 = myHead (ensureNonEmpty [])   -- Allowed
+i1 :: [Int]
+i1 = [2,5,6,3,2]                    -- Not allowed
+-- i1 = myHead (ensureNonEmpty [])   -- Allowed
 
 {-@ type Ordered a = [a]<{\x v -> x <= v}> @-}
 -- Not possible to arbitrarly combine NonEmpty and Ordered into one refinement type
-{-@ type NonEmptyOrdered a = {v:[a]<{\x v -> x >= v}> | len v > 0 }@-}
+{-@ type OrderedList a = {v:[a]<{\x v -> x >= v}> | len v > 0 }@-}
 
 -- Extracting the maximum value from a non-empty and ordered list is equivalent to head
-{-@ maxVal :: NonEmptyOrdered Int -> Int @-}
+{-@ maxVal :: OrderedList Int -> Int @-}
 maxVal :: [Int] -> Int
 maxVal (x:_) = x
+
+r = maxVal i1 
 
 main = putStrLn "Hello, Worldi!"

### 2022-11-14 code/LayeredTypes/compiler/Compiler.py
* Implemented loading of layers that are only referenced implicitly (Due to depends_on)

@@ -1,10 +1,12 @@
 import graphlib
 import os.path
+from warnings import warn
 
 import lark
 from pathlib import Path
 
 from compiler.transformers.RemoveTokens import RemoveTokens
+from layers.Layer import Layer
 from layers.LayerImplWrapper import LayerImplWrapper
 from compiler.transformers.CollectLayers import CollectLayers
 from compiler.transformers.CheckCF import CheckCF
@@ -53,6 +55,7 @@ class LayeredCompiler:
             self.layers[layer_id] = LayerImplWrapper(self.layer_base_dir, lv.layers[layer_id])
             layer_graph[layer_id] = self.layers[layer_id].depends_on()
 
+
         # Check for cycles
         try:
             topo_order = graphlib.TopologicalSorter(layer_graph).static_order()
@@ -61,6 +64,11 @@ class LayeredCompiler:
 
         # Incrementally typecheck each layer based on their topological order
         for layer_id in topo_order:
+            # Dynamically load layer if not already loaded
+            if not layer_id in self.layers:
+                warn(f"Layer {layer_id} was not loaded during CollectLayers, but is required by at least one other layer. Loading it now.")
+                self.layers[layer_id] = LayerImplWrapper(self.layer_base_dir, Layer(layer_id))
+
             tree = self.layers[layer_id].typecheck(tree)
 
         return tree

### 2022-11-14 code/LayeredTypes/tests/layer_implementations/other.py

@@ -1,0 +1,8 @@
+def depends_on():
+    return set()
+
+def typecheck(tree):
+    return tree
+
+def parse_type(type_str):
+    return None

### 2022-11-14 code/LayeredTypes/tests/test_code/test_layers.py

@@ -25,7 +25,7 @@ class TestLayerDefinitions(unittest.TestCase):
     def test_missing_layer(self):
         compiler = get_compiler()
         src_file = full_path("/test_code/layers/non_existant_layer.fl")
-        with self.assertRaises(RuntimeError):
+        with self.assertRaises(FileNotFoundError):
             compiler.typecheck(src_file)
         pass
 
@@ -74,8 +74,13 @@ class TestLayerDependencies(unittest.TestCase):
         compiler = get_compiler()
         src_file = full_path("/test_code/layers/implicit_dependency.fl")
 
-        compiler.typecheck(src_file)
+        call_order.clear()
+
+        with self.assertWarns(UserWarning):
+            compiler.typecheck(src_file)
 
         self.assertEqual(len(compiler.layers), 2)
         self.assertTrue("A" in compiler.layers)
         self.assertTrue("B" in compiler.layers)
+
+        self.assertEqual(call_order, ["A", "B"])

### 2022-11-14 code/LayeredTypes/tests/test_code/test_layers.py
* Added test case for implicit dependency cycle

@@ -84,3 +84,10 @@ class TestLayerDependencies(unittest.TestCase):
         self.assertTrue("B" in compiler.layers)
 
         self.assertEqual(call_order, ["A", "B"])
+
+    def test_implicit_cycle(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/implicit_cycle.fl")
+
+        with self.assertRaises(CycleError):
+            compiler.typecheck(src_file)

### 2022-11-14 code/LayeredTypes/compiler/Compiler.py
* Fixed cycle error handling * Added test case for layers required by implicit layers

@@ -51,24 +51,33 @@ class LayeredCompiler:
 
         layer_graph = {}
         self.layers = {}
+
+        implicit_layers = set()
+
         for layer_id in lv.layers:
             self.layers[layer_id] = LayerImplWrapper(self.layer_base_dir, lv.layers[layer_id])
-            layer_graph[layer_id] = self.layers[layer_id].depends_on()
+            required_layers = self.layers[layer_id].depends_on()
+            implicit_layers.update(required_layers - set(lv.layers.keys()))
+            layer_graph[layer_id] = required_layers
 
+        while len(implicit_layers) > 0:
+            layer_id = implicit_layers.pop()
+            if not layer_id in self.layers:
+                warn(f"Layer {layer_id} was not loaded during CollectLayers, but is required by at least one other layer. Loading it now.")
+                self.layers[layer_id] = LayerImplWrapper(self.layer_base_dir, Layer(layer_id))
+                required_layers = self.layers[layer_id].depends_on()
+                implicit_layers.update(required_layers - set(self.layers.keys()))
+                layer_graph[layer_id] = required_layers
 
         # Check for cycles
         try:
-            topo_order = graphlib.TopologicalSorter(layer_graph).static_order()
+            topo_order = [n for n in graphlib.TopologicalSorter(layer_graph).static_order()]
         except graphlib.CycleError as e:
-            raise RuntimeError(f"Cycle in layer dependencies: {e.args[0]}")
+            raise graphlib.CycleError(f"Cycle in layer dependencies: {e.args[0]}")
 
         # Incrementally typecheck each layer based on their topological order
         for layer_id in topo_order:
             # Dynamically load layer if not already loaded
-            if not layer_id in self.layers:
-                warn(f"Layer {layer_id} was not loaded during CollectLayers, but is required by at least one other layer. Loading it now.")
-                self.layers[layer_id] = LayerImplWrapper(self.layer_base_dir, Layer(layer_id))
-
             tree = self.layers[layer_id].typecheck(tree)
 
         return tree

### 2022-11-14 code/LayeredTypes/tests/test_code/test_layers.py

@@ -89,5 +89,23 @@ class TestLayerDependencies(unittest.TestCase):
         compiler = get_compiler()
         src_file = full_path("/test_code/layers/implicit_cycle.fl")
 
-        with self.assertRaises(CycleError):
+        with self.assertRaises(CycleError) as e:
             compiler.typecheck(src_file)
+
+    def test_implicit_multiple_layers(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/implicit_multiple_layers.fl")
+
+        call_order.clear()
+
+        with self.assertWarns(UserWarning):
+            compiler.typecheck(src_file)
+
+        self.assertEqual(len(compiler.layers), 4)
+        self.assertTrue("A" in compiler.layers)
+        self.assertTrue("B" in compiler.layers)
+        self.assertTrue("C" in compiler.layers)
+        self.assertTrue("D" in compiler.layers)
+
+        self.assertIn(call_order, [["A", "B", "C", "D"], ["A", "C", "B", "D"]])
+

### 2023-01-25 code/LayeredTypes/compiler/transformers/CreateAnnotatedTree.py
* Adapted use case example for file handling

@@ -7,8 +7,7 @@ from lark.tree import Meta
 
 class AnnotatedTree(lark.Tree):
     def __init__(self, data: str, children: 'List[Branch[_Leaf_T]]', meta: Optional[Meta] = None):
-        super().__init__(data, children)
-        self._meta = meta
+        super().__init__(data, children, meta)
         self.__annotations = dict()
 
     def update_layer_annotations(self, layer_identifier: str, identifier: str, values: dict):
@@ -28,7 +27,7 @@ class AnnotatedTree(lark.Tree):
 
     def get_all_layer_annotations(self, layer_id: str, identifier: str):
         if layer_id not in self.__annotations:
-            return None
+            return {}
 
         return self.__annotations[layer_id][identifier]
 

### 2023-01-25 code/LayeredTypes/examples/python_impls/example.py

@@ -1,11 +1,40 @@
-from os import O_RDWR
-
 from builtins import len, print, open
+from pathlib import Path
 
-def openFile(filename:str):
-    return open(filename, "r")
+class FileDescriptor:
+    def __init__(self, filename):
+        self._filename = filename
+        self._filehandle = None
 
-def closeFile(filehandle):
+    def open(self):
+        if self._filehandle is not None:
+            raise Exception("File already open")
+        self._filehandle = open(self._filename, "r")
+
+    def close(self):
+        if self._filehandle is None:
+            raise Exception("File not open")
+        self._filehandle.close()
+        self._filehandle = None
+
+    def read(self):
+        if self._filehandle is None:
+            raise Exception("File not open")
+
+        # Check if the file still has data to read
+        pos = self._filehandle.tell()
+        if pos == self._filehandle.seek(0, 2):
+            raise Exception("File has no more data to read")
+
+        # Seek back to the previous position
+        self._filehandle.seek(pos)
+        return self._filehandle.readlines()
+def createFD(filename:str) -> FileDescriptor:
+    return FileDescriptor(filename)
+def openFile(fileHandle: FileDescriptor):
+    fileHandle.open()
+
+def closeFile(filehandle: FileDescriptor):
     filehandle.close()
 def strToInt(s:str) -> int:
     return int(s)
@@ -14,5 +43,5 @@ def getLine(lines:list, i:int) -> str:
     return lines[i]
 
 # Create a function that takes an open filehandle and returns a list of lines
-def readLines(filehandle) -> list:
-    return filehandle.readlines()
+def readLines(filehandle: FileDescriptor) -> list:
+    return filehandle.read()

### 2023-01-25 code/LayeredTypes/layer_implementations/state.py

@@ -8,9 +8,6 @@ from compiler.transformers.CreateAnnotatedTree import AnnotatedTree
 
 
 class ArgumentState:
-    def __init__(self):
-        self.required_states = set()
-        self.state_transitions = set()
 
     def __init__(self, state: str):
         self.required_states = set()
@@ -65,7 +62,7 @@ class StateLayer(lark.visitors.Interpreter):
 
     def __annotate_states(self, tree):
         if not isinstance(tree, AnnotatedTree):
-            tree = AnnotatedTree(tree)
+            tree = AnnotatedTree(tree.data, tree.children, tree.meta)
 
         for identifier in self.__states:
             tree.add_layer_annotation("state", identifier, "state", self.__states[identifier])

### 2023-01-25 code/LayeredTypes/layer_implementations/typecheck.py

@@ -10,17 +10,6 @@ def depends_on():
 def typecheck(tree):
     class Typechecker(lark.visitors.Interpreter):
         def __init__(self):
-            self.__convertable_types = {
-                "int" : {"long"},
-                "short" : {"int", "long"},
-                "long" : set(),
-                "byte" : {"short", "int", "long"},
-                "float" : {"double"},
-                "double" : set(),
-                "bool" : set(),
-                "str" : set()
-            }
-
             self.__subtype_graph = igraph.Graph(2, directed=True)
             self.__type_ids = {"__num":0,"__logical":1}
 
@@ -45,6 +34,10 @@ def typecheck(tree):
             # Special handling for assignment of num literals
             if t1 == "__num":
                 return self.__is_convertable(t2, "__num")
+
+            # Special handling for assignment of logical literals
+            if t1 == "__logical":
+                return self.__is_convertable(t2, "__logical")
 
             if t1 not in self.__type_ids or t2 not in self.__type_ids:
                 return False
@@ -123,10 +116,10 @@ def typecheck(tree):
             return "__num"
 
         def true(self, tree):
-            return "bool"
+            return "__logical"
 
         def false(self, tree):
-            return "bool"
+            return "__logical"
 
         def bin_op(self, tree):
             lhs_type = self.visit(tree.children[0])
@@ -145,13 +138,13 @@ def typecheck(tree):
 
             if op in {"==", "!=", "<", ">", "<=", ">="}:
                 if self.__is_convertable(lhs_type, rhs_type) or self.__is_convertable(rhs_type, lhs_type):
-                    return "bool"
+                    return "__logical"
 
                 raise TypeError(f"{tree.meta.line}:{tree.meta.column}: Cannot compare types {lhs_type} and {rhs_type}")
 
             if op in {"&&", "||"}:
-                if lhs_type == "bool" and rhs_type == "bool":
-                    return "bool"
+                if self.__is_convertable(lhs_type, "__logical") and self.__is_convertable(rhs_type, "__logical"):
+                    return "__logical"
 
                 raise TypeError(f"{tree.meta.line}:{tree.meta.column}: Cannot perform logical operation on types {lhs_type} and {rhs_type}")
 

### 2023-01-25 code/LayeredTypes/layer_implementations/types.py

@@ -14,7 +14,7 @@ def typecheck(tree):
 
         def __annotate_types(self, tree):
             if not isinstance(tree, AnnotatedTree):
-                tree = AnnotatedTree(tree)
+                tree = AnnotatedTree(tree.data,tree.children,tree.meta)
 
             for var_type in self.variable_types:
                 tree.add_layer_annotation("types",var_type,"type",self.variable_types[var_type])

### 2023-01-25 paper/paper.tex
* Finished paper chapter on file handling use case

@@
 
-Current Liquid Type implementations limit us to only defining one liquid type for variables, arguments or return types.
+% Current Liquid Type implementations limit us to only defining one liquid type for variables, arguments or return types.
 This forces us to combine all properties we want to statically check and verify into one big predicate.
@@
 Additionally previously declared liquid types may not be reused when wanting to refine them further.
+Current verification approaches usually focus on verifying a single property of a program at a time.
+However, in practice, one might needs to verify different properties of the program for different purposes.
+In current implementations, it is not trivially possible to combine these different verification passes.
+One specific examplefor this can be found when looking at current implemetnations of Liquid Types.
+In these, one can only define one liquid type for variables, arguments or return types.
+This forces us to combine all properties we want to statically check and verify into one joint predicate.
+Using such an approach has the drawback that all properties are checked simultaneously which might make it difficult for a developer to identify due to which property a type error occurs.
+Additionally previously declared liquid types may not be reused when wanting to refine them further.
 
-We therefore propose an implementation of \textit{Layered Liquid\todo{Just Layered or Layered Liquid?} Types}, that allows to define types as a conjunction of previously defined Liquid Types.
-These types are then checked iteratively to ensure compability between functions and arguments.
+With this paper, we therefore propose a new incremental verification approach that we will refer to as \textit{Layered Types}.
+The goal of this approach is it for a program to be possible to define individual layers for the verification process that are verified individually.
+By defining dependencies between those layers, it is possible to define a verification order that ensures that all layers are verified in the correct order.
+This allows us to define different properties of a program that can be verified independently from each other.
 
-Assume we have a function \texttt{f} that takes as input a single liquid type $T^{in}$ refined by the predicates $p^{in}_1,\cdots,p^{in}_n$ and result type \texttt{out} and a variable \texttt{arg} of liquid type $T^{arg}$ refined by the predicates $p^{arg}_1,\cdots,p^{arg}_m$:
-
-\begin{lstlisting}[mathescape=true]
-	-- Liquid Type Definitions
-	{-@ type $T^{in}$ t = { t | $p^{in}_1 \wedge ... \wedge p^{in}_n$  } @-}
-	{-@ type $T^{arg}$ t = { t | $p^{arg}_1 \wedge ... \wedge p^{arg}_m$ } @-}
-
-	f : $T^{in}$ -> out 
-	arg : $T^{arg}$
-\end{lstlisting}
-
-When now calling \texttt{f} with \texttt{arg} as an argument, current implementations of liquid types will check for compability of the arguments in a manner similar to:
-
-$$ \left( \bigwedge^{m}_{i=1} p^{arg}_i \right) \implies \left( \bigwedge^{n}_{j=1} p^{in}_j \right) $$
-
-In our layered approach we rather propose that we can define separate layered types for each predicate and easily use their conjunction as the specific refined type for arguments and values:
-
-\begin{lstlisting}[mathescape=true]
-	-- Liquid Type Definitions for inputs to f
-	{-@ type $T^{in}_1$ t = { t | $p^{in}_1$ } @-}
-	{-@ type ... @-}
-	{-@ type $T^{in}_n$ t = { t | $p^{in}_n$ } @-}
-
-	-- Liquid Type Definitions for arg
-	{-@ type $T^{arg}_1$ t = { t | $p^{arg}_1$ } @-}
-	{-@ type ...  @-}
-	{-@ type $T^{arg}_m$ t = { t | $p^{arg}_m$ } @-}
-
-	f : {$T^{in}_1 \wedge ... \wedge T^{in}_n$} -> out 
-	arg : {$T^{arg}_1 \wedge ... \wedge T^{arg}_m$}
-\end{lstlisting}
-
-In this layered approach predicates are not checked simultaneously but rather in an iterative way, which could intuitively rather be expressed as the (which is semantically equivalent to the version above):
-
-$$ \bigwedge^{n}_{j=1} \left( \bigwedge^{m}_{i=1} p^{arg}_i \implies  p^{in}_j \right) $$
+While in this paper, we will focus on specific Use-Cases for Liquid Types, we believe that the proposed approach can be applied to other verification techniques as well.
+For example, we believe that the proposed approach can be extended to verify all sorts of different properties and typesystems such as e.g.
+\textit{Linear Types} \cite{linear_types} or \textit{Dependent Types} \cite{dependent_types}.
 
@@
 While gradual approaches might still type-check with incomplete typing information our approach would fail to typecheck in such a scenario as it still requires complete information.
+However, we believe it would be possible to embed certain gradual verification techniques as independent layers in our approach.
 
@@
 	v \in \texttt{Var}\ 	&::=\ id \\
-	f \in \texttt{Fun}\ 	&::=\ id_1\ id_2\ \{\ s\ \}\\
-	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e) \\
+	f \in \texttt{Fun}\ 	&::=\ id_1\ ...\ id_n\ \{\ s\ \}\\
+	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e)\ |\ f(e_1,\ ...,\ e_n)  \\
 	\texttt{CustomExpr}\ 	&::=\ \texttt{String}\\
@@
 	\texttt{Assign}\ 	&::=\ v\ :=\ e\ |\ v\ :=\ \texttt{CustomExpr}\ \\
-	s \in \texttt{Stmt}\ 	&::=\ \texttt{Assign}\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ \ell\ |\ s_1\ s_2\ \\
+	s \in \texttt{Stmt}\ 	&::=\ \texttt{Assign}\ |\ \texttt{if}\ v\ \texttt{then}\ \lbrace\ s_1\ \rbrace\ \texttt{else}\ \lbrace\ s_2\ \rbrace\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ \lbrace\ s\ \rbrace\ |\ \ell\ |\ s_1\ s_2\ \\
 	\ell \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
@@
 
-These layers can then be used to refine an identifier.
+These layers can then be used to define properties.
+Depending on the Layers' implementation, these properties can apply to identifiers or can be used to define properties of the type system the layer defines.
 In a very basic and minimal approach these layers can be used to annotate types to variables and functions.
@@
 The second layer \textit{liquid} is a refinement layer that refines that the input should be greater or equal to zero and that the output will be greater than zero.
+The implementations of the \textit{base} and \textit{liquid} layer are not part of our language and have to be implemented independently.
 
@@
 The implementation of our language will be done in the \texttt{Python} programming language.
-Roughly speaking, the implementation will consist of three parts, which we will describe throughout this section:
+Generally the implementation can be divided into three individual parts, which we will describe throughout this section:
 
@@
 
-After initially parsing the AST, it will contain explicit nodes for each of the defined layers.
-As a first processing step we use a Transformer to collect these layers and remove the corresponding nodes from the AST.
-The implementation of each node has to be provided by the user as a separate Python module.
-Our language Interpreter will then dynamically load the corresponding module and use it to check the layer.
-To interact with our layered type checker correctly the module has to implement the following functions:
+After parsing the AST there are a few simple checks that are run by an initial interpreter pass.
+Namely, our framework will check that variables are assigned to at least once before they are used in another expression, it will make sure that \texttt{let} does not redeclare an already existing variable.
+Additionally, function names cannot be defined twice and the framework will also check that functons that are called have either been defined previously in the source file or alternatively are defined externally in Python.
+
+After this initial check has passed, the main part of our work, the layered type checker, will start processing the layers.
+A "Layer" here refers to a custom implementation that has to be defined in a separate Python file by the user.
+Based on the defined layer name, our framework will dynamically load the python file as an independent module and will use it to run the type-checker for this layer.
+Our framework supports the definition of the following functions for each layer:
 
 \begin{itemize}
-	\item \texttt{depends\_on} -- This function returns a list of all layers that this layer depends on. This is used to ensure that the layers are checked in the correct order.
+	\item \texttt{depends\_on} -- This function returns a list of strings.
+Each string represents a name of a layer that this layer depends on and thus has to be checked before this layer can be run.
+This is used to ensure that the layers are checked in the correct order.
+	\item \texttt{run\_before} -- This function is similar to \texttt{depends\_on} but returns a list of layers that must not run before this layer has been checked corretly.
+Together with the \texttt{depends\_on} function this allows to flexibly insert layers into any step of the type-checking process.
 	\item \texttt{parse\_type} -- This function may be used to define and parse custom data types that are defined usinf the $CustomExpr$ rule in the grammar.
-\todo{Implementation details -- When and how is this function called?}
-	\item \texttt{check\_layer} -- This function is used to check the layer.
+\todo[inline]{This is a functionality we discussed in the beginning, that would be nice.
+However for our current use cases we do not really need it.
+Should we keep it in the framework or just omit it?}
+	\item \texttt{typecheck} -- This function is used to check the layer.
 It is passed the AST of the program with all the annotated information from the previous layers.
-It should return a boolean value that indicates whether the layer is valid or not.
+It should return the annotated tree with the new layer information added.
 \end{itemize}
@@
 Using the \texttt{depends\_on} function, the layered type checker will first build a dependency graph of all layers.
-It will then topologically sort the layers and check them in the correct order.
-Once all layers have been checked successfully, the interpreter will execute the program.
+By using the dependency graph we can calculate the topological order of layers that can be checked in the correct order.
+Additionally this also allows to check independent layers in parallel.
+Only if all layers have been checked successfully, the program will be executed.
 
-Using this architecture, we provide a flexible framework that allows to easily extend our toy language with various type checking layers.
+In order to communicate annotated information between the different layers, we replace the \texttt{Tree} \Lark{} class with our own \texttt{AnnotatedTree} class.
+This class allows us to store a dictionary of annotations for each node in the tree.
+The annotations are stored as a dictionary of dictionaries.
+The first key of the dictionary is the name of the layer, the second key is the name of the annotation.
+This allows us to store multiple annotations for each layer.
+The \texttt{AnnotatedTree} class also provides a simple interface to access and add annotations which can be used by developers implementing new layers:
+
+\begin{itemize}
+	\item \texttt{get\_layer\_annotation(layer\_id, identifier, key)} -- This function returns the annotation with the given key for the given identifier in the given layer.
+If no annotation is found, it returns \texttt{None}.
+	\item \texttt{get\_all\_layer\_annotations(layer\_id, identifier)} -- This function returns a dictionary of all annotations for the given identifier in the given layer.
+If no annotations are found, it returns an empty dictionary.
+	\item \texttt{add\_layer\_annotation(layer\_id, identifier, key, value)} -- This function adds a annotation with the given key for the given identifier in the given layer to the given value.
+	\item \texttt{update\_layer\_annotations(layer\_id, identifier, values)} -- This updates the layer annotations for the given identifier in the given layer with the given dictionary of values.
+\end{itemize}
+
+Using this architecture, we provide a flexible framework that allows to easily extend our toy language with different layers that can be used to implement different type systems.
 Similar to passes in a compiler, the layers can be added and removed easily.
-Additionally, the layers can be easily combined to create more complex type checking layers.
+Additionally, the layers can be easily combined to create more complex type systems.
 
-\subsection{Examples}
-\label{ssec:examples}
+\section{Example Use-Cases}
+\label{ssec:use_cases}
+For the scope of this paper, we have developed two exemplary use-cases for our framework that demonstrates our layered verification approach.
+In the first use-case, we defined a simple program that uses multiple different, partly independent type systems.
+For our second example, we use our framework to present a more flexible approach to using Liquid Types.
+
+\subsection{Use Case 1 -- File Processing}
+\label{sssec:use_case_1}
+
+In our first use case we implemented a small program that processes the contents of a text file.
+Througout this section we will describe the program and the implemented layers required to realize this program.
+Lastly we will discuss the findings and benefits we gained from using our layered verification approach.
+
+\subsubsection{Description}
+In our first use-case we design a small program that reads contents from a file and performs certain operations based on that.
+The base code for our example can be found in Listing \ref{lst:uc1_code}.
+In our example, the file to be read will contain one number per each line.
+Depending on whether the number is positive or negative, we add the number as is or the square of the number to the resulting value.
+
+\begin{lstlisting}
+processLine lines idx length {
+    end := idx == length
+
+    if end then {
+        0
+    } else {
+
+        curLine := getLine(lines, idx)
+
+        number := strToInt(curLine)
+
+        greaterThanZero := number > 0
+
+        if greaterThanZero then {
+            number + processLine(lines, idx + 1, length)
+        } else {
+            (number * number) + processLine(lines, idx + 1, length)
+        }
+
+    }
+}
+
+processLines lines {
+    length := len(lines)
+
+    idx := 0
+    processLine(lines, idx, length)
+}
+
+fileName := "numbers.data"
+fileHandle := openFile(fileName)
+lines := readLines(fileHandle)
+closeFile(fileHandle)
+total := processLines(lines)
+
+print(total)
+total
+
+\end{lstlisting}
+
+All functions that have not been defined directly \ref{lst:uc1_code} are provided in a separate, external python file.
+As our language on itself does not include a type system and there are no layer annotations in the code above, no type checking will be performed on the code as it is presented.
+
+However even this simple example could already introduce several errors that could be prevented by introducing some static analysis.
+First of all, it does not provide any information about the variable and function type and thus, we cannot ensure that functions are called with arguments of the correct types which might lead to runtime errors.
+
+Additionally we our code interacts with the file it operates on through a file descriptor.
+TThis file descriptor can be viewed as a resource that might have an implicit or explicit internal state.
+In a very simple approach this file descriptor could be in the states "Open" or "Closed".
+In order to read from it, it should be in an "Open" state and additionally we should ensure that by the end of our program the resource has been properly closed again.
+
+In order to improve our example code and allow static analysis, we therefore will introduce three layers:
+\begin{enumerate}
+	\item A \textit{Type} Layer -- This layer will be used to simply define variable and function types
+	\item The \textit{Typechecking} Layer -- This layer will allow to flexibly define a type system and then use the information provided by the \textit{Type} layer to check if the program is well-typed.
+It therefore depends on the \textit{Type} layer to provide the type information.
+	\item A \textit{State} Layer -- This layer will be used to define states on variables.
+Additionally it can model state transitions.
+We will use it to track the state of the file descriptor.
+This layer operates completely independent from the \textit{Type} and \textit{Typechecking} layer.
+\end{enumerate}
+
+\subsubsection{Implemented Layers}
+In the following sections we will explain in more detail how the layers are implemented and what Syntax they use.
+
+\paragraph{Types Layer}
+The \textit{Types} layer is the first layer that is checked.
+It is used to define the types of variables and functions.
+To define the type of a variable we use the \texttt{Layer} syntax of our language.
+A layer definition for the \textit{Types} layer looks as follows:
+
+\begin{lstlisting}
+identifier :: types :: Type
+\end{lstlisting}
+
+The \texttt{identifier} is the name of the variable or function for which we want to define the type.
+The \texttt{types} keyword is used to associate this information with the correct layer.
+The \texttt{Type} is the desired type information to be annotated.
+There are no restrictions on how a \texttt{Type} should look, as the types layer simply annotates the given string into the current node and does not further process it.
+For example to define the type of a variable \texttt{foo} as \texttt{int} we would write:
+
+\begin{lstlisting}
+foo :: types :: int
+\end{lstlisting}
+
+In order to define the type of a function, there is a slight adaption to this syntax.
+While the first two parts are identical to the variable definition, we now need to provide type information for each argument of the function.
+This is done by using the \texttt{Type} syntax for each argument and then separating them with an arrow symbol \texttt{->}.
+The return type of the function is then defined by the last \texttt{Type}.
+If a function does not take an argument the defintion simply starts with the arrow symbol.
+The following code piece contains an example for a function \texttt{f} that takes two arguments of type \texttt{int} and returns a value of type \texttt{bool}, and a function \texttt{g} that does not take any arguments and returns a value of type \texttt{string}:
+
+\begin{lstlisting}
+f :: types :: int -> int -> bool
+g :: types :: -> string
+\end{lstlisting}
+
+It is important to note that the \textit{Types} layer does not perform any type checking.
+It simply annotates the given type information into the AST.
+The \textit{Types} layer only checks for the following inconsistencies:
+
+\begin{itemize}
+	\item The type for a variable can only be defined once in the current scope.
+	\item When encountering a function definition, the type for it needs to be defined before the function signature.
+Additionally the number of arguments in the function signature needs to match the number of arguments in the type definition.
+\end{itemize}
+
+Any further type checking is performed by the \textit{Typechecking} layer.
+
+\paragraph{Typechecking Layer}
+
+The \textit{Typechecking} layer is used to define a type system and then use the information provided by the \textit{Types} layer to check if the program is well-typed.
+It therefore depends on the \textit{Types} layer to provide the type information.
+In our implementation of the type checker, we do not impose any restrictions on the specific type system.
+We rather provide a generic type checking implementation that allows the user to flexibly define their own types and subtype relationships.
+However, since our language implicitly natively requires specific expressions to be of a number or boolean type, there are implicitly part of our type system.
+
+To define the specific type system, we use the \texttt{Layer} syntax of our language.
+However, instead of defining layers for identifiers as in the \textit{Types} layer, we reserve a selected set of keywords to define the type system.
+Depending on the definition provided by the user, the type checker will internally build a type graph that is used to perform the type checking.
+
+The following keywords are reserved and can be used to define parts of the type system:
+
+\begin{itemize}
+	\item \texttt{logicalTypes} -- This keyword is used to define a set of logical types. Multiple types can be defined by separating them with a space.
+	\item \texttt{numTypes} -- This keyword is used to define a set of numeric types. Multiple types can be defined by separating them with a space.
+	\item \texttt{subtype} -- This keyword is used to define a subtype relationship between two or more types.
+The syntax for a subtype relationship is \texttt{<type1> <: <type2> <: ...
+<: <typeN>}.
+The subtype relationship is then defined by the order of the types.
+The first type is the supertype and the following type is a subtype of the previous type.
+\end{itemize}
+
+The following code piece is a short demonstration of a type system containing different logical types, numeric types and subtype relationships:
+
+\begin{lstlisting}
+logicalTypes :: typecheck :: bool
+numTypes :: typecheck :: int float
+subtype :: typecheck :: float <: int
+\end{lstlisting}
+
+The type system defined by the user will be used to perform the typechecking.
+To perform this typecheck the type checker accesses the annotations provided by the \textit{Types} layer.
+For the case that a type is neither a numeric or logical type and it does not have any subtype relationships, it is not required to explicitly define it in the type system.
+The type checker will then simply assume the existance of such a type that is not a subtype of any other type nor does it have any subtypes.
+
+Our type checker will perform the following checks:
+
+\begin{itemize}
+	\item The type of each variable needs to be defined.
+	\item When assigning a value to a variable, the type of the value needs to be a subtype of the type of the variable.
+	\item When performing a logical operations (e.g. \texttt{\&\&}, \texttt{||}) the type of the operands needs to be a logical type. The result of the operation will be a logical type.
+	\item When performing a numeric operation (e.g.
+\texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}) the type of the operands needs to be a numeric type.
+The result of the operation will be the most general numeric type of the operands.
+	\item When performing a comparison operation (e.g.
+\texttt{==}, \texttt{!=}, \texttt{<}, \texttt{>}, \texttt{<=}, \texttt{>=}) the type of the operands needs to be compatible.
+This means that either the right hand side is a subtype of the left hand side or the left hand side is a subtype of the right hand side.
+The result of such an operation is always a boolean value.
+	\item In an \texttt{if} statement the condition needs to be a logical type.
+	\item When performing a function call, the number of arguments needs to match the number of arguments in the function definition.
+Additionally the type of each argument needs to be a subtype of the type of the corresponding argument in the function definition.
+\end{itemize}
+
+Since we require the type of each variable to be defined, this means we require full information about the types of all variables.
+We do not perform any type inference.
+This is a deliberate design decision, as we want to keep the language as simple as possible.
+However, by the design of our language, it would be possible to develop an additional layer that runs after the \textit{Types} layer but before the \textit{Typechecking} layer that performs a type inference based on the incomplete information provided.
+
+\paragraph{State Layer}
+
+The \textit{State} layer is used to define states on variables.
+Similarly, we can use it to check if a variable has a certain state and define state transitions.
+This can for example be used to track the state of resources that are used by a program.
+In our implementation, the state of a variable is completely independent of its type, therefore it does not require any dependencies on the \textit{Types} layer.
+
+The state of a variable is defined by a set of strings.
+Generally there are two ways to check the state of a variable or transition it's state to a new one.
+Firstly, we can explicitly do this by using a layer statement of the language.
+Secondly, we can also impose state requirements on arguments of function calls.
+Then when passing a variable to a function, the layer will check if the variable has the required state and apply the state transition if there are any.
+
+Similar to the \textit{Types} layer, the \textit{State} layer uses the \texttt{Layer} syntax of the language to define the state of a variable.
+The first part of the layer statement defines the variable or function name to define a state on, while the second part is simply the layer identifier \texttt{state}.
+The last part of the layer definition defines the state.
+
+A state definition is enclosed by curly brackets and can contain the following parts:
+
+\begin{itemize}
+	\item A state \textit{requirement} -- Imposes a restrition on the current state the variable needs to have. This is simply a string that needs to be part of the current state of the variable.
+	\item A state \textit{transition} -- Defines a transition to a new state.
+Generally a state transition has the form \texttt{<oldState> => <newState>}.
+However, either the old or new state may be omitted.
+When omitting the old state, this simply means that something is added to the current state of the variable unconditionally.
+When omitting the new state, this means that the corresponding state string will be removed from the set of states.
+\end{itemize}
+
+For our use-case, we want to track the state of the file descriptor.
+Specifically, we want to make sure, that the file handle is open before we can read from it.
+Additionally, we do not want to read from a file handle again once it has already been read.
+Therefore, we define the states \texttt{Open}, \texttt{Consumed} and \texttt{Closed} and define the following state transitions for the functions that interact with the file descriptor:
+
+\begin{lstlisting}
+createFD :: state :: {} -> { Closed }
+openFile :: state :: {Closed => Open} -> {}
+readLines :: state :: {Open => Consumed} -> {}
+closeFile :: state :: {Consumed => Closed} -> {}
+\end{lstlisting}
+
+\subsection{Use Case 2 -- Liquid Types}
+\label{sssec:use_case_2}
 

### 2023-01-27 paper/implementation.tex
* Separated chapters into separate tex files * Finished chapter about first use case * Rewrote parts of the introduction chapter to be more generally about   type systems and not liquid types in particular

--- 2022-10-25 paper/paper.tex
@@
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[acmsmall, review, screen]{acmart}
-
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2022}
-\acmYear{2022}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Conference acronym 'XX]{Make sure to enter the correct
-  conference title from your rights confirmation emai}{June 03--05,
-  2018}{Woodstock, NY}
-%%
-%%  Uncomment \acmBooktitle if the title of the proceedings is different
-%%  from ``Proceedings of ...''!
-%%
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
-
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
-
-%% Packages
-\usepackage{xcolor}
-\usepackage{color}
-\usepackage{listings}
-
-%% Custom Definitions
-\graphicspath{{graph/}}
-\definecolor{ForestGreen}{HTML}{009B55}
-
-\lstset{inputpath=code_examples/lib/Examples}
-\lstset{
-  frame=none,
-  xleftmargin=2pt,
-  stepnumber=1,
-  numbers=left,
-  numbersep=5pt,
-  numberstyle=\ttfamily\tiny\color[gray]{0.3},
-  belowcaptionskip=\bigskipamount,
-  escapeinside={*'}{'*},
-  language=haskell,
-  tabsize=2,
-  emphstyle={\bf},
-  commentstyle=\color{ForestGreen},
-  stringstyle=\mdseries\rmfamily,
-  showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
-  columns=flexible,
-  basicstyle=\small\sffamily,
-  showstringspaces=false,
-  morecomment=[l]\%,
-}
-
-% Shorthand command for Lark
-\newcommand{\Lark}[0]{\textsc{Lark}}
-
-%% CUSTOM PACKAGES - NON-ACCEPTED
-\usepackage{todonotes}
-
-%%
-%% end of the preamble, start of the body of the document source.
-\begin{document}
-
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{LayeredTypes}
-%TODO: Proper Title
-%\title[LayeredTypes]{Layered Types -- A structured approach to arbitrarily combine Liquid Types}
-
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-\author{Lukas Abelt}
-\email{l.abelt@luabelt.de}
-\orcid{1234-5678-9012}
-\affiliation{%
-  \institution{LASIGE}
-  \streetaddress{XXX}
-  \city{Lisbon}
-  \state{}
-  \country{Portugal}
-  \postcode{}
-}
-
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
-\begin{abstract}
-	\todo[inline]{TODO: Abstract}
-\end{abstract}
-
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%% TODO
-\begin{CCSXML}
-<ccs2012>
- <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
-  <concept_significance>500</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
-</ccs2012>
-\end{CCSXML}
-%TODO
-%\ccsdesc[500]{Computer systems organization~Embedded systems}
-%\ccsdesc[300]{Computer systems organization~Redundancy}
-%\ccsdesc{Computer systems organization~Robotics}
-%\ccsdesc[100]{Networks~Network reliability}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-%TODO: Keywords
-\keywords{}
-%% A "teaser" image appears between the author and affiliation
-%% information and the body of the document, and typically spans the
-%% page.
-\begin{teaserfigure}
-  \includegraphics[width=\textwidth]{layered-shots.jpg}
-  \label{fig:teaser}
-  \caption{Layers in Liquids}
-\end{teaserfigure}
-
-\received{20 February 2007}
-\received[revised]{12 March 2009}
-\received[accepted]{5 June 2009}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
-\section{Introduction}
-\subsection{Context}
-Modern program languages such as C++, Haskell, Rust or Java use type systems to ensure certain promises about the correct behavior of an implementation.
-In essence, the type system allows the programer to specify what inputs to accept for a program or function and what outputs it will produce.
-While these guarantees already prevent a multitude of runtime errors, they alone cannot prevent e.g.
-divisions by yero or out of range accesses in arrays.
-
-One approach to bridge this gap is the introduction of \textit{Refinement Types}, also referred to as \textit{Liquid Types}.
-These allow us to annotate the traditional type system with additional logical predicates in order to ensure semantic properties, which can be checked statically at compile time.
-Using these techniques also allows us to define pre- and post-conditions on functions' arguments and return values.
-By properly defining these contracts a developer can clearly communicate and moreover enforce how other developers' code is allowed to interface with their implementations.
-Examples for constraints might include any of the following:
-
-\begin{itemize}
-	\item Ensuring values are in certain bounds
-	\item Ensuring that a list is non-empty
-	\item Ensuring that data structures certain to specific semantic properties e.g. that they are sorted
-\end{itemize}
-
-All in all, refinement types are valuable building blocks to creating reliable software systems.
-Using them can minimize the risk of any unhandled run-time errors occuring which might lead to undesired or even undefined behavior.
-
-\subsection{Motivation}
-
-While Refinement Types already vastly imrpove traditional type systems, current implementations still expose some drawbacks in their practical usage.
-One of these challenges stems from the fact that during the static analysis, each refinement is treated as a single monolithic requirement that needs to be fulfilled.
-However in reality, a refinement might combine different properties of a type that may benefit from being treated individually.
-
-One example for such a refinement in LiquidHaskell is shown in \ref{lst:combined_lt}.
-There, we define a function that extracts the maximum value of a (descendingly) ordered, non-empty list.
-In this example, this operation is equivalent to simply extracting the \texttt{head} of a list.
-While we only define one refinement, it actually operates on two different properties of a list.
-First, it imposes a restriction on it's \texttt{size} by requiring it to be non-empty, while the second restriction imposes an order on the list elements.
-
-\lstinputlisting[firstline=6,caption={Example for a function with a combined Refinement Type in LiquidHaskell},label={lst:combined_lt}]{CombinedLT.hs}
-
-While the refinement concerns two different aspects of the list that are independent from each other, current static analysis implementations of Liquid Types will treat it only as a single combined requirement to be fulfilled.
-However especially developers could benefit from a system that treats them independently: When writing code that interfaces with already existing codes such a separated approach may give a developer more fine-grained information on which requirements they still need to fulfill in their implementation.
-
-Effectively one could treat such different requirements as separate \textit{Layers} that can be verified independently.
-Additionally with a proper implementation using such layers may also result in an implementation benefit, that different Layers can be defined individually and combined arbitrarily, as opposed to the example in Listing \ref{lst:combined_lt} where the combination of the two traits had to be explicitly defined as an refinement on it's own.
-
-One other possible usage of such an iterative approach is that it could allow to more flexibly add different static analysis approaches on top of each other.
-For example \ref{handley2020} have shown how to use liquid types in Haskell to statically perform cost analysis of different resources such as memory consumption or computational complexity.
-In their approach they showed that liquid types can be used to reason about their upper and lower bounds.
-However, in their work they only analysed one resource at a time.
-However, using a layered approach would allow to easily combine these different resource analyses to eg.
-track and reason about both memory usage and computational complexity.
-
-\subsection{Problem Definition}
-
-Current Liquid Type implementations limit us to only defining one liquid type for variables, arguments or return types.
-This forces us to combine all properties we want to statically check and verify into one big predicate.
-Using such an approach has the drawback that all properties are checked simultaneously which might make it difficult for a developer to identify due to which property a type error occurs.
-Additionally previously declared liquid types may not be reused when wanting to refine them further.
-
-We therefore propose an implementation of \textit{Layered Liquid\todo{Just Layered or Layered Liquid?} Types}, that allows to define types as a conjunction of previously defined Liquid Types.
-These types are then checked iteratively to ensure compability between functions and arguments.
-
-Assume we have a function \texttt{f} that takes as input a single liquid type $T^{in}$ refined by the predicates $p^{in}_1,\cdots,p^{in}_n$ and result type \texttt{out} and a variable \texttt{arg} of liquid type $T^{arg}$ refined by the predicates $p^{arg}_1,\cdots,p^{arg}_m$:
-
-\begin{lstlisting}[mathescape=true]
-	-- Liquid Type Definitions
-	{-@ type $T^{in}$ t = { t | $p^{in}_1 \wedge ... \wedge p^{in}_n$  } @-}
-	{-@ type $T^{arg}$ t = { t | $p^{arg}_1 \wedge ... \wedge p^{arg}_m$ } @-}
-
-	f : $T^{in}$ -> out 
-	arg : $T^{arg}$
-\end{lstlisting}
-
-When now calling \texttt{f} with \texttt{arg} as an argument, current implementations of liquid types will check for compability of the arguments in a manner similar to:
-
-$$ \left( \bigwedge^{m}_{i=1} p^{arg}_i \right) \implies \left( \bigwedge^{n}_{j=1} p^{in}_j \right) $$
-
-In our layered approach we rather propose that we can define separate layered types for each predicate and easily use their conjunction as the specific refined type for arguments and values:
-
-\begin{lstlisting}[mathescape=true]
-	-- Liquid Type Definitions for inputs to f
-	{-@ type $T^{in}_1$ t = { t | $p^{in}_1$ } @-}
-	{-@ type ... @-}
-	{-@ type $T^{in}_n$ t = { t | $p^{in}_n$ } @-}
-
-	-- Liquid Type Definitions for arg
-	{-@ type $T^{arg}_1$ t = { t | $p^{arg}_1$ } @-}
-	{-@ type ...  @-}
-	{-@ type $T^{arg}_m$ t = { t | $p^{arg}_m$ } @-}
-
-	f : {$T^{in}_1 \wedge ... \wedge T^{in}_n$} -> out 
-	arg : {$T^{arg}_1 \wedge ... \wedge T^{arg}_m$}
-\end{lstlisting}
-
-In this layered approach predicates are not checked simultaneously but rather in an iterative way, which could intuitively rather be expressed as the (which is semantically equivalent to the version above):
-
-$$ \bigwedge^{n}_{j=1} \left( \bigwedge^{m}_{i=1} p^{arg}_i \implies  p^{in}_j \right) $$
-
-While on first glance this approach could be described as an approach to \textit{Gradual Verification}, we will explicitly refer to our approach as \textit{Layered} to distinguish it from \textit{Gradual Verification} techniques.
-While the latter deal with the gradient between incomplete and complete information (in that case on types), our approach is rather concerned not with what information is available on the typing but rather how it is processed.
-While gradual approaches might still type-check with incomplete typing information our approach would fail to typecheck in such a scenario as it still requires complete information.
-
-\subsection{Impact}
-
-We expect our approach to have the following impacts on Liquid Types in practice:
-
-\paragraph{Improved Error Messages}
-
-One current challenge when using liquid types is that all predicates for a type are checked simultaneously.
-This can make it difficult to identify which predicate caused a type liquid error.
-By separating the predicates into separate \textit{Layers} and verifying them individually, it is possible to precisely identify the predicate causing the type error.
-We expect this to help developers to more easily identify and fix these type errors in their code.
-
-\paragraph{Improved Usability and Reusability}
-
-In current Liquid Type implementations, types that require multiple predicates are defined as a single type.
-This may make it difficult for a developer to directly identify which properties a specific liquid type ensures.
-By separating different properties into a more fine-grained layered approach, this may make it easier to identify these properties.
-Additionally, by separating the predicates into different layers, it is possible to reuse previously defined liquid types in a more flexible way.
-For example, if a developer wants to refine a previously defined liquid type, they can now do so by adding a new layer to it, instead of having to redefine the entire type.
-
-\paragraph{Improved Composability}
-
-In addition to the improved reusability of our approach, we also expect it to improve the composability of different liquid types.
-With our suggested approach, new refinements can simply be created by composing already existing layers.
-This vastly increases modularity of the liquid type approach.
-It also may allow to re-use layers that have already been defined by other developers for ones own implementation.
-
-\subsection{Approach}
-We will start by giving a brief overview of the general principles of refinement types and liquid types in general in section \ref{sec:background}.
-Additionally, we will discuss the current state of the art in liquid types and gradual verification in section \ref{sec:related}.
-
-In section \ref{sec:implementation} we will first give a more detailed description of our approach and define a simple functional language that we will use for further demonstration.
-We will then implement a prototype type checker for our layered approach.
-Using this simple prototype we will showcase how our approach can be used on various simple examples.
-
-In section \ref{sec:evaluation} we will discuss the limitations of our approach and possible extensions.
-Additionally, we will evaluate how our approach compares to using a pure, non-layered, verification approach.
-
-Test
-
-\subsection{Contributions}
-\section{Background}
-\label{sec:background}
-\section{Related Work}
-\label{sec:related}
 \section{Implementation}
@@
 
-An example of a program written in our language can be seen in \ref{lst:fac_example}.
+An example how a program written in our language could look like can be seen in \ref{lst:fac_example}.
 The program defines a factorial function that is refined in two layers.
@@
 Using the \texttt{depends\_on} function, the layered type checker will first build a dependency graph of all layers.
-It will then topologically sort the layers and check them in the correct order.
-Once all layers have been checked successfully, the interpreter will execute the program.
+By using the dependency graph we can calculate the topological order of layers that can be checked in the correct order.
+Additionally this also allows to check independent layers in parallel.
+Only if all layers have been checked successfully, the program will be executed.
 
-\subsection{Examples}
-\label{ssec:examples}
+In order to communicate annotated information between the different layers, we replace the \texttt{Tree} \Lark{} class with our own \texttt{AnnotatedTree} class.
+This class allows us to store a dictionary of annotations for each node in the tree.
+The annotations are stored as a dictionary of dictionaries.
+The first key of the dictionary is the name of the layer, the second key is the name of the annotation.
+This allows us to store multiple annotations for each layer.
+The \texttt{AnnotatedTree} class also provides a simple interface to access and add annotations which can be used by developers implementing new layers:
 
-\section{Evaluation}
-\label{sec:evaluation}
-\section{Conclusion}
-\label{sec:conclusion}
-\section{Acknowledgments}
-\label{sec:acknowledgments}
+\begin{itemize}
+	\item \texttt{get\_layer\_annotation(layer\_id, identifier, key)} -- This function returns the annotation with the given key for the given identifier in the given layer.
+If no annotation is found, it returns \texttt{None}.
+	\item \texttt{get\_all\_layer\_annotations(layer\_id, identifier)} -- This function returns a dictionary of all annotations for the given identifier in the given layer.
+If no annotations are found, it returns an empty dictionary.
+	\item \texttt{add\_layer\_annotation(layer\_id, identifier, key, value)} -- This function adds a annotation with the given key for the given identifier in the given layer to the given value.
+	\item \texttt{update\_layer\_annotations(layer\_id, identifier, values)} -- This updates the layer annotations for the given identifier in the given layer with the given dictionary of values.
+\end{itemize}
 
-\end{document}
-\endinput
-%%
-%% End of file `sample-acmsmall-conf.tex'.
+Using this architecture, we provide a flexible framework that allows to easily extend our toy language with different layers that can be used to implement different type systems.
+Similar to passes in a compiler, the layers can be added and removed easily.
+Additionally, the layers can be easily combined to create more complex type systems.

### 2023-01-27 paper/paper.tex
* Separated chapters into separate tex files * Finished chapter about first use case * Rewrote parts of the introduction chapter to be more generally about   type systems and not liquid types in particular

--- 2022-10-25 paper/paper.tex
@@
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{LayeredTypes}
-%TODO: Proper Title
-%\title[LayeredTypes]{Layered Types -- A structured approach to arbitrarily combine Liquid Types}
+\title[Layered Types]{Layered Types -- A structured approach to arbitrarily combine Type Systems}
 
@@
 %%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-
-%%
 %% The abstract is a short summary of the work to be presented in the
@@
 %% Please copy and paste the code instead of the example below.
-%% TODO
 \begin{CCSXML}
-<ccs2012>
- <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
-  <concept_significance>500</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
-</ccs2012>
 \end{CCSXML}
-%TODO
 %\ccsdesc[500]{Computer systems organization~Embedded systems}
@@
 \maketitle
+\input{introduction}
 
-\section{Introduction}
-\subsection{Context}
-Modern program languages such as C++, Haskell, Rust or Java use type systems to ensure certain promises about the correct behavior of an implementation.
-In essence, the type system allows the programer to specify what inputs to accept for a program or function and what outputs it will produce.
-While these guarantees already prevent a multitude of runtime errors, they alone cannot prevent e.g.
-divisions by yero or out of range accesses in arrays.
+\input{background}
 
-One approach to bridge this gap is the introduction of \textit{Refinement Types}, also referred to as \textit{Liquid Types}.
-These allow us to annotate the traditional type system with additional logical predicates in order to ensure semantic properties, which can be checked statically at compile time.
-Using these techniques also allows us to define pre- and post-conditions on functions' arguments and return values.
-By properly defining these contracts a developer can clearly communicate and moreover enforce how other developers' code is allowed to interface with their implementations.
-Examples for constraints might include any of the following:
+\input{related}
 
-\begin{itemize}
-	\item Ensuring values are in certain bounds
-	\item Ensuring that a list is non-empty
-	\item Ensuring that data structures certain to specific semantic properties e.g. that they are sorted
-\end{itemize}
+\input{implementation}
 
-All in all, refinement types are valuable building blocks to creating reliable software systems.
-Using them can minimize the risk of any unhandled run-time errors occuring which might lead to undesired or even undefined behavior.
+\input{use-cases}
 
-\subsection{Motivation}
+\input{evaluation}
 
-While Refinement Types already vastly imrpove traditional type systems, current implementations still expose some drawbacks in their practical usage.
-One of these challenges stems from the fact that during the static analysis, each refinement is treated as a single monolithic requirement that needs to be fulfilled.
-However in reality, a refinement might combine different properties of a type that may benefit from being treated individually.
+\input{conclusion}
 
-One example for such a refinement in LiquidHaskell is shown in \ref{lst:combined_lt}.
-There, we define a function that extracts the maximum value of a (descendingly) ordered, non-empty list.
-In this example, this operation is equivalent to simply extracting the \texttt{head} of a list.
-While we only define one refinement, it actually operates on two different properties of a list.
-First, it imposes a restriction on it's \texttt{size} by requiring it to be non-empty, while the second restriction imposes an order on the list elements.
-
-\lstinputlisting[firstline=6,caption={Example for a function with a combined Refinement Type in LiquidHaskell},label={lst:combined_lt}]{CombinedLT.hs}
-
-While the refinement concerns two different aspects of the list that are independent from each other, current static analysis implementations of Liquid Types will treat it only as a single combined requirement to be fulfilled.
-However especially developers could benefit from a system that treats them independently: When writing code that interfaces with already existing codes such a separated approach may give a developer more fine-grained information on which requirements they still need to fulfill in their implementation.
-
-Effectively one could treat such different requirements as separate \textit{Layers} that can be verified independently.
-Additionally with a proper implementation using such layers may also result in an implementation benefit, that different Layers can be defined individually and combined arbitrarily, as opposed to the example in Listing \ref{lst:combined_lt} where the combination of the two traits had to be explicitly defined as an refinement on it's own.
-
-One other possible usage of such an iterative approach is that it could allow to more flexibly add different static analysis approaches on top of each other.
-For example \ref{handley2020} have shown how to use liquid types in Haskell to statically perform cost analysis of different resources such as memory consumption or computational complexity.
-In their approach they showed that liquid types can be used to reason about their upper and lower bounds.
-However, in their work they only analysed one resource at a time.
-However, using a layered approach would allow to easily combine these different resource analyses to eg.
-track and reason about both memory usage and computational complexity.
-
-\subsection{Problem Definition}
-
-Current Liquid Type implementations limit us to only defining one liquid type for variables, arguments or return types.
-This forces us to combine all properties we want to statically check and verify into one big predicate.
-Using such an approach has the drawback that all properties are checked simultaneously which might make it difficult for a developer to identify due to which property a type error occurs.
-Additionally previously declared liquid types may not be reused when wanting to refine them further.
-
-We therefore propose an implementation of \textit{Layered Liquid\todo{Just Layered or Layered Liquid?} Types}, that allows to define types as a conjunction of previously defined Liquid Types.
-These types are then checked iteratively to ensure compability between functions and arguments.
-
-Assume we have a function \texttt{f} that takes as input a single liquid type $T^{in}$ refined by the predicates $p^{in}_1,\cdots,p^{in}_n$ and result type \texttt{out} and a variable \texttt{arg} of liquid type $T^{arg}$ refined by the predicates $p^{arg}_1,\cdots,p^{arg}_m$:
-
-\begin{lstlisting}[mathescape=true]
-	-- Liquid Type Definitions
-	{-@ type $T^{in}$ t = { t | $p^{in}_1 \wedge ... \wedge p^{in}_n$  } @-}
-	{-@ type $T^{arg}$ t = { t | $p^{arg}_1 \wedge ... \wedge p^{arg}_m$ } @-}
-
-	f : $T^{in}$ -> out 
-	arg : $T^{arg}$
-\end{lstlisting}
-
-When now calling \texttt{f} with \texttt{arg} as an argument, current implementations of liquid types will check for compability of the arguments in a manner similar to:
-
-$$ \left( \bigwedge^{m}_{i=1} p^{arg}_i \right) \implies \left( \bigwedge^{n}_{j=1} p^{in}_j \right) $$
-
-In our layered approach we rather propose that we can define separate layered types for each predicate and easily use their conjunction as the specific refined type for arguments and values:
-
-\begin{lstlisting}[mathescape=true]
-	-- Liquid Type Definitions for inputs to f
-	{-@ type $T^{in}_1$ t = { t | $p^{in}_1$ } @-}
-	{-@ type ... @-}
-	{-@ type $T^{in}_n$ t = { t | $p^{in}_n$ } @-}
-
-	-- Liquid Type Definitions for arg
-	{-@ type $T^{arg}_1$ t = { t | $p^{arg}_1$ } @-}
-	{-@ type ...  @-}
-	{-@ type $T^{arg}_m$ t = { t | $p^{arg}_m$ } @-}
-
-	f : {$T^{in}_1 \wedge ... \wedge T^{in}_n$} -> out 
-	arg : {$T^{arg}_1 \wedge ... \wedge T^{arg}_m$}
-\end{lstlisting}
-
-In this layered approach predicates are not checked simultaneously but rather in an iterative way, which could intuitively rather be expressed as the (which is semantically equivalent to the version above):
-
-$$ \bigwedge^{n}_{j=1} \left( \bigwedge^{m}_{i=1} p^{arg}_i \implies  p^{in}_j \right) $$
-
-While on first glance this approach could be described as an approach to \textit{Gradual Verification}, we will explicitly refer to our approach as \textit{Layered} to distinguish it from \textit{Gradual Verification} techniques.
-While the latter deal with the gradient between incomplete and complete information (in that case on types), our approach is rather concerned not with what information is available on the typing but rather how it is processed.
-While gradual approaches might still type-check with incomplete typing information our approach would fail to typecheck in such a scenario as it still requires complete information.
-
-\subsection{Impact}
-
-We expect our approach to have the following impacts on Liquid Types in practice:
-
-\paragraph{Improved Error Messages}
-
-One current challenge when using liquid types is that all predicates for a type are checked simultaneously.
-This can make it difficult to identify which predicate caused a type liquid error.
-By separating the predicates into separate \textit{Layers} and verifying them individually, it is possible to precisely identify the predicate causing the type error.
-We expect this to help developers to more easily identify and fix these type errors in their code.
-
-\paragraph{Improved Usability and Reusability}
-
-In current Liquid Type implementations, types that require multiple predicates are defined as a single type.
-This may make it difficult for a developer to directly identify which properties a specific liquid type ensures.
-By separating different properties into a more fine-grained layered approach, this may make it easier to identify these properties.
-Additionally, by separating the predicates into different layers, it is possible to reuse previously defined liquid types in a more flexible way.
-For example, if a developer wants to refine a previously defined liquid type, they can now do so by adding a new layer to it, instead of having to redefine the entire type.
-
-\paragraph{Improved Composability}
-
-In addition to the improved reusability of our approach, we also expect it to improve the composability of different liquid types.
-With our suggested approach, new refinements can simply be created by composing already existing layers.
-This vastly increases modularity of the liquid type approach.
-It also may allow to re-use layers that have already been defined by other developers for ones own implementation.
-
-\subsection{Approach}
-We will start by giving a brief overview of the general principles of refinement types and liquid types in general in section \ref{sec:background}.
-Additionally, we will discuss the current state of the art in liquid types and gradual verification in section \ref{sec:related}.
-
-In section \ref{sec:implementation} we will first give a more detailed description of our approach and define a simple functional language that we will use for further demonstration.
-We will then implement a prototype type checker for our layered approach.
-Using this simple prototype we will showcase how our approach can be used on various simple examples.
-
-In section \ref{sec:evaluation} we will discuss the limitations of our approach and possible extensions.
-Additionally, we will evaluate how our approach compares to using a pure, non-layered, verification approach.
-
-Test
-
-\subsection{Contributions}
-\section{Background}
-\label{sec:background}
-\section{Related Work}
-\label{sec:related}
-\section{Implementation}
-\label{sec:implementation}
-
-In this section we will describe our implementation of a layered typing approach in detail.
-We will start by describing the simple functional language we will use for our examples.
-We will then describe how our layered type checker is implemented.
-Finally, we will showcase how our approach can be used on various simple examples.
-
-\subsection{Language}
-\label{ssec:language}
-
-To demonstrate our layered approach we implemented a simple functional language.
-The syntax of said language is described in \ref{fig:lang}.
-The language is similar to most simple functional languages and generally follows the reduction semantics of a standard Hindley-Milner type system.
-The only data types available in the language are the \texttt{Int} and the \texttt{Bool} type.
-However, the language does not strictly require the programmer to explicitly specify the type of a variable.
-
-\begin{figure}[ht!]
-\begin{align*}
-	id \in \texttt{Ident}\ 	&::=\ [a...zA...Z]+ \\
-	v \in \texttt{Var}\ 	&::=\ id \\
-	f \in \texttt{Fun}\ 	&::=\ id_1\ id_2\ \{\ s\ \}\\
-	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e) \\
-	\texttt{CustomExpr}\ 	&::=\ \texttt{String}\\
-	c \in \texttt{Const}\ 	&::=\ \texttt{True}\ |\ \texttt{False}\ |\ \mathbb{Z} \\
-	\texttt{Assign}\ 	&::=\ v\ :=\ e\ |\ v\ :=\ \texttt{CustomExpr}\ \\
-	s \in \texttt{Stmt}\ 	&::=\ \texttt{Assign}\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ \ell\ |\ s_1\ s_2\ \\
-	\ell \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
-	\circledplus\ 		&::=\ \texttt{+}\ |\ \texttt{-} \\
-	\circleddot\ 		&::=\ \texttt{!=}\ |\ \texttt{==}\ |\ \texttt{<} \ |\ \texttt{>} \ |\ \texttt{<=} \ |\ \texttt{>=} \\
-\end{align*}
-\caption{Syntax of the simple functional language}
-\label{fig:lang}
-\end{figure}
-
-While our language allows to define functions on its own, we will further allow the usage of function names that may have not been defined in the language itself.
-To allow this, the user can create a separate Python file that contains the definitions of additional functions.
-Our interpreter will then forward the calls to these functions to the Python interpreter.
-This allows us to easily access more complicated functionality such as the \texttt{print} function.
-
-Additionally to standard operations, the syntax of the language supports creating layers that can be applied onto each identifier. The syntax of a layer consists of the following:
-\begin{itemize}
-	\item $id_1$ -- The name of the identifier to be refined
-	\item $id_2$ -- The name of the layer
-	\item An arbitrary \textit{String} that represents the refinement.
-Note that the syntax does not restrict the format of this string, as this information will be parsed later by the corresponding type checking layer.
-\end{itemize}
-
-These layers can then be used to refine an identifier.
-In a very basic and minimal approach these layers can be used to annotate types to variables and functions.
-Additionally, one might add additional layers to a variable to ensure more properties.
-Each layer will be checked individually and independently.
-A more detailled approach of the architecture behind our layered approach will be described in section \ref{ssec:architecture}.
-
-An example of a program written in our language can be seen in \ref{lst:fac_example}.
-The program defines a factorial function that is refined in two layers.
-The first layer \textit{base} is a simple type layer that defines that \texttt{fac} takes an \texttt{Int} and returns an \texttt{Int}.
-The second layer \textit{liquid} is a refinement layer that refines that the input should be greater or equal to zero and that the output will be greater than zero.
-
-\begin{lstlisting}[caption={Example of a factorial function in our simple language}, label={lst:fac_example}]
-fac::base:: Int -> Int
-fac::liquid:: {v:Int | v >= 0} -> {v:Int | v > 0}
-fac x { 
-let cond := (x == 0) in
-	if cond then 
-		1 
-	else 
-		x * fac (x - 1) 
-}
-\end{lstlisting}
-
-\subsection{Language Implementation Architecture}
-\label{ssec:architecture}
-
-The implementation of our language will be done in the \texttt{Python} programming language.
-Roughly speaking, the implementation will consist of three parts, which we will describe throughout this section:
-
-\begin{enumerate}
-	\item A parser that parses the input program and creates an abstract syntax tree (AST) of the program.
-	\item A layered type checker that checks the defined type layers for the tree incrementally.
-	\item An interpreter that executes the program.
-\end{enumerate}
-
-To parse a input program, we will use the Python library \Lark{} \cite{Lark}.
-This library allows us to define a grammar for the language and then parse a program according to this grammar.
-All further processing will be performed on the resulting AST.
-To process and execute the AST we will use the Visitors and Transformers provided by the \Lark{} library.
-
-After initially parsing the AST, it will contain explicit nodes for each of the defined layers.
-As a first processing step we use a Transformer to collect these layers and remove the corresponding nodes from the AST.
-The implementation of each node has to be provided by the user as a separate Python module.
-Our language Interpreter will then dynamically load the corresponding module and use it to check the layer.
-To interact with our layered type checker correctly the module has to implement the following functions:
-
-\begin{itemize}
-	\item \texttt{depends\_on} -- This function returns a list of all layers that this layer depends on. This is used to ensure that the layers are checked in the correct order.
-	\item \texttt{parse\_type} -- This function may be used to define and parse custom data types that are defined usinf the $CustomExpr$ rule in the grammar.
-\todo{Implementation details -- When and how is this function called?}
-	\item \texttt{check\_layer} -- This function is used to check the layer.
-It is passed the AST of the program with all the annotated information from the previous layers.
-It should return a boolean value that indicates whether the layer is valid or not.
-\end{itemize}
-
-Using the \texttt{depends\_on} function, the layered type checker will first build a dependency graph of all layers.
-It will then topologically sort the layers and check them in the correct order.
-Once all layers have been checked successfully, the interpreter will execute the program.
-
-\subsection{Examples}
-\label{ssec:examples}
-
-\section{Evaluation}
-\label{sec:evaluation}
-\section{Conclusion}
-\label{sec:conclusion}
-\section{Acknowledgments}
-\label{sec:acknowledgments}
-
+\input{acknowledgements}
 \end{document}

### 2023-01-27 paper/use-cases.tex
* Separated chapters into separate tex files * Finished chapter about first use case * Rewrote parts of the introduction chapter to be more generally about   type systems and not liquid types in particular

\section{Example Use-Cases}
\label{ssec:use_cases}
For the scope of this paper, we have developed two exemplary use-cases for our framework that demonstrates our layered verification approach. In the first use-case, we defined a simple program that uses multiple different, partly independent type systems. For our second example, we use our framework to present a more flexible approach to using Liquid Types.

\subsection{Use Case 1 -- File Processing}
\label{sssec:use_case_1}

In our first use case we implemented a small program that processes the contents of a text file. Througout this section we will describe the program and the implemented layers required to realize this program. Lastly we will discuss the findings and benefits we gained from using our layered verification approach.

\subsubsection{Description}
In our first use-case we design a small program that reads contents from a file and performs certain operations based on that. The base code for our example can be found in Listing \ref{lst:uc1_code}. In our example, the file to be read will contain one number per each line. Depending on whether the number is positive or negative, we add the number as is or the square of the number to the resulting value.

\begin{lstlisting}
processLine lines idx length {
    end := idx == length

    if end then {
        0
    } else {

        curLine := getLine(lines, idx)

        number := strToInt(curLine)

        greaterThanZero := number > 0

        if greaterThanZero then {
            number + processLine(lines, idx + 1, length)
        } else {
            (number * number) + processLine(lines, idx + 1, length)
        }

    }
}

processLines lines {
    length := len(lines)

    idx := 0
    processLine(lines, idx, length)
}

fileName := "numbers.data"
fileHandle := openFile(fileName)
lines := readLines(fileHandle)
closeFile(fileHandle)
total := processLines(lines)

print(total)
total

\end{lstlisting}

All functions that have not been defined directly \ref{lst:uc1_code} are provided in a separate, external python file. As our language on itself does not include a type system and there are no layer annotations in the code above, no type checking will be performed on the code as it is presented.

However even this simple example could already introduce several errors that could be prevented by introducing some static analysis. First of all, it does not provide any information about the variable and function type and thus, we cannot ensure that functions are called with arguments of the correct types which might lead to runtime errors.

Additionally we our code interacts with the file it operates on through a file descriptor. TThis file descriptor can be viewed as a resource that might have an implicit or explicit internal state. In a very simple approach this file descriptor could be in the states "Open" or "Closed". In order to read from it, it should be in an "Open" state and additionally we should ensure that by the end of our program the resource has been properly closed again.

In order to improve our example code and allow static analysis, we therefore will introduce three layers:
\begin{enumerate}
	\item A \textit{Type} Layer -- This layer will be used to simply define variable and function types
	\item The \textit{Typechecking} Layer -- This layer will allow to flexibly define a type system and then use the information provided by the \textit{Type} layer to check if the program is well-typed. It therefore depends on the \textit{Type} layer to provide the type information.
	\item A \textit{State} Layer -- This layer will be used to define states on variables. Additionally it can model state transitions. We will use it to track the state of the file descriptor. This layer operates completely independent from the \textit{Type} and \textit{Typechecking} layer.
\end{enumerate}

\subsubsection{Implemented Layers}
In the following sections we will explain in more detail how the layers are implemented and what Syntax they use.

\paragraph{Types Layer}
The \textit{Types} layer is the first layer that is checked. It is used to define the types of variables and functions. To define the type of a variable we use the \texttt{Layer} syntax of our language. A layer definition for the \textit{Types} layer looks as follows:

\begin{lstlisting}
identifier :: types :: Type
\end{lstlisting}

The \texttt{identifier} is the name of the variable or function for which we want to define the type. The \texttt{types} keyword is used to associate this information with the correct layer. The \texttt{Type} is the desired type information to be annotated. There are no restrictions on how a \texttt{Type} should look, as the types layer simply annotates the given string into the current node and does not further process it. For example to define the type of a variable \texttt{foo} as \texttt{int} we would write:

\begin{lstlisting}
foo :: types :: int
\end{lstlisting}

In order to define the type of a function, there is a slight adaption to this syntax. While the first two parts are identical to the variable definition, we now need to provide type information for each argument of the function. This is done by using the \texttt{Type} syntax for each argument and then separating them with an arrow symbol \texttt{->}. The return type of the function is then defined by the last \texttt{Type}. If a function does not take an argument the defintion simply starts with the arrow symbol. The following code piece contains an example for a function \texttt{f} that takes two arguments of type \texttt{int} and returns a value of type \texttt{bool}, and a function \texttt{g} that does not take any arguments and returns a value of type \texttt{string}:

\begin{lstlisting}
f :: types :: int -> int -> bool
g :: types :: -> string
\end{lstlisting}

It is important to note that the \textit{Types} layer does not perform any type checking. It simply annotates the given type information into the AST. The \textit{Types} layer only checks for the following inconsistencies:

\begin{itemize}
	\item The type for a variable can only be defined once in the current scope.
	\item When encountering a function definition, the type for it needs to be defined before the function signature. Additionally the number of arguments in the function signature needs to match the number of arguments in the type definition.
\end{itemize}

Any further type checking is performed by the \textit{Typechecking} layer.

\paragraph{Typechecking Layer}

The \textit{Typechecking} layer is used to define a type system and then use the information provided by the \textit{Types} layer to check if the program is well-typed. It therefore depends on the \textit{Types} layer to provide the type information. In our implementation of the type checker, we do not impose any restrictions on the specific type system. We rather provide a generic type checking implementation that allows the user to flexibly define their own types and subtype relationships. However, since our language implicitly natively requires specific expressions to be of a number or boolean type, there are implicitly part of our type system.

To define the specific type system, we use the \texttt{Layer} syntax of our language. However, instead of defining layers for identifiers as in the \textit{Types} layer, we reserve a selected set of keywords to define the type system. Depending on the definition provided by the user, the type checker will internally build a type graph that is used to perform the type checking.

The following keywords are reserved and can be used to define parts of the type system:

\begin{itemize}
	\item \texttt{logicalTypes} -- This keyword is used to define a set of logical types. Multiple types can be defined by separating them with a space.
	\item \texttt{numTypes} -- This keyword is used to define a set of numeric types. Multiple types can be defined by separating them with a space.
	\item \texttt{subtype} -- This keyword is used to define a subtype relationship between two or more types. The syntax for a subtype relationship is \texttt{<type1> <: <type2> <: ... <: <typeN>}. The subtype relationship is then defined by the order of the types. The first type is the supertype and the following type is a subtype of the previous type.
\end{itemize}

The following code piece is a short demonstration of a type system containing different logical types, numeric types and subtype relationships:

\begin{lstlisting}
logicalTypes :: typecheck :: bool
numTypes :: typecheck :: int float
subtype :: typecheck :: float <: int
\end{lstlisting}

The type system defined by the user will be used to perform the typechecking. To perform this typecheck the type checker accesses the annotations provided by the \textit{Types} layer. For the case that a type is neither a numeric or logical type and it does not have any subtype relationships, it is not required to explicitly define it in the type system. The type checker will then simply assume the existance of such a type that is not a subtype of any other type nor does it have any subtypes.

Our type checker will perform the following checks:

\begin{itemize}
	\item The type of each variable needs to be defined.
	\item When assigning a value to a variable, the type of the value needs to be a subtype of the type of the variable.
	\item When performing a logical operations (e.g. \texttt{\&\&}, \texttt{||}) the type of the operands needs to be a logical type. The result of the operation will be a logical type.
	\item When performing a numeric operation (e.g. \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}) the type of the operands needs to be a numeric type. The result of the operation will be the most general numeric type of the operands.
	\item When performing a comparison operation (e.g. \texttt{==}, \texttt{!=}, \texttt{<}, \texttt{>}, \texttt{<=}, \texttt{>=}) the type of the operands needs to be compatible. This means that either the right hand side is a subtype of the left hand side or the left hand side is a subtype of the right hand side. The result of such an operation is always a boolean value.
	\item In an \texttt{if} statement the condition needs to be a logical type.
	\item When performing a function call, the number of arguments needs to match the number of arguments in the function definition. Additionally the type of each argument needs to be a subtype of the type of the corresponding argument in the function definition.
\end{itemize}

Since we require the type of each variable to be defined, this means we require full information about the types of all variables. We do not perform any type inference. This is a deliberate design decision, as we want to keep the language as simple as possible. However, by the design of our language, it would be possible to develop an additional layer that runs after the \textit{Types} layer but before the \textit{Typechecking} layer that performs a type inference based on the incomplete information provided.

\paragraph{State Layer}

The \textit{State} layer is used to define states on variables. Similarly, we can use it to check if a variable has a certain state and define state transitions. This can for example be used to track the state of resources that are used by a program. In our implementation, the state of a variable is completely independent of its type, therefore it does not require any dependencies on the \textit{Types} layer.

The state of a variable is defined by a set of strings. Generally there are two ways to check the state of a variable or transition it's state to a new one. Firstly, we can explicitly do this by using a layer statement of the language. Secondly, we can also impose state requirements on arguments of function calls. Then when passing a variable to a function, the layer will check if the variable has the required state and apply the state transition if there are any.

Similar to the \textit{Types} layer, the \textit{State} layer uses the \texttt{Layer} syntax of the language to define the state of a variable. The first part of the layer statement defines the variable or function name to define a state on, while the second part is simply the layer identifier \texttt{state}. The last part of the layer definition defines the state.

A state definition is enclosed by curly brackets and can contain the following parts:

\begin{itemize}
	\item A state \textit{requirement} -- Imposes a restrition on the current state the variable needs to have. This is simply a string that needs to be part of the current state of the variable.
	\item A state \textit{transition} -- Defines a transition to a new state. Generally a state transition has the form \texttt{<oldState> => <newState>}. However, either the old or new state may be omitted. When omitting the old state, this simply means that something is added to the current state of the variable unconditionally. When omitting the new state, this means that the corresponding state string will be removed from the set of states.
\end{itemize}

For our use-case, we want to track the state of the file descriptor. Specifically, we want to make sure, that the file handle is open before we can read from it. Additionally, we do not want to read from a file handle again once it has already been read. Therefore, we define the states \texttt{Open}, \texttt{Consumed} and \texttt{Closed} and define the appropriate state transistions as seen in Listing \ref{lst:uc1_state}. Additionally, at the end of the program, we want to make sure that the file handle is closed. Therefore, we define a state requirement that the file handle needs to be in the state \texttt{Closed}.

\begin{lstlisting}
createFD :: state :: {} -> { Closed }
openFile :: state :: {Closed => Open} -> {}
readLines :: state :: {Open => Consumed} -> {}
closeFile :: state :: {Consumed => Closed} -> {}

-- State requirement at the end of the program
fileDescriptor :: state :: {Closed}
\end{lstlisting}

With these three layers, we have defined the complete type system for this use-case. Using these three layers we can now ensure the following properties of our program statically:

\begin{itemize}
	\item Our program is type safe. This means that we can be sure that the program will not crash due to type errors.
	\item Using the state layer, we ensure that the file descriptor is handled correctly and is in a closed state at the end of the program.
\end{itemize}

\subsubsection{Discussion}

Implementing this first use-case has given us valuable results and insights on our approach. We have successfully implemented three layers independently from each other, however since the \textit{Typecheck} layer depends on the information provided by the \textit{Types} layer, we successfully demonstrated, that information can be communicated between layers. We have also successfully demonstrated that the layers can be used to define a type system for a program. This means that we can use the layers to define a type system for a program and then use the type checker to check if the program is type safe.

Furthermore, as a result of this use-case, we provide an implementation of a state layer. This layer can be used to define states on variables and functions and to check if a variable has a certain state. Additionally, we can use the state layer to define state transitions. This can be used to track the state of resources that are used by a program. We believe that this layer is suitable for a wide range of applications far beyond the simple demonstration we have shown in this use-case.

One shortcoming of our implementation is that we require that the \textit{Types} layer needs to provide full information on the types of all variables and functions. For a developer it would be much more flexible if it would be possible to only provide partial information and use type-inference to deduce the missing information. However, by the design of our framework extending the example to also support type inference is possible. We could for example add a new layer that runs after the \textit{Types} layer but before the \textit{Typecheck} layer. This layer could then perform type inference and annotate the missing information to the AST. This way the \textit{Typecheck} layer would have all the information it needs to perform the type checking. However for the scope of this paper, we have decided not to implement this feature.

\subsection{Use Case 2 -- Liquid Types}
\label{sssec:use_case_2}

### 2023-01-30 paper/use-cases.tex
* Removed unused sample tex files * Slight adaptions to first use case file

@@
 
-With these three layers, we have defined the complete type system for this use-case. Using these three layers we can now ensure the following properties of our program statically:
+The final program with all layer annotations can be found in Listing \ref{lst:uc1_annotated}.
+With these three layers, we have defined the complete type system for this use-case.
+Using these three layers we can now ensure the following properties of our program statically:
 
@@
 \end{itemize}
+
+\begin{lstlisting}
+-- Type Layer Definitions for externally defined functions
+getLine :: types :: [string] -> int -> string
+strToInt :: types :: string -> int
+len :: types :: [string] -> int
+createFD :: types :: string -> FileHandle
+openFile :: types :: FileHandle -> void
+readLines :: types :: FileHandle -> [string]
+closeFile :: types :: FileHandle -> void
+print :: types :: Any -> void
+
+-- Layer definitions for type checking
+logicalTypes :: typecheck :: bool
+numTypes :: typecheck :: int
+subtype :: typecheck :: int <: Any
+subtype :: typecheck :: string <: Any
+
+-- State layer definitions
+createFD :: state :: {} -> { Closed }
+openFile :: state :: {Closed => Open} -> {}
+readLines :: state :: {Open => Consumed} -> {}
+closeFile :: state :: {Consumed => Closed} -> {}
+
+processLine :: types :: [string] -> int -> int -> int
+processLine lines idx length {
+
+    end :: types :: bool
+    end := idx == length
+
+    if end then {
+        0
+    } else {
+
+        curLine :: types :: string
+        curLine := getLine(lines, idx)
+
+        number :: types :: int
+        number := strToInt(curLine)
+
+        greaterThanZero :: types :: bool
+        greaterThanZero := number > 0
+
+        if greaterThanZero then {
+            number + processLine(lines, idx + 1, length)
+        } else {
+            (number * number) + processLine(lines, idx + 1, length)
+        }
+
+    }
+}
+
+processLines :: types :: [string] -> int
+processLines lines {
+    length :: types :: int
+    length := len(lines)
+
+    idx :: types :: int
+    idx := 0
+    processLine(lines, idx, length)
+}
+
+fileName :: types :: string
+fileName := "numbers.data"
+
+fileDescriptor :: types :: FileHandle
+fileDescriptor := createFD(fileName)
+
+openFile(fileDescriptor)
+
+lines :: types :: [string]
+lines := readLines(fileDescriptor)
+
+closeFile(fileDescriptor)
+
+total :: types :: int
+total := processLines(lines)
+
+print(total)
+total
+
+-- State post-conditions
+-- The file handle is closed after the program has finished
+fileDescriptor :: state :: {Closed}
+\end{lstlisting}
 

### 2023-01-30 paper/implementation.tex
* Added outline description for use case 2

@@
 To demonstrate our layered approach we implemented a simple functional language.
-The syntax of said language is described in \ref{fig:lang}.
-The language is similar to most simple functional languages and generally follows the reduction semantics of a standard Hindley-Milner type system.
-The only data types available in the language are the \texttt{Int} and the \texttt{Bool} type.
-However, the language does not strictly require the programmer to explicitly specify the type of a variable.
+The syntax of said language is described in Fig.
+\ref{fig:lang}.
+The language is similar to most simple functional languages.
+Implicitly, our language natively supports \textt{Boolean} and \texttt{Integer} datatypes through the available constants.
+However, the language on it's own does not require any type annotations.
 
@@
 
-These layers can then be used to define properties.
-Depending on the Layers' implementation, these properties can apply to identifiers or can be used to define properties of the type system the layer defines.
+These layers can be used to define properties.
+Depending on the specific implementation of a layer, these properties can apply to identifiers or can be used to define properties of the type system the layer defines.
 In a very basic and minimal approach these layers can be used to annotate types to variables and functions.

### 2023-01-31 code/LayeredTypes/examples/python_impls/use_case_2.py
* Added bubble sort implementation for use case 2

@@ -1,0 +1,11 @@
+from builtins import len, print
+
+def set_element(ls, idx, val):
+    ls[idx] = val
+
+def get_element(ls, idx):
+    return ls[idx]
+
+def get_list():
+    # Return some unordered list with 12 elements
+    return [12, 11, 10, 9, 8, 7, 6, 5, 4, 3, 2, 1]

### 2023-03-06 code/LayeredTypes/tests/layer_implementations/B.py
* Adapted pytest configuration to fix Github pipeline

@@ -1,4 +1,4 @@
-from tests.utils import call_order
+from utils import call_order
 
 def depends_on():
     return {"A"}

### 2023-03-06 code/LayeredTypes/tests/layer_implementations/D.py

@@ -1,4 +1,4 @@
-from tests.utils import call_order
+from utils import call_order
 
 def depends_on():
     return { "B", "C"}

### 2023-03-06 code/LayeredTypes/tests/test_layers.py

@@ -1,7 +1,7 @@
 import unittest
 from graphlib import CycleError
 
-from tests.utils import parse_file, get_compiler, full_path, call_order
+from utils import get_compiler, full_path, call_order
 
 class TestLayerDefinitions(unittest.TestCase):
 

### 2023-03-06 code/LayeredTypes/tests/test_liquid.py

@@ -1,8 +1,8 @@
 import unittest
 
-from aeon.frontend.parser import mk_parser, parse_type
+from aeon.frontend.parser import parse_type
 from compiler.Exceptions import LayerException
-from tests.utils import parse_file, typecheck_correct_file, get_compiler, full_path
+from utils import typecheck_correct_file, get_compiler, full_path
 
 
 class TestLiquidLayer(unittest.TestCase):
@@ -13,7 +13,7 @@ class TestLiquidLayer(unittest.TestCase):
         typecheck_correct_file(self, "/test_code/liquid/simple_assignment_infer.fl")
 
     def test_simple_assignment_fail(self):
-        compiler = get_compiler(layer_path="../layer_implementations")
+        compiler = get_compiler(layer_path="layer_implementations")
         src_file = full_path("/test_code/liquid/simple_assignment_fail.fl")
 
         with self.assertRaises(LayerException) as context:
@@ -31,7 +31,7 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual(type_expected, e.type_expected)
 
     def test_simple_assignment_infer_fail(self):
-        compiler = get_compiler(layer_path="../layer_implementations")
+        compiler = get_compiler(layer_path="layer_implementations")
         src_file = full_path("/test_code/liquid/simple_assignment_infer_fail.fl")
 
         with self.assertRaises(LayerException) as context:
@@ -73,7 +73,7 @@ class TestLiquidLayer(unittest.TestCase):
         typecheck_correct_file(self, "/test_code/liquid/let_inner_layer.fl")
 
     def test_let_inner_layer_fail(self):
-        compiler = get_compiler(layer_path="../layer_implementations")
+        compiler = get_compiler(layer_path="layer_implementations")
         src_file = full_path("/test_code/liquid/let_inner_layer_fail.fl")
 
         with self.assertRaises(LayerException) as context:
@@ -87,7 +87,7 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual("y", e.identifier)
 
     def test_ident_type_undefined(self):
-        compiler = get_compiler(layer_path="../layer_implementations")
+        compiler = get_compiler(layer_path="layer_implementations")
         src_file = full_path("/test_code/liquid/ident_type_undefined.fl")
 
         with self.assertRaises(LayerException) as context:
@@ -101,7 +101,7 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual("x", e.identifier)
 
[…]

### 2023-03-06 code/LayeredTypes/tests/test_state_layer.py

@@ -1,8 +1,8 @@
 import unittest
 
 from compiler.Exceptions import LayerException
-from layer_implementations.state import StateError as StateError, ArgumentState
-from tests.utils import typecheck_correct_file, get_compiler, full_path
+from layer_implementations.state import ArgumentState
+from utils import typecheck_correct_file, get_compiler, full_path
 
 
 class StateLayer(unittest.TestCase):

### 2023-03-06 code/LayeredTypes/tests/test_liquid.py
* Added tests for function definitions

@@ -128,8 +128,11 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual(1, e.offset)
 
     def test_fun_def(self):
-        #TODO
-        self.assertTrue(False)
+        typecheck_correct_file(self, "/test_code/liquid/fun_def.fl")
+
+
+    def test_fun_def_multiple_args(self):
+        typecheck_correct_file(self, "/test_code/liquid/fun_def_multiple_args.fl")
 
     def test_nested_fun_call_multiple_args(self):
         typecheck_correct_file(self, "/test_code/liquid/nested_fun_call_multiple_args.fl")

### 2023-03-06 code/LayeredTypes/tests/test_liquid.py
* Added tests for function definitions

@@ -130,9 +130,32 @@ class TestLiquidLayer(unittest.TestCase):
     def test_fun_def(self):
         typecheck_correct_file(self, "/test_code/liquid/fun_def.fl")
 
+    def test_fun_def_fail_inner(self):
+        compiler = get_compiler(layer_path="layer_implementations")
+        src_file = full_path("/test_code/liquid/fun_def_fail_inner.fl")
+
+        with self.assertRaises(LayerException) as context:
+            compiler.typecheck(src_file)
+
+        #TODO
+        self.assertTrue(False)
+
+        self.assertEqual("liquid", context.exception.layer_name)
+        e = context.exception.original_exception
+        self.assertEqual("LiquidSubtypeException", e.__class__.__name__)
+        self.assertEqual(5, e.lineno)
+        self.assertEqual(1, e.offset)
+
+    def test_fun_def_fail_oncall(self):
+        # TODO
+        self.assertTrue(False)
 
     def test_fun_def_multiple_args(self):
         typecheck_correct_file(self, "/test_code/liquid/fun_def_multiple_args.fl")
+
+    def test_fun_def_multiple_args_fail(self):
+        # TODO
+        self.assertTrue(False)
 
     def test_nested_fun_call_multiple_args(self):
         typecheck_correct_file(self, "/test_code/liquid/nested_fun_call_multiple_args.fl")

### 2023-03-06 code/LayeredTypes/tests/test_liquid.py
* Finished tests for function definition that fails on inner assign

@@ -137,14 +137,15 @@ class TestLiquidLayer(unittest.TestCase):
         with self.assertRaises(LayerException) as context:
             compiler.typecheck(src_file)
 
-        #TODO
-        self.assertTrue(False)
-
         self.assertEqual("liquid", context.exception.layer_name)
         e = context.exception.original_exception
         self.assertEqual("LiquidSubtypeException", e.__class__.__name__)
-        self.assertEqual(5, e.lineno)
-        self.assertEqual(1, e.offset)
+        self.assertEqual(6, e.lineno)
+        self.assertEqual(5, e.offset)
+        type_expected = parse_type("{v:Int | v > 0}")
+        type_actual = parse_type("{v:Int | v >= 0}")
+        self.assertEqual(type_actual, e.type_actual)
+        self.assertEqual(type_expected, e.type_expected)
 
     def test_fun_def_fail_oncall(self):
         # TODO

### 2023-03-06 code/LayeredTypes/tests/test_liquid.py
* Finished tests for function definition that fails on call of user-defined functions

@@ -148,8 +148,21 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual(type_expected, e.type_expected)
 
     def test_fun_def_fail_oncall(self):
-        # TODO
-        self.assertTrue(False)
+        compiler = get_compiler(layer_path="layer_implementations")
+        src_file = full_path("/test_code/liquid/fun_def_fail_oncall.fl")
+
+        with self.assertRaises(LayerException) as context:
+            compiler.typecheck(src_file)
+
+        self.assertEqual("liquid", context.exception.layer_name)
+        e = context.exception.original_exception
+        self.assertEqual("LiquidSubtypeException", e.__class__.__name__)
+        self.assertEqual(10, e.lineno)
+        self.assertEqual(1, e.offset)
+        type_expected = parse_type("{v:Int | v > 0}")
+        type_actual = parse_type("{v:Int | v == 0}")
+        self.assertEqual(type_actual, e.type_actual)
+        self.assertEqual(type_expected, e.type_expected)
 
     def test_fun_def_multiple_args(self):
         typecheck_correct_file(self, "/test_code/liquid/fun_def_multiple_args.fl")

### 2023-03-06 code/LayeredTypes/tests/test_liquid.py
* Added tests that check correct definition of refinements

@@ -168,8 +168,56 @@ class TestLiquidLayer(unittest.TestCase):
         typecheck_correct_file(self, "/test_code/liquid/fun_def_multiple_args.fl")
 
     def test_fun_def_multiple_args_fail(self):
-        # TODO
-        self.assertTrue(False)
+        compiler = get_compiler(layer_path="layer_implementations")
+        src_file = full_path("/test_code/liquid/fun_def_multiple_args_fail.fl")
+
+        with self.assertRaises(LayerException) as context:
+            compiler.typecheck(src_file)
+
+        self.assertEqual("liquid", context.exception.layer_name)
+        e = context.exception.original_exception
+        self.assertEqual("LiquidSubtypeException", e.__class__.__name__)
+        self.assertEqual(8, e.lineno)
+        self.assertEqual(5, e.offset)
+        type_expected = parse_type("{v:Int | v > 0}")
+        type_actual = parse_type("{v:Int | v >= 0}")
+        self.assertEqual(type_actual, e.type_actual)
+        self.assertEqual(type_expected, e.type_expected)
+
+    def test_fun_def_multiple_args_fail_oncall(self):
+        compiler = get_compiler(layer_path="../layer_implementations")
+        src_file = full_path("/test_code/liquid/fun_def_multiple_args_fail_oncall.fl")
+
+        with self.assertRaises(LayerException) as context:
+            compiler.typecheck(src_file)
+
+        self.assertEqual("liquid", context.exception.layer_name)
+        e = context.exception.original_exception
+        self.assertEqual("LiquidSubtypeException", e.__class__.__name__)
+        self.assertEqual(13, e.lineno)
+        self.assertEqual(1, e.offset)
+        type_expected = parse_type("{v:Int | x > v}")
+        type_actual = parse_type("{v:Int | v == 1}")
+        self.assertEqual(type_actual, e.type_actual)
+        self.assertEqual(type_expected, e.type_expected)
+
 
     def test_nested_fun_call_multiple_args(self):
         typecheck_correct_file(self, "/test_code/liquid/nested_fun_call_multiple_args.fl")
+
+    def test_liquid_fun_refinement_noref(self):
+        typecheck_correct_file(self, "/test_code/liquid/liquid_fun_refinement_noref.fl")
+
+    def test_liquid_fun_refinement_no_overlap(self):
+        typecheck_correct_file(self, "/test_code/liquid/liquid_fun_refinement_no_overlap.fl")
+
[…]

### 2023-03-07 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Added checking for duplicate argument name references in function definitions

@@ -1,8 +1,9 @@
+from collections import Counter
 from copy import copy
 
 import lark
 
-from aeon.core.liquid import LiquidLiteralInt, LiquidLiteralBool, LiquidVar
+from aeon.core.liquid import LiquidLiteralInt, LiquidLiteralBool, LiquidVar, liquid_free_vars
 from aeon.core.substitutions import substitution_in_type
 from aeon.core.terms import Var
 from aeon.core.types import t_int, t_bool, t_string, RefinedType
@@ -37,6 +38,21 @@ class LiquidTypeUndefinedError(TypecheckException):
     def __init__(self, identifier, line, column):
         self.identifier = identifier
         super().__init__(f"Type of identifier {identifier} is undefined", line, column)
+        pass
+
+class LiquidFunctionDefinitionError(TypecheckException):
+    """Exception that will be raised when during liquid typechecking a function definition is invalid.
+
+    Attributes:
+        identifier -- the identifier that is undefined
+    """
+    def __init__(self, fun_name, argument_type, argument_idx, duplicate_var, line, column):
+        self.argument_type = argument_type
+        self.argument_idx = argument_idx
+        self.duplicate_var = duplicate_var
+        self.fun_name = fun_name
+
+        super().__init__(f"Function definition for {fun_name}:\rInvalid refinement at index {argument_idx}: {argument_idx}. (References non-unique variable {duplicate_var})", line, column)
         pass
 
 def make_name_unique(name :str, context :TypingContext) -> str:
@@ -124,6 +140,18 @@ class LiquidLayer(lark.visitors.Interpreter):
             raise TypecheckException("Could not parse refinement type: " + str(e), tree.meta.line, tree.meta.column)
 
         if is_fn:
+            # We check that the refinements are valid
+            # Create a map from name to count of x.name in refinement types:
+
+            refinement_var_counts = Counter([x.name for x in refinement_types])
+            for i, ref in enumerate(refinement_types):
+                for var in liquid_free_vars(ref.refinement):
+                    if var == ref.name:
+                        # We ignore the variable that is bound by the refinement
+                        continue
+                    if refinement_var_counts[var] > 1:
+                        raise LiquidFunctionDefinitionError(identifier, ref, i, var, tree.meta.line, tree.meta.column)
+
             self.__fun_types[identifier] = refinement_types
         else:
             self.__types[identifier] = refinement_types[0]
[…]

### 2023-03-07 code/LayeredTypes/tests/test_liquid.py

@@ -211,13 +211,22 @@ class TestLiquidLayer(unittest.TestCase):
     def test_liquid_fun_refinement_no_overlap(self):
         typecheck_correct_file(self, "/test_code/liquid/liquid_fun_refinement_no_overlap.fl")
 
-    def test_liquid_fun_ref_fail(self):
+    def test_liquid_fun_refinement_ref_fail(self):
         compiler = get_compiler(layer_path="layer_implementations")
-        src_file = full_path("/test_code/liquid/liquid_fun_ref_fail.fl")
+        src_file = full_path("/test_code/liquid/liquid_fun_refinement_ref_fail.fl")
 
         with self.assertRaises(LayerException) as context:
             compiler.typecheck(src_file)
 
         self.assertEqual("liquid", context.exception.layer_name)
         e = context.exception.original_exception
-        self.assertTrue(False)
+        self.assertEqual("LiquidFunctionDefinitionError", e.__class__.__name__)
+        self.assertEqual(5, e.lineno)
+        self.assertEqual(1, e.offset)
+        self.assertEqual("fun", e.fun_name)
+        self.assertEqual("v", e.duplicate_var)
+        self.assertEqual(2, e.argument_idx)
+        type_expected = parse_type("{x:Int | x > v}")
+        self.assertEqual(type_expected, e.argument_type)
+
+        

### 2023-03-07 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Added test skeletons for substitution

@@ -1,5 +1,6 @@
 from collections import Counter
 from copy import copy
+import typing as tp
 
 import lark
 
@@ -72,6 +73,26 @@ def make_name_unique(name :str, context :TypingContext) -> str:
         name = f"{name}_{p}"
 
     return name
+
+
+def substitute_refinement_names(fun_types: tp.List[RefinedType]):
+    ref_var_names = [var.name for var in fun_types[:-1]]
+    ref_var_names_cnt = Counter(ref_var_names)
+    # We rename the variables in the refinements of the arguments
+    for idx, name in enumerate(ref_var_names):
+        fun_types[idx] = substitution_in_type(fun_types[idx], Var(f"$arg{idx}"), name)
+        if ref_var_names_cnt[name] == 1:
+            for i in range(idx + 1, len(ref_var_names)):
+                fun_types[i] = substitution_in_type(fun_types[i], Var(f"$arg{idx}"), name)
+
+    return fun_types
+
+def substitute_argument_names(arg_names: tp.List[str], arg_types: tp.List[RefinedType]):
+    for i in range(len(arg_names)):
+        ref_name = arg_types[i].name
+        for j in range(i + 1, len(arg_names)):
+            arg_types[j] = substitution_in_type(arg_types[j], arg_names[i], ref_name)
+
 class LiquidLayer(lark.visitors.Interpreter):
     def __init__(self):
         self.__types = {}
@@ -140,8 +161,6 @@ class LiquidLayer(lark.visitors.Interpreter):
             raise TypecheckException("Could not parse refinement type: " + str(e), tree.meta.line, tree.meta.column)
 
         if is_fn:
-            # We check that the refinements are valid
-            # Create a map from name to count of x.name in refinement types:
 
             refinement_var_counts = Counter([x.name for x in refinement_types])
             for i, ref in enumerate(refinement_types):
@@ -224,6 +243,8 @@ class LiquidLayer(lark.visitors.Interpreter):
             free_var_name = make_name_unique(free_var_name, context)
 
             # We need to rename the variable in all the refinements for the remaining arguments
+            # TODO: Differentiate cases where different arguments use the same variable name
+            # This is possible as long as the variable is not referenced in another refinement
             for j in range(i+1, len(expected_arg_types)):
                 expected_arg_types[j] = substitution_in_type(expected_arg_types[j], Var(free_var_name), original_name)
 
@@ -260,7 +281,8 @@ class LiquidLayer(lark.visitors.Interpreter):
[…]

### 2023-03-07 code/LayeredTypes/tests/test_liquid.py

@@ -1,7 +1,10 @@
 import unittest
 
+from aeon.core.liquid import LiquidApp, LiquidVar, LiquidLiteralInt
+from aeon.core.types import RefinedType, t_int
 from aeon.frontend.parser import parse_type
 from compiler.Exceptions import LayerException
+from compiler.interpreters.LiquidTypeChecker import substitute_argument_names, substitute_refinement_names
 from utils import typecheck_correct_file, get_compiler, full_path
 
 
@@ -185,8 +188,10 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual(type_expected, e.type_expected)
 
     def test_fun_def_multiple_args_fail_oncall(self):
-        compiler = get_compiler(layer_path="../layer_implementations")
+        compiler = get_compiler(layer_path="layer_implementations")
         src_file = full_path("/test_code/liquid/fun_def_multiple_args_fail_oncall.fl")
+
+        compiler.typecheck(src_file)
 
         with self.assertRaises(LayerException) as context:
             compiler.typecheck(src_file)
@@ -229,4 +234,35 @@ class TestLiquidLayer(unittest.TestCase):
         type_expected = parse_type("{x:Int | x > v}")
         self.assertEqual(type_expected, e.argument_type)
 
-        
+class TestSubstitutions(unittest.TestCase):
+
+    def test_single_refinement_substituted(self):
+        type_before = parse_type("{v:Int | v > 0}")
+        type_after = RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] ))
+
+        types_substituted = substitute_refinement_names([type_before])
+        self.assertEqual([type_after], types_substituted)
+
+    def test_two_refinements_references(self):
+        self.assertTrue(False)
+
+    def test_multiple_refinements_references(self):
+        self.assertTrue(False)
+
+    def test_multiple_refinements_no_references(self):
+        self.assertTrue(False)
+
+    def test_multiple_refinements_mixed(self):
+        self.assertTrue(False)
+
+    def test_substitute_arg_names_single_refinement(self):
+        self.assertTrue(False)
+
+    def test_substitute_arg_names_multiple_refinements_no_ref(self):
+        self.assertTrue(False)
+
+    def test_substitute_arg_names_multiple_refinements_mixed(self):
+        self.assertTrue(False)
+
+    def test_substitute_arg_names_multiple_refinements_references(self):
+        self.assertTrue(False)

### 2023-03-07 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Fixed implementation of refinement replacement used in fun_def

@@ -5,7 +5,7 @@ import typing as tp
 import lark
 
 from aeon.core.liquid import LiquidLiteralInt, LiquidLiteralBool, LiquidVar, liquid_free_vars
-from aeon.core.substitutions import substitution_in_type
+from aeon.core.substitutions import substitution_in_type, substitution_in_liquid
 from aeon.core.terms import Var
 from aeon.core.types import t_int, t_bool, t_string, RefinedType
 from aeon.frontend.parser import mk_parser
@@ -80,12 +80,11 @@ def substitute_refinement_names(fun_types: tp.List[RefinedType]):
     ref_var_names_cnt = Counter(ref_var_names)
     # We rename the variables in the refinements of the arguments
     for idx, name in enumerate(ref_var_names):
-        fun_types[idx] = substitution_in_type(fun_types[idx], Var(f"$arg{idx}"), name)
+        fun_types[idx].name = f"$arg{idx}"
+        fun_types[idx].refinement = substitution_in_liquid(fun_types[idx].refinement, LiquidVar(f"$arg{idx}"), name)
         if ref_var_names_cnt[name] == 1:
             for i in range(idx + 1, len(ref_var_names)):
-                fun_types[i] = substitution_in_type(fun_types[i], Var(f"$arg{idx}"), name)
-
-    return fun_types
+                fun_types[i].refinement = substitution_in_liquid(fun_types[i].refinement, LiquidVar(f"$arg{idx}"), name)
 
 def substitute_argument_names(arg_names: tp.List[str], arg_types: tp.List[RefinedType]):
     for i in range(len(arg_names)):

### 2023-03-07 code/LayeredTypes/tests/test_liquid.py

@@ -240,8 +240,8 @@ class TestSubstitutions(unittest.TestCase):
         type_before = parse_type("{v:Int | v > 0}")
         type_after = RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] ))
 
-        types_substituted = substitute_refinement_names([type_before])
-        self.assertEqual([type_after], types_substituted)
+        substitute_refinement_names([type_before, None])
+        self.assertEqual(type_after, type_before)
 
     def test_two_refinements_references(self):
         self.assertTrue(False)

### 2023-03-07 code/LayeredTypes/tests/test_liquid.py
* Parametrized test for refinement replacement

@@ -1,5 +1,7 @@
 import unittest
 
+import pytest
+
 from aeon.core.liquid import LiquidApp, LiquidVar, LiquidLiteralInt
 from aeon.core.types import RefinedType, t_int
 from aeon.frontend.parser import parse_type
@@ -235,7 +237,6 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual(type_expected, e.argument_type)
 
 class TestSubstitutions(unittest.TestCase):
-
     def test_single_refinement_substituted(self):
         type_before = parse_type("{v:Int | v > 0}")
         type_after = RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] ))
@@ -243,17 +244,6 @@ class TestSubstitutions(unittest.TestCase):
         substitute_refinement_names([type_before, None])
         self.assertEqual(type_after, type_before)
 
-    def test_two_refinements_references(self):
-        self.assertTrue(False)
-
-    def test_multiple_refinements_references(self):
-        self.assertTrue(False)
-
-    def test_multiple_refinements_no_references(self):
-        self.assertTrue(False)
-
-    def test_multiple_refinements_mixed(self):
-        self.assertTrue(False)
 
     def test_substitute_arg_names_single_refinement(self):
         self.assertTrue(False)
@@ -266,3 +256,65 @@ class TestSubstitutions(unittest.TestCase):
 
     def test_substitute_arg_names_multiple_refinements_references(self):
         self.assertTrue(False)
+
+test_vals = [
+    ([parse_type("{v:Int | v > 0}")],
+     [RefinedType("$arg0", t_int,
+                  LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] ))
+     ]
+    ),
+    (
+        [parse_type("{x:Int | x > 0}"), parse_type("{y:Int | y > x}")],
+        [
+            RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] )),
+            RefinedType("$arg1", t_int, LiquidApp(">", [LiquidVar("$arg1"), LiquidVar("$arg0")]))
+        ]
+    ),
+    (
[…]

### 2023-03-07 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Added parametrized test for argument replacement

@@ -90,7 +90,7 @@ def substitute_argument_names(arg_names: tp.List[str], arg_types: tp.List[Refine
     for i in range(len(arg_names)):
         ref_name = arg_types[i].name
         for j in range(i + 1, len(arg_names)):
-            arg_types[j] = substitution_in_type(arg_types[j], arg_names[i], ref_name)
+            arg_types[j].refinement = substitution_in_liquid(arg_types[j].refinement, LiquidVar(arg_names[i]), ref_name)
 
 class LiquidLayer(lark.visitors.Interpreter):
     def __init__(self):
@@ -281,6 +281,7 @@ class LiquidLayer(lark.visitors.Interpreter):
         self.__types = {}
 
         substitute_refinement_names(fun_type)
+        substitute_argument_names(arg_names, fun_type)
 
         for i in range(len(arg_names)):
             self.__ctx = VariableBinder(self.__ctx, arg_names[i], fun_type[i])

### 2023-03-07 code/LayeredTypes/tests/test_liquid.py

@@ -236,28 +236,7 @@ class TestLiquidLayer(unittest.TestCase):
         type_expected = parse_type("{x:Int | x > v}")
         self.assertEqual(type_expected, e.argument_type)
 
-class TestSubstitutions(unittest.TestCase):
-    def test_single_refinement_substituted(self):
-        type_before = parse_type("{v:Int | v > 0}")
-        type_after = RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] ))
-
-        substitute_refinement_names([type_before, None])
-        self.assertEqual(type_after, type_before)
-
-
-    def test_substitute_arg_names_single_refinement(self):
-        self.assertTrue(False)
-
-    def test_substitute_arg_names_multiple_refinements_no_ref(self):
-        self.assertTrue(False)
-
-    def test_substitute_arg_names_multiple_refinements_mixed(self):
-        self.assertTrue(False)
-
-    def test_substitute_arg_names_multiple_refinements_references(self):
-        self.assertTrue(False)
-
-test_vals = [
+refinement_replacement_test_vals = [
     ([parse_type("{v:Int | v > 0}")],
      [RefinedType("$arg0", t_int,
                   LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] ))
@@ -314,7 +293,80 @@ test_vals = [
     )
 ]
 
-@pytest.mark.parametrize("types_before,types_after", test_vals)
+argument_names_substitution_test_vals = [
+    (
+        ["x"],
+        [
+            parse_type("{v:Int | v > 0}"),
+        ],
+        [
+            parse_type("{v:Int | v > 0}"),
+        ]
+    ),
+    (
+        ["x", "y"],
+        [
+            parse_type("{v1:Int | v1 > 0}"),
+            parse_type("{v2:Int | v2 > v1}"),
+        ],
+        [
+            parse_type("{v1:Int | v1 > 0}"),
+            parse_type("{v2:Int | v2 > x}")
+        ]
+    ),
+    (
+        ["x", "y", "z"],
+        [
+            parse_type("{v1:Int | v1 > 0}"),
+            parse_type("{v2:Int | v2 > v1}"),
+            parse_type("{v3:Int | v3 > v2}"),
+        ],
+        [
+            parse_type("{v1:Int | v1 > 0}"),
+            parse_type("{v2:Int | v2 > x}"),
+            parse_type("{v3:Int | v3 > y}")
+        ]
[…]

### 2023-03-07 code/LayeredTypes/tests/test_liquid.py
* Added end-to-end tests for function definition replacement of refined types

@@ -356,6 +356,51 @@ argument_names_substitution_test_vals = [
         ]
     ),
     ]
+
+end_to_end_test_vals = [
+    (
+        ["x"],
+        [ parse_type("{x:Int | x > 0}") ],
+        [ RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] )) ]
+    ),
+    (
+        ["x", "y"],
+        [
+            parse_type("{y:Int | y > 0}"),
+            parse_type("{x:Int | x > y}")
+        ],
+        [
+            RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] )),
+            RefinedType("$arg1", t_int, LiquidApp(">", [LiquidVar("$arg1"), LiquidVar("x")]))
+        ]
+    ),
+    (
+        ["x", "y", "z"],
+        [
+            parse_type("{z:Int | z > 0}"),
+            parse_type("{x:Int | x > z}"),
+            parse_type("{y:Int | y > x}")
+        ],
+        [
+            RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] )),
+            RefinedType("$arg1", t_int, LiquidApp(">", [LiquidVar("$arg1"), LiquidVar("x")])),
+            RefinedType("$arg2", t_int, LiquidApp(">", [LiquidVar("$arg2"), LiquidVar("y")])),
+        ]
+    ),
+    (
+        ["x", "y", "z"],
+        [
+            parse_type("{x:Int | x > 0}"),
+            parse_type("{x:Int | x > 1}"),
+            parse_type("{x:Int | x > 2}"),
+        ],
+        [
+            RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)] )),
+            RefinedType("$arg1", t_int, LiquidApp(">", [LiquidVar("$arg1"), LiquidLiteralInt(1)])),
+            RefinedType("$arg2", t_int, LiquidApp(">", [LiquidVar("$arg2"), LiquidLiteralInt(2)])),
+        ]
+    ),
+]
 @pytest.mark.parametrize("types_before,types_after", refinement_replacement_test_vals)
 def test_refinement_substitutions(types_before, types_after):
     substitute_refinement_names([*types_before, None])
[…]

### 2023-03-07 code/LayeredTypes/tests/test_liquid.py
* Fixed tests failing due to new variable renaming in function definition

@@ -148,7 +148,7 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual(6, e.lineno)
         self.assertEqual(5, e.offset)
         type_expected = parse_type("{v:Int | v > 0}")
-        type_actual = parse_type("{v:Int | v >= 0}")
+        type_actual = RefinedType("$arg0", t_int, LiquidApp(">=", [LiquidVar("$arg0"), LiquidLiteralInt(0)]))
         self.assertEqual(type_actual, e.type_actual)
         self.assertEqual(type_expected, e.type_expected)
 
@@ -164,7 +164,7 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual("LiquidSubtypeException", e.__class__.__name__)
         self.assertEqual(10, e.lineno)
         self.assertEqual(1, e.offset)
-        type_expected = parse_type("{v:Int | v > 0}")
+        type_expected = RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)]))
         type_actual = parse_type("{v:Int | v == 0}")
         self.assertEqual(type_actual, e.type_actual)
         self.assertEqual(type_expected, e.type_expected)
@@ -185,7 +185,7 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual(8, e.lineno)
         self.assertEqual(5, e.offset)
         type_expected = parse_type("{v:Int | v > 0}")
-        type_actual = parse_type("{v:Int | v >= 0}")
+        type_actual = RefinedType("$arg1", t_int, LiquidApp(">=", [LiquidVar("$arg1"), LiquidLiteralInt(0)]))
         self.assertEqual(type_actual, e.type_actual)
         self.assertEqual(type_expected, e.type_expected)
 

### 2023-03-07 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Adapted tests to new replacement behavior

@@ -1,5 +1,5 @@
 from collections import Counter
-from copy import copy
+from copy import copy, deepcopy
 import typing as tp
 
 import lark
@@ -257,7 +257,9 @@ class LiquidLayer(lark.visitors.Interpreter):
 
         # First check if we have a type definition for the function
         fun_identifier = tree.children[0].value
-        fun_type = tree.get_layer_annotation("liquid", fun_identifier, "function_type")
+        # We want to explicitly copy the function type, as we will modify it
+        # If we wouldn't copy it this would have implications when calling the function
+        fun_type = deepcopy(tree.get_layer_annotation("liquid", fun_identifier, "function_type"))
 
         if fun_type is None:
             raise LiquidTypeUndefinedError(fun_identifier, tree.meta.line, tree.meta.column)

### 2023-03-07 code/LayeredTypes/tests/test_liquid.py

@@ -164,7 +164,7 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual("LiquidSubtypeException", e.__class__.__name__)
         self.assertEqual(10, e.lineno)
         self.assertEqual(1, e.offset)
-        type_expected = RefinedType("$arg0", t_int, LiquidApp(">", [LiquidVar("$arg0"), LiquidLiteralInt(0)]))
+        type_expected = parse_type("{v:Int | v > 0}")
         type_actual = parse_type("{v:Int | v == 0}")
         self.assertEqual(type_actual, e.type_actual)
         self.assertEqual(type_expected, e.type_expected)
@@ -193,8 +193,6 @@ class TestLiquidLayer(unittest.TestCase):
         compiler = get_compiler(layer_path="layer_implementations")
         src_file = full_path("/test_code/liquid/fun_def_multiple_args_fail_oncall.fl")
 
-        compiler.typecheck(src_file)
-
         with self.assertRaises(LayerException) as context:
             compiler.typecheck(src_file)
 
@@ -203,7 +201,7 @@ class TestLiquidLayer(unittest.TestCase):
         self.assertEqual("LiquidSubtypeException", e.__class__.__name__)
         self.assertEqual(13, e.lineno)
         self.assertEqual(1, e.offset)
-        type_expected = parse_type("{v:Int | x > v}")
+        type_expected = parse_type("{x:Int | x > liquidFun_v}")
         type_actual = parse_type("{v:Int | v == 1}")
         self.assertEqual(type_actual, e.type_actual)
         self.assertEqual(type_expected, e.type_expected)

### 2023-03-08 paper/types2023.tex
Added paper template for TYPES2023 conference

% easychair.tex,v 3.5 2017/03/15

\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[EPiCempty]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
%\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}

\usepackage{doc}

% use this if you have a long article and want to create an index
% \usepackage{makeidx}

% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

% Some of our commands for this guide.
%
\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

%\makeindex

%% Front Matter
%%
% Regular title as in the article class.
%
\title{LayeredTypes -- A structured approach to arbitrarily combine type systems}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Lukas Abelt\inst{1}
\and
    Alcides Fonseca\inst{2}
}

% Institutes for affiliations are also joined by \and,
\institute{
  Saarland University,
  Saarbrücken, Germany\\
  \email{lukasa@cs.fs.uni-saarland.de}
\and
   Universidade de Lisboa,
   Lisbon, Portugal\\
   \email{amfonseca@fc.ul.pt}\\
}

%  \authorrunning{} has to be set for the shorter version of the authors' names;
% otherwise a warning will be rendered in the running heads. When processed by
% EasyChair, this command is mandatory: a document without \authorrunning
% will be rejected by EasyChair

\authorrunning{Abelt and Fonseca}

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{Layered Types}

\begin{document}

\maketitle

\begin{abstract}
\end{abstract}

% The table of contents below is added for your convenience. Please do not use
% the table of contents if you are preparing your paper for publication in the
% EPiC Series or Kalpa Publications series

\setcounter{tocdepth}{2}
{\small
\tableofcontents}

%\section{To mention}
%
%Processing in EasyChair - number of pages.
%
%Examples of how EasyChair processes papers. Caveats (replacement of EC
%class, errors).

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}

%------------------------------------------------------------------------------

%------------------------------------------------------------------------------
\end{document}

### 2023-03-08 paper/types2023.tex
Added first draft of TYPES2023 submission

@@
 
+\lstset{
+  frame=none,
+  xleftmargin=2pt,
+  stepnumber=1,
+  numbers=left,
+  numbersep=5pt,
+  numberstyle=\ttfamily\tiny\color[gray]{0.3},
+  belowcaptionskip=\bigskipamount,
+  escapeinside={*'}{'*},
+  tabsize=2,
+  emphstyle={\bf},
+  commentstyle=\color{ForestGreen},
+  stringstyle=\mdseries\rmfamily,
+  showspaces=false,
+  keywordstyle=\bfseries\rmfamily,
+  columns=flexible,
+  basicstyle=\small\sffamily,
+  showstringspaces=false,
+  morecomment=[l]\%,
+}
+\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
 % In order to save space or manage large tables or figures in a
@@
 
+\newcommand{\LayeredTypes}{\textsc{LayeredTypes}}
 %\makeindex
@@
 %
-\title{LayeredTypes -- A structured approach to arbitrarily combine type systems}
+\title{LayeredTypes -- Independently combine type systems}
 
@@
 \author{
-Lukas Abelt\inst{1}
+Author 1\inst{1}
 \and
-    Alcides Fonseca\inst{2}
+Author 2\inst{2}
 }
@@
 \institute{
-  Saarland University,
-  Saarbrücken, Germany\\
-  \email{lukasa@cs.fs.uni-saarland.de}
+  Institute 1,
+  Location 1\\
+  \email{mail1}
 \and
-   Universidade de Lisboa,
-   Lisbon, Portugal\\
-   \email{amfonseca@fc.ul.pt}\\
+   Intitute 2,
+   Location 2\\
+   \email{mail2}\\
 }
@@
 
-\authorrunning{Abelt and Fonseca}
+\authorrunning{Authorrunning}
 
@@
 
-\begin{abstract}
-\end{abstract}
-
 % The table of contents below is added for your convenience. Please do not use
@@
 
-\setcounter{tocdepth}{2}
-{\small
-\tableofcontents}
+%\setcounter{tocdepth}{2}
+%{\small
+%\tableofcontents}
 
@@
 %------------------------------------------------------------------------------
-\section{Introduction}
-\label{sect:introduction}
+%\section{Context}
+\label{sec:context}
+Most programmers and computer scientists will be familiar with simple type systems that ensure that the code they write are type safe in the context of the language they are writing it in.
+However, much more sophisticated type systems exist that can be used to ensure specific properties of a program.
+For example Session Types \cite{session} can be used to show correctness of concurrent programs, linear type systems \cite{linear} ensure that objects are used exactly once and Refinement Types \cite{refinement} can narrow types by imposing predicates onto them.
+Another example is the use of liquid types to show bounds on resource usage \cite{liquidate-assets}.
+
+Depending on the specific type systems, they can be viewed as independently and orthogonal from one another, however sometimes type systems also depend on other type systems and thus extending them.
+One example of this are refinement types which build on top of traditional type systems and extend these types by making them more specific.
+Most importantly not all type systems are useful for any program; When developing some low-level code meant to be run in an operating systems' kernel, they might want to use the help of a specific type system that ensure that they do not exceed certain bounds in memory usage.
+However, when developing an application that is meant to be run in a highly distributed environment, using Session Types might be a more appropriate type system to ensure the correct implementation of the communication of the distirbuted components.
+
+%\section{Challenges in Type Systems}
+\label{sec:challenges-in-type-systems}
+Most commonly used programming languages such as Java, C++ or Haskell embed a single type system that checks, either statically or dynamically, that functions and variables are used in a compatible way.
+To use more sophisticated type systems such as Liquid Types, one has to rely on specific tools or compiler extensions such as LiquidHaskell\cite{LiquidHaskell} or LiquidJava\cite{LiquidJava}.
+This leads to the issue that it is usually not trivially possible to combine different type systems in a single program.
+However in some cases one might require multiple different type systems to ensure different properties of a program.
+
+Consider the very simple code example in \autoref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line.
+Even in this short example there are multiple different concerns to be addressed:
+\begin{itemize}
+	\item From a type-safety perspective, one wants to make sure that e.g. \texttt{get(...)} is only called with a list and an integer index as arguments
+	\item Additionally, one wants to ensure that \texttt{get} is never accessed with an index that exceeds the lists length to avoid out-of-bounds access errors at runtime
+	\item When working with resources such as file descriptors one wants to ensure that these are in a proper (closed) state by the end of the program execution
+\end{itemize}
+Optimally, these properties should be verified independently from each other wherever possible.
+For example tracking the state of a resource can be done independently of it's specific type, so in case that for whatever reason this check fails we still want to be able to independently verify whether or not the state behaves as we intend it.
+
+\begin{minipage}{0.4\linewidth}
+\begin{lstlisting}[caption={Simple example code},label={lst:code_before}]
+printLines lines idx len {
+  if len != idx then {
+    line = get(lines, idx)
+    print(line)
+    printLines(lines, idx+1, len)
+  }
+}
+
+filename = "input.txt"
+fileDescriptor = createFD(filename)
+
+open(fileDescriptor)
+
+lines = readLines(fileDescriptor)
+len = length(lines)
+printLines(lines, 0, len)
+
+close(fileDescriptor)
+\end{lstlisting}
+\end{minipage}%
+\begin{minipage}{0.59\linewidth}
+\begin{lstlisting}[caption={Annotations for \LayeredTypes},label={lst:code_after}]
+-- State Layer definitions
+createFD :: state :: {} -> { Closed }
+openFile :: state :: {Closed => Open} -> {}
+readLines :: state :: {Open => Consumed} -> {}
+closeFile :: state :: {Consumed => Closed} -> {}
+
+-- Type layer definitions
+get :: types :: List -> int -> string
+length :: types :: List -> int
+createFD :: types :: string -> FileHandle
+open :: types :: FileHandle -> void
+readLines :: types :: FileHandle -> List
+close :: types :: FileHandle -> void
+print :: types :: string -> void
+printLines :: types :: List -> int -> int -> void
+
+-- Liquid layer definitions
+length :: {List | true} -> {v:int | v>=0}
+printLines :: liquid :: { List | true } -> { l:int | l>=0 } -> { i:int | i<=l }
+get :: liquid :: { List | true } -> { i:int | i<len }
+
+-- State requirement at the end of the program
+fileDescriptor :: state :: {Closed}	
+\end{lstlisting}
+\end{minipage}
+
+A very similar issue can also be observed when operating on Liquid Types: In certain cases we want to define predicates on orthogonal properties that could be verified independently from one another; However, current implementation of liquid types do not allow us to define distinct predicates to be verified independently.
+In such cases we have to merge all relevant properties into one joint predicate which can make it hard to identify the specific issue when the verification fails.
+
+%\section{Proposed Approach}
+\label{sec:proposed-approach}
+
+To tackle these issues, we present \LayeredTypes.
+In our approach we view different type systems as indepent layers that can be verified in an incremental way.
+Layers can either verify properties in a complete stand-alone matter, or they can define dependencies to other layers which allow to build more complex type systems in an incremental manner.
+
+We want to note an important distinction to Gradual Typing\cite{gradual}: While the latter is concerned with only providing partial typing information and making it gradually more precise, our approach does not focus on the information provided to the verifier, but rather on \textit{how} this informati.
+
+In our work we present a python-based demonstration framework that implements this layered verification approach.
+In a similar manner to passes in compilers, users can provide their own implementations for verification layers that suit their workflow.
+In \autoref{lst:code_after} we see a set of annotations that can be added to the base code of \autoref{lst:code_before} to tackle the issues described.
+We define three separate layers \texttt{State}, \texttt{types} and \texttt{liquid}.
+Our framework will build a dependency graph of the layers which allows the type "path" to be verified independently from the state path.
+
+During our work we implemented two specific use-cases, one similar to the code snippet presented above and one using multiple liquid layers to verify different liquid properties independently.
 

### 2023-03-08 paper/types2023.tex
Rephrased some parts TYPES submission

@@
 %------------------------------------------------------------------------------
-%\section{Context}
+\section{Context}
 \label{sec:context}
@@
 However, much more sophisticated type systems exist that can be used to ensure specific properties of a program.
-For example Session Types \cite{session} can be used to show correctness of concurrent programs, linear type systems \cite{linear} ensure that objects are used exactly once and Refinement Types \cite{refinement} can narrow types by imposing predicates onto them.
-Another example is the use of liquid types to show bounds on resource usage \cite{liquidate-assets}.
+Examples for this are Session Types \cite{session}, used for concurrent programs, linear type systems \cite{linear} or Refinement Types \cite{refinement}.
 
-Depending on the specific type systems, they can be viewed as independently and orthogonal from one another, however sometimes type systems also depend on other type systems and thus extending them.
-One example of this are refinement types which build on top of traditional type systems and extend these types by making them more specific.
-Most importantly not all type systems are useful for any program; When developing some low-level code meant to be run in an operating systems' kernel, they might want to use the help of a specific type system that ensure that they do not exceed certain bounds in memory usage.
-However, when developing an application that is meant to be run in a highly distributed environment, using Session Types might be a more appropriate type system to ensure the correct implementation of the communication of the distirbuted components.
+Depending on the specific type systems, they can be viewed as independently and orthogonal from one another, however sometimes type systems can also extend existing ones such as refinement types which build on top of traditional type systems and make these more specific.
+Most importantly not all type systems are useful for any program; When developing some low-level code meant to be run in an operating systems' kernel, they might want to use the help of a specific type system that ensure that they do not exceed certain bounds in memory usage such as shown in \cite{liquidate-assets}.
+However, when developing an application that is meant to be run in a highly distributed environment, using Session Types might be a more appropriate type system to ensure the correct implementation of the communication of the distributed components.
 
-%\section{Challenges in Type Systems}
+\section{Challenges in Type Systems}
 \label{sec:challenges-in-type-systems}
@@
 
-Consider the very simple code example in \autoref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line.
-Even in this short example there are multiple different concerns to be addressed:
-\begin{itemize}
-	\item From a type-safety perspective, one wants to make sure that e.g. \texttt{get(...)} is only called with a list and an integer index as arguments
-	\item Additionally, one wants to ensure that \texttt{get} is never accessed with an index that exceeds the lists length to avoid out-of-bounds access errors at runtime
-	\item When working with resources such as file descriptors one wants to ensure that these are in a proper (closed) state by the end of the program execution
-\end{itemize}
+Consider the simple code example in \autoref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line.
+In this short snippet, there are multiple concerns to be addressed.
+Firstly, one wants to ensure that all function calls are done with variables of the appropriate data types.
+Secondly, to avoid errors due to an out of bound access, \texttt{get} should only be called with an index that exceeds the list length.
+Lastly, when working with resources such as a file descriptor it might be useful to track it's state and ensure that it is in a proper (closed) state at the end of program execution.
+%\begin{itemize}
+%	\item From a type-safety perspective, one wants to make sure that e.g. \texttt{get(...)} is only called with a list and an integer index as arguments
+%	\item Additionally, one wants to ensure that \texttt{get} is never accessed with an index that exceeds the lists length to avoid out-of-bounds access errors at runtime
+%	\item When working with resources such as file descriptors one wants to ensure that these are in a proper (closed) state by the end of the program execution
+%\end{itemize}
+
 Optimally, these properties should be verified independently from each other wherever possible.
@@
 
-%\section{Proposed Approach}
+\section{Proposed Approach}
 \label{sec:proposed-approach}
@@
 In our approach we view different type systems as indepent layers that can be verified in an incremental way.
-Layers can either verify properties in a complete stand-alone matter, or they can define dependencies to other layers which allow to build more complex type systems in an incremental manner.
+Layers can either verify properties in a complete stand-alone matter, or they can define dependencies to other layers which allow to build more complex type systems.
 
-We want to note an important distinction to Gradual Typing\cite{gradual}: While the latter is concerned with only providing partial typing information and making it gradually more precise, our approach does not focus on the information provided to the verifier, but rather on \textit{how} this informati.
+We want to note an important distinction to Gradual Typing\cite{gradual}: While the latter is concerned with only providing partial typing information and making it gradually more precise, our approach does not focus on the information provided to the verifier, but rather on \textit{how} this information.
 
 In our work we present a python-based demonstration framework that implements this layered verification approach.
-In a similar manner to passes in compilers, users can provide their own implementations for verification layers that suit their workflow.
+Users can provide their own implementations for verification layers and define dependencies between them.
 In \autoref{lst:code_after} we see a set of annotations that can be added to the base code of \autoref{lst:code_before} to tackle the issues described.
-We define three separate layers \texttt{State}, \texttt{types} and \texttt{liquid}.
-Our framework will build a dependency graph of the layers which allows the type "path" to be verified independently from the state path.
+We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}.
+Internally, the framework will build a dependency graph thus allowing to verify properties independently from one another where appropriate.
 
-During our work we implemented two specific use-cases, one similar to the code snippet presented above and one using multiple liquid layers to verify different liquid properties independently.
+We believe that this layered, incremental approach can help build more powerful and independent type systems while at the same time making it simpler to understand the errors that might arise during verification.
 

### 2023-03-09 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Extracted variable renaming in function call to separate function

@@ -92,6 +92,17 @@ def substitute_argument_names(arg_names: tp.List[str], arg_types: tp.List[Refine
         for j in range(i + 1, len(arg_names)):
             arg_types[j].refinement = substitution_in_liquid(arg_types[j].refinement, LiquidVar(arg_names[i]), ref_name)
 
+
+def rename_variable(context, expected_arg_types, fun_identifier, i):
+    original_name = expected_arg_types[i].name
+    free_var_name = f"{fun_identifier}_{original_name}"
+    free_var_name = make_name_unique(free_var_name, context)
+
+    for j in range(i + 1, len(expected_arg_types)):
+        expected_arg_types[j] = substitution_in_type(expected_arg_types[j], Var(free_var_name), original_name)
+    return free_var_name
+
+
 class LiquidLayer(lark.visitors.Interpreter):
     def __init__(self):
         self.__types = {}
@@ -237,15 +248,8 @@ class LiquidLayer(lark.visitors.Interpreter):
             # As other refinements may reference it
             # To ensure uniqueness, we rename the variable
             assert type(arg_type) == RefinedType
-            original_name = expected_arg_types[i].name
-            free_var_name = f"{fun_identifier}_{original_name}"
-            free_var_name = make_name_unique(free_var_name, context)
-
-            # We need to rename the variable in all the refinements for the remaining arguments
-            # TODO: Differentiate cases where different arguments use the same variable name
-            # This is possible as long as the variable is not referenced in another refinement
-            for j in range(i+1, len(expected_arg_types)):
-                expected_arg_types[j] = substitution_in_type(expected_arg_types[j], Var(free_var_name), original_name)
+            # TODO: do not always do this, only if the variable name is unique
+            free_var_name = rename_variable(context, expected_arg_types, fun_identifier, i)
 
             context = context.with_var(free_var_name, arg_type)
             self.__ctx = context

### 2023-03-09 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Refactored fun_call renaming to correctly consider arguments with identical names

@@ -93,14 +93,14 @@ def substitute_argument_names(arg_names: tp.List[str], arg_types: tp.List[Refine
             arg_types[j].refinement = substitution_in_liquid(arg_types[j].refinement, LiquidVar(arg_names[i]), ref_name)
 
 
-def rename_variable(context, expected_arg_types, fun_identifier, i):
-    original_name = expected_arg_types[i].name
+def rename_variable(context, original_name, fun_identifier, all_arg_names, remaining_args):
     free_var_name = f"{fun_identifier}_{original_name}"
     free_var_name = make_name_unique(free_var_name, context)
 
-    for j in range(i + 1, len(expected_arg_types)):
-        expected_arg_types[j] = substitution_in_type(expected_arg_types[j], Var(free_var_name), original_name)
-    return free_var_name
+    if all_arg_names.count(original_name) == 1:
+        for i in range(len(remaining_args)):
+            remaining_args[i] = substitution_in_type(remaining_args[i], Var(free_var_name), original_name)
+    return free_var_name, remaining_args
 
 
 class LiquidLayer(lark.visitors.Interpreter):
@@ -231,6 +231,7 @@ class LiquidLayer(lark.visitors.Interpreter):
                                               tree.meta.column)
 
         context = tree.get_layer_annotation("liquid", "contexts", "context")
+        all_arg_names = [x.name for x in expected_arg_types]
         # Check that the types of the arguments are correct
         for i in range(actual_num_args):
             arg_type, context = self.visit(tree.children[i+1])
@@ -248,8 +249,10 @@ class LiquidLayer(lark.visitors.Interpreter):
             # As other refinements may reference it
             # To ensure uniqueness, we rename the variable
             assert type(arg_type) == RefinedType
-            # TODO: do not always do this, only if the variable name is unique
-            free_var_name = rename_variable(context, expected_arg_types, fun_identifier, i)
+
+            original_name = expected_arg_types[i].name
+            free_var_name, expected_arg_types[i+1:] = rename_variable(context, original_name, fun_identifier,
+                                                                      all_arg_names , expected_arg_types[i+1:])
 
             context = context.with_var(free_var_name, arg_type)
             self.__ctx = context

### 2023-03-09 code/LayeredTypes/tests/test_liquid.py
* Added tests for function call renaming

@@ -5,8 +5,11 @@ import pytest
 from aeon.core.liquid import LiquidApp, LiquidVar, LiquidLiteralInt
 from aeon.core.types import RefinedType, t_int
 from aeon.frontend.parser import parse_type
+from aeon.typing.context import EmptyContext
+from aeon.utils.ctx_helpers import build_context
 from compiler.Exceptions import LayerException
-from compiler.interpreters.LiquidTypeChecker import substitute_argument_names, substitute_refinement_names
+from compiler.interpreters.LiquidTypeChecker import substitute_argument_names, substitute_refinement_names, \
+    rename_variable
 from utils import typecheck_correct_file, get_compiler, full_path
 
 
@@ -399,6 +402,112 @@ end_to_end_test_vals = [
         ]
     ),
 ]
+
+fun_call_var_renaming_test_vals = [
+    (
+        (
+            EmptyContext(),
+            "arg",
+            "fun",
+            ["arg"],
+            []
+        ),
+        (
+            "fun_arg",
+            []
+        )
+    ),
+    (
+        (
+            EmptyContext(),
+            "arg",
+            "fun",
+            ["arg","v"],
+            [parse_type("{v:Int | v > 0}")]
+        ),
+        (
+            "fun_arg",
+            [parse_type("{v:Int | v > 0}")]
+        )
+    ),
+    (
+        (
+            EmptyContext(),
+            "arg",
+            "fun",
+            ["arg","v"],
+            [parse_type("{v:Int | v > arg}")]
+        ),
+        (
+            "fun_arg",
+            [parse_type("{v:Int | v > fun_arg}")]
+        )
+    ),
+    (
+        (
+            EmptyContext(),
+            "arg",
+            "fun",
+            ["arg","v","v"],
+            [
+                parse_type("{v:Int | v > 0}"),
+                parse_type("{v:Int | v > arg}")
+            ]
+        ),
+        (
+            "fun_arg",
+            [
+                parse_type("{v:Int | v > 0}"),
+                parse_type("{v:Int | v > fun_arg}")
+            ]
+        )
+    ),
+    (
+        (
+            EmptyContext(),
+            "v",
+            "fun",
+            ["v","v","v"],
+            [
+                parse_type("{v:Int | v > 0}"),
+                parse_type("{v:Int | v > 1}")
+            ]
+        ),
+        (
+            "fun_v",
+            [
+                parse_type("{v:Int | v > 0}"),
+                parse_type("{v:Int | v > 1}")
+            ]
+        )
+    ),
+    (
[…]

### 2023-03-10 paper/types2023.tex
Added references to types submission

@@
 \begin{document}
-
 \maketitle
@@
 
-We want to note an important distinction to Gradual Typing\cite{gradual}: While the latter is concerned with only providing partial typing information and making it gradually more precise, our approach does not focus on the information provided to the verifier, but rather on \textit{how} this information.
+We want to note an important distinction to Gradual Typing\cite{gradual}\cite{gradual-objects}: While the latter is concerned with only providing partial typing information and making it gradually more precise, our approach does not focus on the information provided to the verifier, but rather on \textit{how} this information.
 
@@
 %------------------------------------------------------------------------------
+\bibliographystyle{plain}
+\bibliography{literature}
 

### 2023-03-10 paper/types2023.tex
Small fix and add acknowledgement

@@
 
+\usepackage{cleveref}
 \usepackage{doc}
@@
 %
+\authorrunning{Abelt and Fonseca}
+
 \newcommand{\easychair}{\textsf{easychair}}
@@
 %
-\title{LayeredTypes -- Independently combine type systems}
+\title{LayeredTypes -- Combining dependent and independent type systems}
 
@@
 \author{
-Author 1\inst{1}
+Lukas Abelt\inst{1,2}
 \and
-Author 2\inst{2}
+Alcides Fonseca\inst{1}
 }
@@
 \institute{
-  Institute 1,
-  Location 1\\
-  \email{mail1}
+  LASIGE,
+  Faculdade de Ciências da Universidade de Lisboa, Portugal\\
+  \email{labelt@lasige.di.fc.ul.pt},
+  \email{alcides@ciencias.ulisboa.pt}
 \and
-   Intitute 2,
-   Location 2\\
-   \email{mail2}\\
+   Saarland University,
+   Germany\\
 }
@@
 % will be rejected by EasyChair
-
-\authorrunning{Authorrunning}
 
@@
 \label{sec:context}
-Most programmers and computer scientists will be familiar with simple type systems that ensure that the code they write are type safe in the context of the language they are writing it in.
-However, much more sophisticated type systems exist that can be used to ensure specific properties of a program.
-Examples for this are Session Types \cite{session}, used for concurrent programs, linear type systems \cite{linear} or Refinement Types \cite{refinement}.
+Most programmers and computer scientists will be familiar with simple type systems that ensure that the code they write is type safe in the context of the language they are writing it in.
+However, much more sophisticated type systems exist that can be used to ensure specific properties of a program, such as resource usage (Linear Types~\cite{linear}), value predicates (Liquid Types~\cite{DBLP:conf/pldi/RondonKJ08}), or correct following of a distributed protocol (Session Types~\cite{session}).
 
-Depending on the specific type systems, they can be viewed as independently and orthogonal from one another, however sometimes type systems can also extend existing ones such as refinement types which build on top of traditional type systems and make these more specific.
-Most importantly not all type systems are useful for any program; When developing some low-level code meant to be run in an operating systems' kernel, they might want to use the help of a specific type system that ensure that they do not exceed certain bounds in memory usage such as shown in \cite{liquidate-assets}.
-However, when developing an application that is meant to be run in a highly distributed environment, using Session Types might be a more appropriate type system to ensure the correct implementation of the communication of the distributed components.
+Some of these type systems can be dependent or independent from each other.
+Both Liquid Types and Session Types require a base type system with primitive types (such as int, bool) to exist, but they can be orthogonal to each other, as you can have a Liquid Type whose base type is int, and a Session Type that also relies on the base value being int, ignore the liquid predicate.
 
-\section{Challenges in Type Systems}
-\label{sec:challenges-in-type-systems}
-Most commonly used programming languages such as Java, C++ or Haskell embed a single type system that checks, either statically or dynamically, that functions and variables are used in a compatible way.
-To use more sophisticated type systems such as Liquid Types, one has to rely on specific tools or compiler extensions such as LiquidHaskell\cite{LiquidHaskell} or LiquidJava\cite{LiquidJava}.
-This leads to the issue that it is usually not trivially possible to combine different type systems in a single program.
-However in some cases one might require multiple different type systems to ensure different properties of a program.
+Given this plethora of type systems, different applications require a different subset of type systems.
+Not all applications require these advanced type systems (e.g., fully dependent types), trading off static guarantees for (even if short-term) productivity.
+Untyped languages have been popular in both web applications, scripting and data science.
+On the other hand, if you are writing a device driver, you will probably want to take advantage of a type system that guarantees that memory usage is constrained to a given bound~\cite{liquidate-assets}.
+Or if you are writing a complex, distributed protocol, you might want to take advantage of Session Types for that part of the program.
 
-Consider the simple code example in \autoref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line.
+In this talk, we try to address the challenge of how can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.
+
+ % Alcides is here
+
+ To illustrate this in a small example, consider the program in \Cref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line.
 In this short snippet, there are multiple concerns to be addressed.
 Firstly, one wants to ensure that all function calls are done with variables of the appropriate data types.
-Secondly, to avoid errors due to an out of bound access, \texttt{get} should only be called with an index that exceeds the list length.
+Secondly, to avoid errors due to an out of bound access, \texttt{get} should only be called with an index that does not exceed the list length.
 Lastly, when working with resources such as a file descriptor it might be useful to track it's state and ensure that it is in a proper (closed) state at the end of program execution.
+Note that the same standard library functions (\lstinline|createFD|, \lstinline|readLines|, etc...) may be needed in other programs with different type systems requirements.
+
 %\begin{itemize}
@@
 
-Optimally, these properties should be verified independently from each other wherever possible.
-For example tracking the state of a resource can be done independently of it's specific type, so in case that for whatever reason this check fails we still want to be able to independently verify whether or not the state behaves as we intend it.
+These properties can be verified in two major ways.
+
+All features from all type systems can be combined in single, monolithic type system, containing all the complexity that all the different type systems entail.
+A very similar issue can also be observed when operating on Liquid Types: In certain cases we want to define predicates on orthogonal properties that could be verified independently from one another; However, current implementation of liquid types do not allow such distinction, leading to confusing and unintelligible error messages~\cite{wits-error-messages}.
 
@@
 
-A very similar issue can also be observed when operating on Liquid Types: In certain cases we want to define predicates on orthogonal properties that could be verified independently from one another; However, current implementation of liquid types do not allow us to define distinct predicates to be verified independently.
-In such cases we have to merge all relevant properties into one joint predicate which can make it hard to identify the specific issue when the verification fails.
-
 \section{Proposed Approach}
@@
 
-To tackle these issues, we present \LayeredTypes.
-In our approach we view different type systems as indepent layers that can be verified in an incremental way.
-Layers can either verify properties in a complete stand-alone matter, or they can define dependencies to other layers which allow to build more complex type systems.
+We propose a second, more principled alternative: \LayeredTypes.
+In \LayeredTypes, developers can write programs, but can also define additional type systems as layers.
+Each layer defines the basic types, and how typechecking happens.
+Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.
 
-We want to note an important distinction to Gradual Typing\cite{gradual}\cite{gradual-objects}: While the latter is concerned with only providing partial typing information and making it gradually more precise, our approach does not focus on the information provided to the verifier, but rather on \textit{how} this information.
+Type checking of programs can be partial in the sense that a program may only use some layers (even if the libraries used are defined in more layers), requiring only that each function contains type information for that layer and all dependencies.
+The missing layers can be added over time, if they are found to be relevant.
 
-In our work we present a python-based demonstration framework that implements this layered verification approach.
+We want to note an important distinction to Gradual Typing~\cite{gradual,gradual-objects}: Gradual Typing allows to only provide partial typing information in one system, but does not allow to choose different sets of types to verify.
+It allows to blend between having no types (0) and having types (1).
+Our proposal supports multiple type systems, with different combinations among them.
+
+We evaluated our approach in a prototype language.
 Users can provide their own implementations for verification layers and define dependencies between them.
@@
 
+\input{chapters/acknowledgements.tex}
+
+\bibliographystyle{plain}
+\bibliography{bibliography}
+
+\end{document}
 %------------------------------------------------------------------------------

### 2023-03-10 paper/types2023.tex
References
Small pass

@@
 %
-\authorrunning{Abelt and Fonseca}
-
 \newcommand{\easychair}{\textsf{easychair}}
@@
 
+\authorrunning{Authorrunning}
+
 % \titlerunning{} has to be set to either the main title or its shorter
@@
 
-In this talk, we try to address the challenge of how can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.
+\section{Challenges in Type Systems}
+\label{sec:challenges-in-type-systems}
+
+Right now, to choose a particular type system, developers have to change programming languages, which might introduce interoperability problems with other components of their system.
+Furthermore, authors of new programming languages have to write extensive standard libraries in order for their language to be adopted by the masses.
+ Most commonly used programming languages such as Java, C++ or Haskell embed a single type system that checks, either statically or dynamically, that functions and variables are used in a compatible way.
+
+ To use more sophisticated type systems such as Liquid Types, developers either port their programs to a different programming language, or rely on compiler plugins to add another type-checking phase, like in LiquidHaskell\cite{LiquidHaskell} or LiquidJava\cite{LiquidJava}.
+However, it is not clear how multiple type systems would interact in the same programming language.
+
+ \paragraph{Problem statement:} How can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.
 
@@
 Firstly, one wants to ensure that all function calls are done with variables of the appropriate data types.
-Secondly, to avoid errors due to an out of bound access, \texttt{get} should only be called with an index that does not exceed the list length.
+Secondly, to avoid errors due to an out of bound access, \texttt{get} should only be called with an index that exceeds the list length.
 Lastly, when working with resources such as a file descriptor it might be useful to track it's state and ensure that it is in a proper (closed) state at the end of program execution.
-Note that the same standard library functions (\lstinline|createFD|, \lstinline|readLines|, etc...) may be needed in other programs with different type systems requirements.
 
@@
 
-These properties can be verified in two major ways.
-
-All features from all type systems can be combined in single, monolithic type system, containing all the complexity that all the different type systems entail.
-A very similar issue can also be observed when operating on Liquid Types: In certain cases we want to define predicates on orthogonal properties that could be verified independently from one another; However, current implementation of liquid types do not allow such distinction, leading to confusing and unintelligible error messages~\cite{wits-error-messages}.
+Optimally, these properties should be verified independently from each other wherever possible.
+For example tracking the state of a resource can be done independently of it's specific type, so in case that for whatever reason this check fails we still want to be able to independently verify whether or not the state behaves as we intend it.
 
@@
 
+A very similar issue can also be observed when operating on Liquid Types: In certain cases we want to define predicates on orthogonal properties that could be verified independently from one another; However, current implementation of liquid types do not allow us to define distinct predicates to be verified independently.
+In such cases we have to merge all relevant properties into one joint predicate which can make it hard to identify the specific issue when the verification fails.
+
 \section{Proposed Approach}
@@
 
-We propose a second, more principled alternative: \LayeredTypes.
-In \LayeredTypes, developers can write programs, but can also define additional type systems as layers.
-Each layer defines the basic types, and how typechecking happens.
-Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.
+To tackle these issues, we present \LayeredTypes.
+In our approach we view different type systems as indepent layers that can be verified in an incremental way.
+Layers can either verify properties in a complete stand-alone matter, or they can define dependencies to other layers which allow to build more complex type systems.
 
-Type checking of programs can be partial in the sense that a program may only use some layers (even if the libraries used are defined in more layers), requiring only that each function contains type information for that layer and all dependencies.
-The missing layers can be added over time, if they are found to be relevant.
+We want to note an important distinction to Gradual Typing\cite{gradual}\cite{gradual-objects}: While the latter is concerned with only providing partial typing information and making it gradually more precise, our approach does not focus on the information provided to the verifier, but rather on \textit{how} this information.
 
-We want to note an important distinction to Gradual Typing~\cite{gradual,gradual-objects}: Gradual Typing allows to only provide partial typing information in one system, but does not allow to choose different sets of types to verify.
-It allows to blend between having no types (0) and having types (1).
-Our proposal supports multiple type systems, with different combinations among them.
-
-We evaluated our approach in a prototype language.
+In our work we present a python-based demonstration framework that implements this layered verification approach.
 Users can provide their own implementations for verification layers and define dependencies between them.
@@
 
-\input{chapters/acknowledgements.tex}
-
-\bibliographystyle{plain}
-\bibliography{bibliography}
-
-\end{document}
 %------------------------------------------------------------------------------

### 2023-03-15 tex/chapters/introduction.tex
Restructured LaTeX directories
- Now represents individual submissions/styles

\section{Introduction}
\label{sec:introduction}
\subsection{Context}
\label{ssec:context}
Type systems are a core part of most modern programming languages. Most programmers will be familiar with the concept of simple static type systems as they are present in languages such as C++, Java, Rust or Haskell, that assign types to variables and functions and checks that types are used in a correct way. Using static type systems bears a lot of benefits as it can already prevent a multitude of runtime errors.

However, static type systems alone cannot prevent runtime errors alone as they do not present strong enough guarantees. Therefore many other type systems have been proposed over the years. One example of this are \textit{Refinement Types}, also referred to as \textit{Liquid Types}. These allow us to annotate the traditional type system with additional logical predicates in order to ensure semantic properties, which can be checked statically at compile time. Using these techniques also allows us to define pre- and post-conditions on functions' arguments and return values. By properly defining these contracts a developer can clearly communicate and moreover enforce how other developers' code is allowed to interface with their implementations. Examples for constraints might include any of the following:

\begin{itemize}
	\item Ensuring values are in certain bounds
	\item Ensuring that a list is non-empty
	\item Ensuring that data structures certain to specific semantic properties e.g. that they are sorted
\end{itemize}

All in all, refinement types are valuable building blocks to creating reliable software systems. Using them can minimize the risk of any unhandled run-time errors occuring which might lead to undesired or even undefined behavior.

Refinement Types are but one example of the many different typesystems that have been introduced over the years. In this paper we want to discuss how it is possible to combine different type systems and combine their advantages.

\subsection{Motivation}
\label{ssec:motivation}

While various type systems have been proposed over the years, using even one of them in an existing program is not a trivial task. In order to use these more advanced type systems, programmers often depend on specialized tools such as separate type checkers or compilers. Depending on the specific implementation these tools may or may not be easily integratable into existing development workflows. It gets even more complicated when trying to combine multiple type systems. In order to use multiple type systems, programmers then rely on a multitude of different tools that often cannot be arbitrarily combined. This can lead to a lot of overhead and complexity in the development process.

Our motivation for this paper was to propose an approach, that allows to easily extend and combine existing type systems, with the goal of making them more accessible to programmers. We want to achieve this by providing a framework that allows to easily combine different type systems and ensures a consistent type checking process.

The original motivation for this paper however, stems from a problem often prevalent while using refinement types; While Refinement Types vastly imrpove traditional type systems, current implementations still expose some drawbacks in their practical usage. One of these challenges stems from the fact that during the static analysis, each refinement is treated as a single monolithic requirement that needs to be fulfilled. However in reality, a refinement might combine different properties of a type that may benefit from being treated individually.

One example for such a refinement in LiquidHaskell is shown in \ref{lst:combined_lt}. There, we define a function that extracts the maximum value of a (descendingly) ordered, non-empty list. In this example, this operation is equivalent to simply extracting the \texttt{head} of a list. While we only define one refinement, it actually operates on two different properties of a list. First, it imposes a restriction on it's \texttt{size} by requiring it to be non-empty, while the second restriction imposes an order on the list elements.

\lstinputlisting[firstline=6,caption={Example for a function with a combined Refinement Type in LiquidHaskell},label={lst:combined_lt}]{CombinedLT.hs}

While the refinement concerns two different aspects of the list that are independent from each other, current static analysis implementations of Liquid Types will treat it only as a single combined requirement to be fulfilled. However especially developers could benefit from a system that treats them independently: When writing code that interfaces with already existing codes such a separated approach may give a developer more fine-grained information on which requirements they still need to fulfill in their implementation.

Effectively one could treat such different requirements as separate \textit{Layers} that can be verified independently. Additionally with a proper implementation using such layers may also result in an implementation benefit, that different Layers can be defined individually and combined arbitrarily, as opposed to the example in Listing \ref{lst:combined_lt} where the combination of the two traits had to be explicitly defined as an refinement on it's own.

One other possible usage of such an iterative approach is that it could allow to more flexibly add different static analysis approaches on top of each other. For example \ref{handley2020} have shown how to use liquid types in Haskell to statically perform cost analysis of different resources such as memory consumption or computational complexity. In their approach they showed that liquid types can be used to reason about their upper and lower bounds. However, in their work they only analysed one resource at a time. However, using a layered approach would allow to easily combine these different resource analyses to eg. track and reason about both memory usage and computational complexity.

\subsection{Problem Definition}
\label{ssec:problem_definition}

Current verification approaches usually focus on verifying a single property of a program at a time. However, in practice, one might needs to verify different properties of the program for different purposes. In current implementations, it is not trivially possible to combine these different verification passes. One specific examplefor this can be found when looking at current implemetnations of Liquid Types. In these, one can only define one liquid type for variables, arguments or return types. This forces us to combine all properties we want to statically check and verify into one joint predicate. Using such an approach has the drawback that all properties are checked simultaneously which might make it difficult for a developer to identify due to which property a type error occurs. Additionally previously declared liquid types may not be reused when wanting to refine them further.

With this paper, we therefore propose a new incremental verification approach that we will refer to as \textit{Layered Types}. The goal of this approach is it for a program to be possible to define individual layers for the verification process that are verified individually. By defining dependencies between those layers, it is possible to define a verification order that ensures that all layers are verified in the correct order. This allows us to define different properties of a program that can be verified independently from each other.

With this paper, we want to tackle the following research questions:

\begin{enumerate}
	\item RQ1: Can we define a layered type system that allows to treat different properties of a program independently from each other?
	\item RQ2: Can we use such a layered type system to improve challenges that are present in current Liquid Type implementations?
\end{enumerate}

While in this paper, we will focus on specific Use-Cases that aim towards using it with Liquid Types, we believe that the proposed approach can be applied to other verification techniques as well. For example, we believe that the proposed approach can be extended to verify all sorts of different properties and typesystems such as e.g. \textit{Linear Types} \cite{linear_types} or \textit{Dependent Types} \cite{dependent_types}.

We also want to make an important distinction on the terminology used within the scope of this paper: While on first glance this approach could be described as an approach to \textit{Gradual Verification}, we will explicitly refer to our approach as \textit{Layered} to distinguish it from \textit{Gradual Verification} techniques. While the latter deal with the gradient between incomplete and complete information (in that case on types), our approach is rather concerned not with what information is available on the typing but rather how it is processed. However, we believe that our approach can also be used to implement gradual verification techniques.

\subsection{Impact}
\label{ssec:impact}

The main impact of this paper is to showcase an approach of how different, dependent or independent, type systems can be combined using a single framework without the need to rely on multiple different tools. This allows to combine the advantages of different type systems and allows to use them in a more flexible way. Additionally, we believe that this approach can be used to improve the usability of Liquid Types and other verification techniques. Specifically, we believe our approach has the potential to have the following impacts on Liquid Types in practice:

\paragraph{Improved Error Messages}

One current challenge when using liquid types is that all predicates for a type are checked simultaneously. This can make it difficult to identify which predicate caused a type liquid error. By separating the predicates into separate \textit{Layers} and verifying them individually, it is possible to precisely identify the predicate causing the type error. We expect this to help developers to more easily identify and fix these type errors in their code.

\paragraph{Improved Usability and Reusability}

In current Liquid Type implementations, types that require multiple predicates are defined as a single type. This may make it difficult for a developer to directly identify which properties a specific liquid type ensures. By separating different properties into a more fine-grained layered approach, this may make it easier to identify these properties. Additionally, by separating the predicates into different layers, it is possible to reuse liquid types in a more flexible way. For example, if a developer wants to refine a previously defined liquid type, they can now do so by adding a new layer to it, instead of having to redefine the entire type.

\paragraph{Improved Composability}

In addition to the improved reusability of our approach, we also expect it to improve the composability of different liquid types. With our suggested approach, new refinements can simply be created by composing already existing layers. This vastly increases modularity of the liquid type approach. It also may allow to re-use layers that have already been defined by other developers for ones own implementation.

\paragraph{Improved Modularity}

By separating the predicates into different layers, it is possible to reuse previously defined liquid types in a more flexible way. For example, if a developer wants to refine a previously defined liquid type, they can now do so by adding a new layer to it, instead of having to redefine the entire type. Additionally, this also supports the seperation of concerns when collaborating with other developers. In current implementations, when interfacing with other developers' code, one needs to be aware of all the refinements required by such code in order to interact with it. When different properties are divided into proper layers, this is no longer required.

\subsection{Approach}
\label{ssec:approach}

We will start by giving a brief overview of the general principles of refinement types and liquid types in general in section \ref{sec:background}. Additionally, we will discuss the current state of the art in liquid types and gradual verification in section \ref{sec:related}.

In section \ref{sec:implementation} we will first give a more detailed description of our approach and define a simple functional language that we will use for further demonstration. We will then describe in detail the proposed architecture of our layered type checker and our layered verification framework.

In section \ref{sec:use_cases} we will present two specific use cases. One of which will showcase a way of combining different independent type systems using our layered approach. The other use case will showcase how our approach can be used to verify different properties independently in a liquid type system.

In section \ref{sec:evaluation} we will discuss the limitations of our approach and possible extensions. Additionally, we will evaluate how our approach compares to using a pure, non-layered, verification approach.

\subsection{Contributions}
\label{ssec:contributions}

With this paper we deliver the following contributions:

\begin{enumerate}
	\item We propose a new approach to incremental verification that we will refer to as \textit{Layered Types}.
	\item We present an architecture for a layered verification framework that allows to verify different properties of a program independently from each other.
	\item We present two specific use cases that showcase how our approach can be used to verify different properties independently in a liquid type system.
	\item We deliver a prototype implementation of our layered type checker framework, including some example implementations for layers in said framework. Our implementation is available at \url{https://www.github.com/LuAbelt/LayeredTypes}.
\end{enumerate}

### 2023-03-15 paper/code_examples/lib/Examples/BasicLT.hs

@@ -1,18 +1,0 @@
-module Examples.BasicLT where
-
--- Refinement type that describes a non-empty list
-{-@ type NonEmpty a = {v:[a] | len v > 0 } @-}
-
--- Define a head function that only works on non-emtpy lists
-{-@ myHead :: NonEmpty Int -> Int @-}
-myHead :: [Int] -> Int
-myHead (x:_) = x
-
-{-@ ensureNonEmpty :: [Int] -> NonEmpty Int @-}
-ensureNonEmpty :: [Int] -> [Int]
-ensureNonEmpty [] = [0]
-ensureNonEmpty x = x
-
-i1 :: Int
--- i1 = myHead []                    -- Not allowed
-i1 = myHead (ensureNonEmpty [])   -- Allowed

### 2023-03-15 paper/code_examples/lib/Examples/CombinedLT.hs

@@ -1,11 +1,0 @@
-module Examples.CombinedLT where
-
-{-@ type NonEmpty a = {v:[a] | len v > 0 } @-}
-{-@ type Ordered a = [a]<{\x v -> x <= v}> @-}
--- Not possible to arbitrarly combine NonEmpty and Ordered into one refinement type
-{-@ type NonEmptyOrdered a = {v:[a]<{\x v -> x >= v}> | len v > 0 }@-}
-
--- Extracting the maximum value from a non-empty and ordered list is equivalent to head
-{-@ maxVal :: NonEmptyOrdered Int -> Int @-}
-maxVal :: [Int] -> Int
-maxVal (x:_) = x

### 2023-03-15 tex/code_examples/lib/Examples/BasicLT.hs

@@ -1,0 +1,18 @@
+module Examples.BasicLT where
+
+-- Refinement type that describes a non-empty list
+{-@ type NonEmpty a = {v:[a] | len v > 0 } @-}
+
+-- Define a head function that only works on non-emtpy lists
+{-@ myHead :: NonEmpty Int -> Int @-}
+myHead :: [Int] -> Int
+myHead (x:_) = x
+
+{-@ ensureNonEmpty :: [Int] -> NonEmpty Int @-}
+ensureNonEmpty :: [Int] -> [Int]
+ensureNonEmpty [] = [0]
+ensureNonEmpty x = x
+
+i1 :: Int
+-- i1 = myHead []                    -- Not allowed
+i1 = myHead (ensureNonEmpty [])   -- Allowed

### 2023-03-15 tex/code_examples/lib/Examples/CombinedLT.hs

@@ -1,0 +1,11 @@
+module Examples.CombinedLT where
+
+{-@ type NonEmpty a = {v:[a] | len v > 0 } @-}
+{-@ type Ordered a = [a]<{\x v -> x <= v}> @-}
+-- Not possible to arbitrarly combine NonEmpty and Ordered into one refinement type
+{-@ type NonEmptyOrdered a = {v:[a]<{\x v -> x >= v}> | len v > 0 }@-}
+
+-- Extracting the maximum value from a non-empty and ordered list is equivalent to head
+{-@ maxVal :: NonEmptyOrdered Int -> Int @-}
+maxVal :: [Int] -> Int
+maxVal (x:_) = x

### 2023-03-15 tex/code_examples/lib/Main.hs

@@ -1,0 +1,31 @@
+main :: IO ()
+
+-- Refinement type that describes a non-empty list
+{-@ type NonEmpty a = {v:[a] | len v > 0 } @-}
+
+-- Define a head function that only works on non-emtpy lists
+{-@ myHead :: NonEmpty Int -> Int @-}
+myHead :: [Int] -> Int
+myHead (x:_) = x
+
+{-@ ensureNonEmpty :: [Int] -> NonEmpty Int @-}
+ensureNonEmpty :: [Int] -> [Int]
+ensureNonEmpty [] = [0]
+ensureNonEmpty x = x
+
+i1 :: [Int]
+i1 = [2,5,6,3,2]                    -- Not allowed
+-- i1 = myHead (ensureNonEmpty [])   -- Allowed
+
+{-@ type Ordered a = [a]<{\x v -> x <= v}> @-}
+-- Not possible to arbitrarly combine NonEmpty and Ordered into one refinement type
+{-@ type OrderedList a = {v:[a]<{\x v -> x >= v}> | len v > 0 }@-}
+
+-- Extracting the maximum value from a non-empty and ordered list is equivalent to head
+{-@ maxVal :: OrderedList Int -> Int @-}
+maxVal :: [Int] -> Int
+maxVal (x:_) = x
+
+r = maxVal i1 
+
+main = putStrLn "Hello, Worldi!"

### 2023-03-16 code/LayeredTypes/examples/python_impls/use_case_2.py
* Updated use case 2 to only use pure functions and do not re-assign variables

@@ -1,11 +1,13 @@
+from random import randrange
 from builtins import len, print
 
-def set_element(ls, idx, val):
+def set_list_element(ls, idx, val):
     ls[idx] = val
+    return ls
 
 def get_element(ls, idx):
     return ls[idx]
 
 def get_list():
-    # Return some unordered list with 12 elements
-    return [12, 11, 10, 9, 8, 7, 6, 5, 4, 3, 2, 1]
+    # Return list with 12 random elements
+    return [randrange( 100) for _ in range(12)]

### 2023-03-16 code/LayeredTypes/compiler/Compiler.py
* Refactored layer processing to not use static order

@@ -1,9 +1,11 @@
 import graphlib
 import os.path
+from enum import Enum
 from warnings import warn
 
 import lark
 from pathlib import Path
+from termcolor import colored
 
 from compiler.Exceptions import LayerException
 from compiler.transformers.RemoveTokens import RemoveTokens
@@ -15,9 +17,19 @@ from compiler.interpreters.CheckCF import CheckCF
 from compiler.transformers.CreateAnnotatedTree import CreateAnnotatedTree as AnnotateTree
 from compiler.Interpreters import SimpleInterpreter
 
+class LayerVerificationState(Enum):
+    UNPROCESSED = 0
+    VERIFYING = 1
+    SUCCESS = 2
+    FAILURE = 3
+    BLOCKED = 4
+    CYCLE = 5
+
 class LayeredCompiler:
     def __init__(self, layer_base_dir, implementations_file=None):
 
+        self.layer_states = dict()
+        self.layer_errors = dict()
         self.layer_base_dir = Path(layer_base_dir)
 
         if implementations_file is None:
@@ -40,26 +52,85 @@ class LayeredCompiler:
         self.layers = dict()
         self.interpreter = SimpleInterpreter(self.implementations_file)
 
-    def typecheck(self, input_file, check_cf=True):
+    def typecheck(self, input_file, check_cf=True, verbose=False, raise_on_error=True):
         tree = self.parse(input_file)
 
-        self.__typecheck(tree, check_cf)
+        self.__typecheck(tree, check_cf, raise_on_error)
+
+        if verbose:
+            self.print_layer_states()
 
         return True
 
-    def __typecheck(self, tree, check_cf=True):
-        lv = CollectLayers()
-        cf_check = CheckCF(self.interpreter.external_functions_names)
-
-        tree = lv.transform(tree)
+    def __typecheck(self, tree, check_cf=True, raise_on_error=True):
         tree = AnnotateTree().transform(tree)
 
         if check_cf:
+            cf_check = CheckCF(self.interpreter.external_functions_names)
             cf_check.visit(tree)
+
+        layer_graph = self.build_layer_graph(tree)
+
+        self.layer_states = {layer_id: LayerVerificationState.UNPROCESSED for layer_id in layer_graph}
+        # Store the processing states for all layers
+        for layer_id in layer_graph:
+            self.layer_states.update({layer_id: LayerVerificationState.UNPROCESSED for layer_id in layer_graph[layer_id]})
+
+        # We use graphlib to process layers in topological order
[…]

### 2023-03-16 code/LayeredTypes/compiler/Compiler.py
* Fixed layer processing order to not run out of nodes too early

@@ -93,7 +93,9 @@ class LayeredCompiler:
                 self.layer_states[layer_id] = LayerVerificationState.CYCLE
 
         while topological_sorter.is_active():
+            processed_one = False
             for node in topological_sorter.get_ready():
+                processed_one = True
                 assert self.layer_states[node] == LayerVerificationState.UNPROCESSED
                 layer_handle = self.layers[node]
                 try:
@@ -109,7 +111,7 @@ class LayeredCompiler:
             # We do not want to process nodes that depend on a failed node
             # This is done by marking the node as done, so it is not returned by get_ready()
             # We do this until there are no more nodes to process
-            if topological_sorter.is_active() and len(topological_sorter.get_ready()) == 0:
+            if not processed_one:
                 break
 
         # There might be unparsed layers, which are blocked due to failed dependencies

### 2023-03-16 code/LayeredTypes/tests/test_layers.py

@@ -83,7 +83,7 @@ class TestLayerDependencies(unittest.TestCase):
         self.assertTrue("A" in compiler.layers)
         self.assertTrue("B" in compiler.layers)
 
-        self.assertEqual(call_order, ["A", "B"])
+        self.assertEqual(["A", "B"], call_order)
 
     def test_implicit_cycle(self):
         compiler = get_compiler()

### 2023-03-16 code/LayeredTypes/compiler/Compiler.py
* Added additional test to correctly check that layers are verified as far as possible

@@ -17,13 +17,15 @@ from compiler.interpreters.CheckCF import CheckCF
 from compiler.transformers.CreateAnnotatedTree import CreateAnnotatedTree as AnnotateTree
 from compiler.Interpreters import SimpleInterpreter
 
+
 class LayerVerificationState(Enum):
-    UNPROCESSED = 0
-    VERIFYING = 1
-    SUCCESS = 2
-    FAILURE = 3
-    BLOCKED = 4
-    CYCLE = 5
+    UNPROCESSED = 0 # Layer has not been processed yet
+    VERIFYING = 1   # Unused at the moment. Used for parallel verification
+    SUCCESS = 2     # Layer has been successfully verified
+    FAILURE = 3     # Layer has failed verification
+    BLOCKED = 4     # Layer has been blocked by a layer that failed verification
+    CYCLE = 5       # Layer is part of a cycle
+    CYCLE_BLOCKED = 6   # Layer is blocked by a layer that is part of a cycle
 
 class LayeredCompiler:
     def __init__(self, layer_base_dir, implementations_file=None):
@@ -89,7 +91,7 @@ class LayeredCompiler:
 
             has_cycle = True
 
-            for layer_id in e.args[0]:
+            for layer_id in e.args[1]:
                 self.layer_states[layer_id] = LayerVerificationState.CYCLE
 
         while topological_sorter.is_active():
@@ -123,6 +125,11 @@ class LayeredCompiler:
                     for node in topological_sorter.get_ready():
                         self.layer_states[node] = LayerVerificationState.BLOCKED
                         topological_sorter.done(node)
+
+        # All layers that are still unprocessed are blocked because they depend on a layer that is part of a cycle
+        for layer_id in self.layer_states:
+            if self.layer_states[layer_id] == LayerVerificationState.UNPROCESSED:
+                self.layer_states[layer_id] = LayerVerificationState.CYCLE_BLOCKED
 
         # Return true iff all layers are successfully verified
         return all([self.layer_states[layer_id] == LayerVerificationState.SUCCESS for layer_id in self.layer_states])

### 2023-03-16 code/LayeredTypes/tests/layer_implementations/blockedLayerC.py

@@ -1,0 +1,8 @@
+def depends_on():
+    return {"failLayerA"}
+
+def typecheck(tree):
+    return tree
+
+def parse_type(type_str):
+    return None

### 2023-03-16 code/LayeredTypes/tests/layer_implementations/dependOnCycleA.py

@@ -1,0 +1,8 @@
+def depends_on():
+    return {"cycleA"}
+
+def typecheck(tree):
+    return tree
+
+def parse_type(type_str):
+    return None

### 2023-03-16 code/LayeredTypes/tests/layer_implementations/failLayerA.py

@@ -1,0 +1,8 @@
+def depends_on():
+    return set()
+
+def typecheck(tree):
+    raise Exception("Layer A failed")
+
+def parse_type(type_str):
+    return None

### 2023-03-16 code/LayeredTypes/tests/test_layers.py

@@ -1,6 +1,7 @@
 import unittest
 from graphlib import CycleError
 
+from compiler.Compiler import LayerVerificationState
 from utils import get_compiler, full_path, call_order
 
 class TestLayerDefinitions(unittest.TestCase):
@@ -70,6 +71,15 @@ class TestLayerDependencies(unittest.TestCase):
         with self.assertRaises(CycleError):
             compiler.typecheck(src_file)
 
+    def test_circular_dependency_2(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/create_cycle.fl")
+
+        compiler.typecheck(src_file, raise_on_error=False)
+
+        self.assertEqual(compiler.layer_states["cycleA"], LayerVerificationState.CYCLE)
+        self.assertEqual(compiler.layer_states["cycleB"], LayerVerificationState.CYCLE)
+
     def test_implicit_dependency(self):
         compiler = get_compiler()
         src_file = full_path("/test_code/layers/implicit_dependency.fl")
@@ -92,6 +102,15 @@ class TestLayerDependencies(unittest.TestCase):
         with self.assertRaises(CycleError) as e:
             compiler.typecheck(src_file)
 
+    def test_implicit_cycle_2(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/implicit_cycle.fl")
+
+        compiler.typecheck(src_file, raise_on_error=False)
+
+        self.assertEqual(compiler.layer_states["cycleA"], LayerVerificationState.CYCLE)
+        self.assertEqual(compiler.layer_states["cycleB"], LayerVerificationState.CYCLE)
+
     def test_implicit_multiple_layers(self):
         compiler = get_compiler()
         src_file = full_path("/test_code/layers/implicit_multiple_layers.fl")
@@ -108,4 +127,23 @@ class TestLayerDependencies(unittest.TestCase):
         self.assertTrue("D" in compiler.layers)
 
         self.assertIn(call_order, [["A", "B", "C", "D"], ["A", "C", "B", "D"]])
+
+    def test_complex_layers(self):
+        compiler = get_compiler()
+        src_file = full_path("/test_code/layers/complex_layers.fl")
+
+        call_order.clear()
+
+        compiler.typecheck(src_file, raise_on_error=False)
+
+        self.assertEqual(LayerVerificationState.SUCCESS, compiler.layer_states["A"])
[…]

### 2023-03-16 code/LayeredTypes/compiler/Compiler.py
* Fixed main to print verification summary by default

@@ -204,7 +204,8 @@ class LayeredCompiler:
         return self.interpreter.run(tree)
 
     def print_layer_states(self):
-        for layer_id, state in self.layer_states:
+        for layer_id in self.layer_states:
+            state = self.layer_states[layer_id]
             # Print layer name and state
             # Color of state depends on state:
             #   - Success: Green
@@ -217,7 +218,8 @@ class LayeredCompiler:
                 LayerVerificationState.FAILURE: "red",
                 LayerVerificationState.BLOCKED: "yellow",
                 LayerVerificationState.UNPROCESSED: "white",
-                LayerVerificationState.CYCLE: "magenta"
+                LayerVerificationState.CYCLE: "magenta",
+                LayerVerificationState.CYCLE_BLOCKED: "cyan"
             }
 
 

### 2023-03-16 code/LayeredTypes/main.py

@@ -31,7 +31,7 @@ if __name__ == "__main__":
 
     try:
         if options.typecheck:
-            compiler.typecheck(file)
+            compiler.typecheck(file, verbose=True, raise_on_error=False)
         else:
             result = compiler.run(file)
             print(f"Program returned: {result}")

### 2023-03-20 tex/report_fct/report.tex
Added FCT report

--- 2023-03-10 paper/types2023.tex
@@
 %
-\authorrunning{Abelt and Fonseca}
+\authorrunning{Abelt}
 
@@
 %
-\title{LayeredTypes -- Combining dependent and independent type systems}
+\title{CAMELOT Research Scholarship -- Final Report}
 
@@
 Lukas Abelt\inst{1,2}
-\and
-Alcides Fonseca\inst{1}
 }
@@
 % will be rejected by EasyChair
-\titlerunning{Layered Types}
+\titlerunning{Project Report CAMELOT}
 
@@
 %------------------------------------------------------------------------------
-\section{Context}
+\section{Introduction}
+
+This document is the final project report for the research scholarship granted in the scope of the CAMELOT project for the duration of 01.09.2022-30.11.2022.
+It was carried out by Lukas Abelt in the LASIGE Research Group within the Reliable Software Systems Research Line.
+The work was supercised by Alcides Fonseca, Assistant Professor at the Faculdade de Ciências da Universidade de Lisboa.
+The work was part of an Erasmus internship that ends in March 2023.
+
+\section{Description of the Scholarship}
+\subsection{Context}
 \label{sec:context}
@@
 
- % Alcides is here
-
  To illustrate this in a small example, consider the program in \Cref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line.
@@
 Note that the same standard library functions (\lstinline|createFD|, \lstinline|readLines|, etc...) may be needed in other programs with different type systems requirements.
-
-%\begin{itemize}
-%	\item From a type-safety perspective, one wants to make sure that e.g. \texttt{get(...)} is only called with a list and an integer index as arguments
-%	\item Additionally, one wants to ensure that \texttt{get} is never accessed with an index that exceeds the lists length to avoid out-of-bounds access errors at runtime
-%	\item When working with resources such as file descriptors one wants to ensure that these are in a proper (closed) state by the end of the program execution
-%\end{itemize}
 
@@
 
-\section{Proposed Approach}
+\subsection{Goals}
 \label{sec:proposed-approach}
 
-We propose a second, more principled alternative: \LayeredTypes.
-In \LayeredTypes, developers can write programs, but can also define additional type systems as layers.
+The Goals of the Scholarship was to explore a new verification approach, called \LayeredTypes.
+The Goal was, that in \LayeredTypes, developers can write programs, but can also define additional type systems as layers.
 Each layer defines the basic types, and how typechecking happens.
@@
 
-We want to note an important distinction to Gradual Typing~\cite{gradual,gradual-objects}: Gradual Typing allows to only provide partial typing information in one system, but does not allow to choose different sets of types to verify.
-It allows to blend between having no types (0) and having types (1).
-Our proposal supports multiple type systems, with different combinations among them.
+For this purpose, a prototype was to be developed consisting of a small and simple functional language accompanied by a prototype layered Verification tool that can realise the described layered verification approach.
 
-We evaluated our approach in a prototype language.
-Users can provide their own implementations for verification layers and define dependencies between them.
-In \autoref{lst:code_after} we see a set of annotations that can be added to the base code of \autoref{lst:code_before} to tackle the issues described.
+We believe that this layered, incremental approach can help build more powerful and independent type systems while at the same time making it simpler to understand the errors that might arise during verification.
+
+\subsection{Results obtained}
+During the duration of the scholarship a prototype compiler and verifier was built in a prototype language.
+This work layed an important foundation for the subsequent work for the remaining time of the ERASMUS+ Internship in which the implementation was further extended and improved.
+Two exemplary use-cases were developed showing the potential of the layered approach.
+
+One example of how the framework works can be seen in the accompanging code fragments.
+In \Cref{lst:code_after} we see a set of annotations that can be added to the base code of \Cref{lst:code_before}.
 We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}.
@@
 
-We believe that this layered, incremental approach can help build more powerful and independent type systems while at the same time making it simpler to understand the errors that might arise during verification.
+\subsection{Deliverables and Publications}
 
-\input{chapters/acknowledgements.tex}
+The full implementation of the developed prototype will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}.
+In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection.
+Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.
+
+\input{../chapters/acknowledgements.tex}
 
 \bibliographystyle{plain}
-\bibliography{bibliography}
+\bibliography{literature.bib}
 
 \end{document}
-%------------------------------------------------------------------------------
-\bibliographystyle{plain}
-\bibliography{literature}
 
 %------------------------------------------------------------------------------
-\end{document}

### 2023-03-21 tex/report_uds/report.tex
Added base for uds internship report

--- 2023-03-20 tex/report_fct/report.tex
@@
 \usepackage{doc}
-
+\usepackage{tabularx}
 % use this if you have a long article and want to create an index
@@
 }
+
 \lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
@@
 %
-\authorrunning{Abelt}
-
-\newcommand{\easychair}{\textsf{easychair}}
-\newcommand{\miktex}{MiK{\TeX}}
-\newcommand{\texniccenter}{{\TeX}nicCenter}
-\newcommand{\makefile}{\texttt{Makefile}}
-\newcommand{\latexeditor}{LEd}
+\authorrunning{Lukas Abelt}
 
 \newcommand{\LayeredTypes}{\textsc{LayeredTypes}}
+\newcommand{\Lark}{\textsc{Lark}}
 %\makeindex
+
+\makeatletter
+\def\input@path{{../chapters}}
+%or: \def\input@path{{/path/to/folder/}{/path/to/other/folder/}}
+\makeatother
+\graphicspath{{../graph}}
 
@@
 %
-\title{CAMELOT Research Scholarship -- Final Report}
+\title{Industry Internship -- Final Report}
 
@@
 \institute{
+   Saarland University,
+   Germany\\
+   \email{luab00001@stud.uni-saarland.de}
+  \and
    LASIGE,
@@
   \email{labelt@lasige.di.fc.ul.pt},
-  \email{alcides@ciencias.ulisboa.pt}
-\and
-   Saarland University,
-   Germany\\
 }
 
-%  \authorrunning{} has to be set for the shorter version of the authors' names;
-% otherwise a warning will be rendered in the running heads. When processed by
-% EasyChair, this command is mandatory: a document without \authorrunning
-% will be rejected by EasyChair
-
-% \titlerunning{} has to be set to either the main title or its shorter
-% version for the running heads. When processed by
-% EasyChair, this command is mandatory: a document without \titlerunning
-% will be rejected by EasyChair
-\titlerunning{Project Report CAMELOT}
+\titlerunning{Industrial Internship Report}
 
@@
 
-% The table of contents below is added for your convenience. Please do not use
-% the table of contents if you are preparing your paper for publication in the
-% EPiC Series or Kalpa Publications series
+%------------------------------------------------------------------------------
+\section{About this document}
 
-%\setcounter{tocdepth}{2}
-%{\small
-%\tableofcontents}
+This document composes the internship report for the industrial internship of Lukas Abelt, carried out from October 2022 until March 2023 at the LASIGE Research Centre in Lisbon, Portugal.
 
-%\section{To mention}
-%
-%Processing in EasyChair - number of pages.
-%
-%Examples of how EasyChair processes papers. Caveats (replacement of EC
-%class, errors).
+\section{Internship Overview}
+This seciton gives an overview of the relevant key data of persons and institutions involved in this internship.
 
-%------------------------------------------------------------------------------
-\section{Introduction}
+\begin{tabularx}{\textwidth}{rX}
+	\textbf{Name of Intern/Student:} & Lukas Abelt\\
+	\textbf{Student Contact Information:} & \email{luab00001@stud.uni-saarland.de}\\
+	\textbf{Matriculation Number at UdS:} & 7009048\\
+	\textbf{Course of Study at UdS:} & Master of Science Computer Science\\
+	\textbf{Start of Internship:} & 01.01.2022\\
+	\textbf{End of Internship:} & 31.03.2023\\
+	\textbf{Hosting Insitution:} & LASIGE Research Unit\footnote{\url{https://www.lasige.pt/}}\newline Reliable Software Systems Research Line\newline Faculdade de Ciências da Universidade de Lisboa\\
+	\textbf{Internship Supervisor:} & Alcides Fonseca, Assistant Professor\\
+	\textbf{Supervisor Contact Information:} & \email{amfonseca@fc.ul.pt}\\
+	\textbf{Internship Title:} & \textit{Type Layers for Gradual Verification}
+\end{tabularx}
 
-This document is the final project report for the research scholarship granted in the scope of the CAMELOT project for the duration of 01.09.2022-30.11.2022.
-It was carried out by Lukas Abelt in the LASIGE Research Group within the Reliable Software Systems Research Line.
-The work was supercised by Alcides Fonseca, Assistant Professor at the Faculdade de Ciências da Universidade de Lisboa.
-The work was part of an Erasmus internship that ends in March 2023.
+\section{Internship Description}
+This section will give a bread overview of the work that was carried out during the internship.
+It wstart by providing context to the problem at hand and continue further with the proposed approach and goals of the internship.
+A technical description of the implemented solution will be provided and finally a short overview of the acquired skills and produced deliverables will be provided.
 
-\section{Description of the Scholarship}
 \subsection{Context}
@@
 
-In this talk, we try to address the challenge of how can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.
+During this internship, we tried to address the challenge of how can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.
 
@@
 
+\subsection{Goals}
+\label{sec:proposed-approach}
+The Goals of the internship was to explore a new verification approach, which was later decided to be called \LayeredTypes.
+In \LayeredTypes, developers can write programs, but can also define additional type systems as layers.
+Each layer defines the basic types, and how typechecking happens.
+Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.
+
+Type checking of programs can be partial in the sense that a program may only use some layers (even if the libraries used are defined in more layers), requiring only that each function contains type information for that layer and all dependencies.
+The missing layers can be added over time, if they are found to be relevant.
+
+For this purpose, a prototype was to be developed consisting of a small and simple functional language accompanied by a prototype layered Verification tool that can realise the described layered verification approach.
+
+We believe that this layered, incremental approach can help build more powerful and independent type systems while at the same time making it simpler to understand the errors that might arise during verification.
+
+\subsection{Implementation Details}
+
+We developed a prototype for a layered verification approach called \LayeredTypes.
+In this prototype, programs can be written in a simple functional language that supports some primitive operations such as variables, function definitions, basic binary operations and \texttt{let} constructs.
+To enable the use of more complex functionality, it is also possible to define additional functions in python that can be called in this simple language.
+An example of a program written in that language can be seen in \Cref{lst:code_before}.
+
+The prototype consists of a simple compiler and interpreter for said functional language.
+The prototype has been developed in Python using the \Lark compiler framework.
+The core of our compiler is a \textit{layered static verifier}.
+Within a program, the programer can define annotations to be passed on to one or multiple verification layers.
+The implementation of each verification layer is provided by the programer in a separate python file with a pre-defined set of functions that will be called by the verifier.
+Specifically, each layer can provide the following information:
+
+\begin{itemize}
+	\item A list of all other layers that it depends on -- That is, all layers that need to be verified successfully before the layer itself is verified
+	\item A list of all layers that should be run \textit{after} this layer has been successfully verified
+	\item A \texttt{typecheck} function that gets passed the AST of the program and can perform the specific verification for the current layer.
+\end{itemize}
+
+The framework will build the dependency graph of all layers defined and used in a program and run them in a topological order so that all dependencies are satisfied.
+A conceptual overview of the verification framework can be seen in \Cref{fig:framework_overview}
+
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=.8\textwidth]{framework_overview}
+	\label{fig:framework_overview}
+	\caption{Conceptual Overview of \LayeredTypes}
+\end{figure}
+
+To evaluate our work, we also implemented two different use-cases:
+
+In the first Use-Case, we developed a simple program that reads input from a file and processes it line by line.
+In this use-case we use three different layers.
+The first layer, \texttt{types} is a simple layer that is used to annotate typing information to variables and functions.
+The second layer, \texttt{typecheck}, depends on the first one and defines rules on how the program should be type-checked.
+It then uses the annotations provided by the \texttt{type} layer to check if the program is well-typed.
+The third layer, \texttt{state}, can be used to define, check and transistion states on variables.
+In our example we use it to track the state of the file descriptor and ensure that it is closed at the end of the program.
+The \texttt{state} layer acts independently from the other two layers.
+
+In the second Use-Case we focus on a problem currently present in implementations for \textit{Liquid Types}:
+
+\subsection{Deliverables and Publications}
+
+The full implementation of the developed prototype of \LayeredTypes will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}.
+In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection.
+Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.
+
+\input{acknowledgements}
+
+\bibliographystyle{plain}
+\bibliography{literature.bib}
 \begin{minipage}{0.4\linewidth}
@@
 
-\subsection{Goals}
-\label{sec:proposed-approach}
-
-The Goals of the Scholarship was to explore a new verification approach, called \LayeredTypes.
-The Goal was, that in \LayeredTypes, developers can write programs, but can also define additional type systems as layers.
-Each layer defines the basic types, and how typechecking happens.
-Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.
-
-Type checking of programs can be partial in the sense that a program may only use some layers (even if the libraries used are defined in more layers), requiring only that each function contains type information for that layer and all dependencies.
-The missing layers can be added over time, if they are found to be relevant.
-
-For this purpose, a prototype was to be developed consisting of a small and simple functional language accompanied by a prototype layered Verification tool that can realise the described layered verification approach.
-
-We believe that this layered, incremental approach can help build more powerful and independent type systems while at the same time making it simpler to understand the errors that might arise during verification.
-
-\subsection{Results obtained}
-During the duration of the scholarship a prototype compiler and verifier was built in a prototype language.
-This work layed an important foundation for the subsequent work for the remaining time of the ERASMUS+ Internship in which the implementation was further extended and improved.
-Two exemplary use-cases were developed showing the potential of the layered approach.
-
-One example of how the framework works can be seen in the accompanging code fragments.
-In \Cref{lst:code_after} we see a set of annotations that can be added to the base code of \Cref{lst:code_before}.
-We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}.
-Internally, the framework will build a dependency graph thus allowing to verify properties independently from one another where appropriate.
-
-\subsection{Deliverables and Publications}
-
-The full implementation of the developed prototype will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}.
-In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection.
-Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.
-
-\input{../chapters/acknowledgements.tex}
-
-\bibliographystyle{plain}
-\bibliography{literature.bib}
-
 \end{document}

### 2023-03-21 code/LayeredTypes/tests/implementations.py
* Added example for uninterpreted functions

@@ -24,3 +24,13 @@ def f():
 
 def noArgs():
     return 1
+
+def initList():
+    return []
+
+def append(ls, val):
+    ls.append(val)
+    return ls
+
+def create_list(n: int):
+    return list(range(n))

### 2023-03-22 code/LayeredTypes/tests/test_liquid.py
* Updated unit tests for uninterpreted function

@@ -249,6 +249,9 @@ class TestLiquidLayer(unittest.TestCase):
     def test_bin_op_cmp(self):
         typecheck_correct_file(self, "/test_code/liquid/bin_op_cmp.fl")
 
+    def test_uninterpreted_fun(self):
+        typecheck_correct_file(self, "/test_code/liquid/uninterpreted_fun.fl")
+
 refinement_replacement_test_vals = [
     ([parse_type("{v:Int | v > 0}")],
      [RefinedType("$arg0", t_int,

### 2023-03-22 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Generalized liquid layer with layer identifier

@@ -103,10 +103,11 @@ def rename_variable(context, original_name, fun_identifier, all_arg_names, remai
 
 
 class LiquidLayer(lark.visitors.Interpreter):
-    def __init__(self):
+    def __init__(self, layer_identifier: str= "liquid"):
         self.__types = {}
         self.__fun_types = {}
         self.__ctx = EmptyContext()
+        self.__layer_identifier = layer_identifier
         pass
 
     def visit(self, tree):
@@ -121,12 +122,12 @@ class LiquidLayer(lark.visitors.Interpreter):
     def __annotate(self, tree):
 
         for identifier in self.__types:
-            tree.add_layer_annotation("liquid", identifier, "type", self.__types[identifier])
+            tree.add_layer_annotation(self.__layer_identifier, identifier, "type", self.__types[identifier])
 
         for fun_identifier in self.__fun_types:
-            tree.add_layer_annotation("liquid", fun_identifier, "function_type", self.__fun_types[fun_identifier])
+            tree.add_layer_annotation(self.__layer_identifier, fun_identifier, "function_type", self.__fun_types[fun_identifier])
 
-        tree.add_layer_annotation("liquid", "contexts", "context", self.__ctx)
+        tree.add_layer_annotation(self.__layer_identifier, "contexts", "context", self.__ctx)
 
 
     def assign(self, tree: AnnotatedTree):
@@ -136,13 +137,13 @@ class LiquidLayer(lark.visitors.Interpreter):
         identifier = tree.children[0].value
 
         # Try to get the type from context if it has already been assigned
-        ctx = tree.get_layer_annotation("liquid", "contexts", "context")
+        ctx = tree.get_layer_annotation(self.__layer_identifier, "contexts", "context")
 
         id_type = ctx.type_of(identifier)
 
         if id_type is None:
             # Type has not been assigned yet, try to get it from the types dictionary
-            id_type = tree.get_layer_annotation("liquid", identifier, "type")
+            id_type = tree.get_layer_annotation(self.__layer_identifier, identifier, "type")
 
         if id_type is None:
             # Type is still undefined, raise an error
@@ -154,7 +155,7 @@ class LiquidLayer(lark.visitors.Interpreter):
         identifier = tree.children[0].children[0].value
         layer_id = tree.children[1].children[0].value
 
-        if layer_id != "liquid":
+        if layer_id != self.__layer_identifier:
             return
 
[…]

### 2023-03-24 tex/report_uds/report.tex
* Generalized liquid layer with layer identifier
Added missing graph

@@
 %
-\title{Industry Internship -- Final Report}
+\title{Type Layers for Gradual Verification -- Industrial Internship Report}
 
@@
 
-\titlerunning{Industrial Internship Report}
+\titlerunning{Type Layers for Gradual Verification}
 
@@
 	\textbf{Hosting Insitution:} & LASIGE Research Unit\footnote{\url{https://www.lasige.pt/}}\newline Reliable Software Systems Research Line\newline Faculdade de Ciências da Universidade de Lisboa\\
+	\textbf{Working hours per week} & 35 \\
 	\textbf{Internship Supervisor:} & Alcides Fonseca, Assistant Professor\\
 	\textbf{Supervisor Contact Information:} & \email{amfonseca@fc.ul.pt}\\
-	\textbf{Internship Title:} & \textit{Type Layers for Gradual Verification}
+	\textbf{Internship Title:} & \textit{Type Layers for Gradual Verification}\\ 
 \end{tabularx}
+\section{Signatures}
 
+I hereby acknowledge that the information provided in this report is accurate and sufficiently describes the work carried out.
+\vfill
+\par\noindent\makebox[2.5in]{\hrulefill} \hfill\makebox[2.5in]{\hrulefill}%
+\par\noindent\makebox[2.5in][l]{Lukas Abelt, Intern}      \hfill\makebox[2.5in][l]{Alcides Fonseca, Supervisor}%
+\par\noindent\makebox[2.5in][l]{Lisbon, \today}      \hfill\makebox[2.5in][l]{Lisbon, \today}%
+\newpage
 \section{Internship Description}
@@
 
-The prototype consists of a simple compiler and interpreter for said functional language.
-The prototype has been developed in Python using the \Lark compiler framework.
-The core of our compiler is a \textit{layered static verifier}.
-Within a program, the programer can define annotations to be passed on to one or multiple verification layers.
-The implementation of each verification layer is provided by the programer in a separate python file with a pre-defined set of functions that will be called by the verifier.
-Specifically, each layer can provide the following information:
-
-\begin{itemize}
-	\item A list of all other layers that it depends on -- That is, all layers that need to be verified successfully before the layer itself is verified
-	\item A list of all layers that should be run \textit{after} this layer has been successfully verified
-	\item A \texttt{typecheck} function that gets passed the AST of the program and can perform the specific verification for the current layer.
-\end{itemize}
-
-The framework will build the dependency graph of all layers defined and used in a program and run them in a topological order so that all dependencies are satisfied.
-A conceptual overview of the verification framework can be seen in \Cref{fig:framework_overview}
-
-\begin{figure}[ht!]
-	\centering
-	\includegraphics[width=.8\textwidth]{framework_overview}
-	\label{fig:framework_overview}
-	\caption{Conceptual Overview of \LayeredTypes}
-\end{figure}
-
-To evaluate our work, we also implemented two different use-cases:
-
-In the first Use-Case, we developed a simple program that reads input from a file and processes it line by line.
-In this use-case we use three different layers.
-The first layer, \texttt{types} is a simple layer that is used to annotate typing information to variables and functions.
-The second layer, \texttt{typecheck}, depends on the first one and defines rules on how the program should be type-checked.
-It then uses the annotations provided by the \texttt{type} layer to check if the program is well-typed.
-The third layer, \texttt{state}, can be used to define, check and transistion states on variables.
-In our example we use it to track the state of the file descriptor and ensure that it is closed at the end of the program.
-The \texttt{state} layer acts independently from the other two layers.
-
-In the second Use-Case we focus on a problem currently present in implementations for \textit{Liquid Types}:
-
-\subsection{Deliverables and Publications}
-
-The full implementation of the developed prototype of \LayeredTypes will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}.
-In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection.
-Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.
-
-\input{acknowledgements}
-
-\bibliographystyle{plain}
-\bibliography{literature.bib}
 \begin{minipage}{0.4\linewidth}
@@
 
+The prototype consists of a simple compiler and interpreter for said functional language.
+The prototype has been developed in Python using the \Lark compiler framework.
+The core of our compiler is a \textit{layered static verifier}.
+Within a program, the programer can define annotations to be passed on to one or multiple verification layers.
+An example of such annotations is shown in \Cref{lst:code_after} The implementation of each verification layer is provided by the programer in a separate python file with a pre-defined set of functions that will be called by the verifier.
+Specifically, each layer can provide the following information:
+
+\begin{itemize}
+	\item A list of all other layers that it depends on -- That is, all layers that need to be verified successfully before the layer itself is verified
+	\item A list of all layers that should be run \textit{after} this layer has been successfully verified
+	\item A \texttt{typecheck} function that gets passed the AST of the program and can perform the specific verification for the current layer.
+\end{itemize}
+
+The framework will build the dependency graph of all layers defined and used in a program and run them in a topological order so that all dependencies are satisfied.
+A conceptual overview of the verification framework can be seen in \Cref{fig:framework_overview}
+
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=.8\textwidth]{framework_overview}
+	\caption{Conceptual Overview of \LayeredTypes}
+	\label{fig:framework_overview}
+\end{figure}
+
+To evaluate our work, we also implemented two different use-cases:
+
+In the first Use-Case, we developed a simple program that reads input from a file and processes it line by line.
+In this use-case we use three different layers.
+The first layer, \texttt{types} is a simple layer that is used to annotate typing information to variables and functions.
+The second layer, \texttt{typecheck}, depends on the first one and defines rules on how the program should be type-checked.
+It then uses the annotations provided by the \texttt{type} layer to check if the program is well-typed.
+The third layer, \texttt{state}, can be used to define, check and transistion states on variables.
+In our example we use it to track the state of the file descriptor and ensure that it is closed at the end of the program.
+The \texttt{state} layer acts independently from the other two layers.
+
+In the second Use-Case we focus on a problem currently present in implementations for \textit{Liquid Types}.
+When operating on liquid types, often we want to reason about orthogonal properties of variables and functions.
+One example could be the verification of machine learning dataset pipelines: Some operations may only depend and influence the rows (Datapoints) of a dataset (e.g.
+adding a datapoint does not affect the number of columns), while others only only affect the columns (Features).
+Finally, some actions require information on both rows and columns.
+For example when training a model based on a dataset, we may want to ensure that this dataset contains enough datapoints for the amount of features.
+
+In our specific running example we added a few simple functions with that specific behavior.
+We have several methods (Adding, removing, extracting a certain amount of elements) that only affect rows and columns separately but also a \texttt{train} method, that introduces a relationship between the number of columns and the number of rows: We assume that we need at least ten times as many datapoints as we have features.
+
+In our example we showcased the issues that arise with hard to interpret error messages that occur when typechecking fails when verifying all properties at once.
+Subsequently we separated the individual properties and interactions into separate layers to demonstrate how this separation affects the error messages produced.
+
+\subsection{Deliverables, Publications and Skills acquired}
+
+During the work on this internship and the subsequent implementation, the intern acquired theoretical as well as practical knowledge on the following aspects:
+\begin{itemize}
+	\item Compiler Construction
+	\item Verification Techniques
+	\item Importance and Distinction of different Type Systems including, but not limited to:
+		\begin{itemize}
+			\item Linear Types
+			\item Dependent Types
+			\item Liquid Types
+		\end{itemize}
+\end{itemize}
+
+The full implementation of the developed prototype of \LayeredTypes will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}.
+In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection.
+Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.
+
+\input{acknowledgements}
+
+\bibliographystyle{plain}
+\bibliography{literature.bib}
+
 \end{document}

### 2023-03-24 code/LayeredTypes/layer_implementations/rowcols.py
* Started finalization of second use-case

@@ -1,0 +1,9 @@
+from compiler.interpreters import LiquidTypeChecker
+
+def depends_on():
+    return {"rows","cols"}
+def typecheck(tree):
+    liquid_checker = LiquidTypeChecker.LiquidLayer(layer_identifier="rowcols")
+    liquid_checker.visit(tree)
+
+    return tree

### 2023-03-24 code/LayeredTypes/tests/test_liquid.py

@@ -252,6 +252,9 @@ class TestLiquidLayer(unittest.TestCase):
     def test_uninterpreted_fun(self):
         typecheck_correct_file(self, "/test_code/liquid/uninterpreted_fun.fl")
 
+    def test_additional_context(self):
+        typecheck_correct_file(self, "/test_code/liquid/additional_context.fl")
+
 refinement_replacement_test_vals = [
     ([parse_type("{v:Int | v > 0}")],
      [RefinedType("$arg0", t_int,

### 2023-03-24 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Added proper dependencies between liquid layers

@@ -12,6 +12,7 @@ from aeon.frontend.parser import mk_parser
 from aeon.typing.context import EmptyContext, VariableBinder, TypingContext
 from aeon.typing.entailment import entailment
 from aeon.verification.sub import sub
+from aeon.verification.vcs import Conjunction
 from compiler.Exceptions import TypecheckException, WrongArgumentCountException, FeatureNotSupportedError
 from compiler.transformers.CreateAnnotatedTree import AnnotatedTree, make_annotated_tree
 
@@ -101,25 +102,96 @@ def rename_variable(context, original_name, fun_identifier, all_arg_names, remai
             remaining_args[i] = substitution_in_type(remaining_args[i], Var(free_var_name), original_name)
     return free_var_name, remaining_args
 
+def rename_in_context(context: TypingContext, original_name, new_name):
+    if not isinstance(context, VariableBinder):
+        return
+
+    if context.name == original_name:
+        context.name = new_name
+
+
+    assert isinstance(context.type, RefinedType)
+    context.type.refinement = substitution_in_liquid(context.type.refinement, LiquidVar(new_name), original_name)
+
+    rename_in_context(context.prev, original_name, new_name)
+def combine_contexts(original_context: TypingContext, additional_context: TypingContext, refined_type: RefinedType):
+    copy_context = deepcopy(additional_context)
+
+    while not isinstance(copy_context, EmptyContext):
+        assert isinstance(copy_context, VariableBinder)
+
+        original_name = copy_context.name
+        fresh_name = original_name
+
+        while original_context.type_of(fresh_name) or additional_context.type_of(fresh_name):
+
+            if original_context.type_of(fresh_name):
+                fresh_name = make_name_unique(fresh_name, original_context)
+
+            if additional_context.type_of(fresh_name):
+                fresh_name = make_name_unique(fresh_name, additional_context)
+
+        if original_name in liquid_free_vars(refined_type.refinement):
+            refined_type.refinement = substitution_in_liquid(refined_type.refinement, LiquidVar(original_name), fresh_name)
+
+        # Rename the variable in the copy context
+        rename_in_context(copy_context, original_name, fresh_name)
+
+        original_context = original_context.with_var(fresh_name, copy_context.type)
+        copy_context = copy_context.prev
+
+    return original_context
 
 class LiquidLayer(lark.visitors.Interpreter):
-    def __init__(self, layer_identifier: str= "liquid"):
[…]

### 2023-03-24 code/LayeredTypes/layer_implementations/rowcols.py

@@ -3,7 +3,7 @@ from compiler.interpreters import LiquidTypeChecker
 def depends_on():
     return {"rows","cols"}
 def typecheck(tree):
-    liquid_checker = LiquidTypeChecker.LiquidLayer(layer_identifier="rowcols")
+    liquid_checker = LiquidTypeChecker.LiquidLayer(layer_identifier="rowcols", additional_contexts={"rows", "cols"})
     liquid_checker.visit(tree)
 
     return tree

### 2023-03-24 code/LayeredTypes/layer_implementations/rows.py

@@ -1,7 +1,7 @@
 from compiler.interpreters import LiquidTypeChecker
 
 def typecheck(tree):
-    liquid_checker = LiquidTypeChecker.LiquidLayer(layer_identifier="cols")
+    liquid_checker = LiquidTypeChecker.LiquidLayer(layer_identifier="rows")
     liquid_checker.visit(tree)
 
     return tree

### 2023-03-27 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py
* Correct variable renaming in refinements when combining types

@@ -195,8 +195,10 @@ class LiquidLayer(lark.visitors.Interpreter):
             # We combine the contexts
             ctx = combine_contexts(ctx, other_ctx, refined_type)
 
+            # We rename the variables in the refinement of the other layer
+            other_refinement = substitution_in_liquid(other_type.refinement, LiquidVar(refined_type.name), other_type.name)
             # In the end we combine the refined predicate with the other predicate
-            refined_type.refinement = LiquidApp("&&",[refined_type.refinement, other_type.refinement])
+            refined_type.refinement = LiquidApp("&&",[refined_type.refinement, other_refinement])
         
         # We add a layer annotation to the tree that can be used by other liquid layers
         tree.add_layer_annotation(self.__layer_identifier, "liquid", "type", refined_type)

### 2023-03-28 code/LayeredTypes/compiler/Compiler.py
* Fixed some bugs in combining contexts

@@ -123,9 +123,14 @@ class LayeredCompiler:
                 # We mark the node as done, so we can mark all nodes that depend on it as blocked
                 topological_sorter.done(layer_id)
                 while topological_sorter.is_active():
+                    processed_one = False
                     for node in topological_sorter.get_ready():
+                        processed_one = True
                         self.layer_states[node] = LayerVerificationState.BLOCKED
                         topological_sorter.done(node)
+
+                    if not processed_one:
+                        break
 
         # All layers that are still unprocessed are blocked because they depend on a layer that is part of a cycle
         for layer_id in self.layer_states:

### 2023-03-28 code/LayeredTypes/compiler/interpreters/LiquidTypeChecker.py

@@ -24,11 +24,16 @@ class LiquidSubtypeException(TypecheckException):
         right_type -- the right type of the subtype check
         context -- the context in which the subtype check failed
     """
-    def __init__(self, msg, type_actual, type_expected, context, line, column):
+    def __init__(self, msg, type_actual, type_expected, context: TypingContext, line, column):
         self.type_actual = type_actual
         self.type_expected = type_expected
         self.context = context
-        super().__init__(f"{msg}:\n{type_actual} is not a subtype of {type_expected} (Context: {context})", line, column)
+
+        pretty_context = ""
+        while not isinstance(context, EmptyContext):
+            pretty_context += f"{context.name} : {context.type}\n"
+            context = context.prev
+        super().__init__(f"{msg}:\n{type_actual} is not a subtype of {type_expected}\nIn Context:\n{pretty_context}", line, column)
         pass
 
 class LiquidTypeUndefinedError(TypecheckException):
@@ -114,43 +119,44 @@ def rename_in_context(context: TypingContext, original_name, new_name):
     context.type.refinement = substitution_in_liquid(context.type.refinement, LiquidVar(new_name), original_name)
 
     rename_in_context(context.prev, original_name, new_name)
+
+def fresh_in_both(context_1: TypingContext, context_2: TypingContext):
+    original_name = context_1.name
+    fresh_name = original_name
+
+    while context_1.type_of(fresh_name) or context_2.type_of(fresh_name):
+
+        if context_1.type_of(fresh_name):
+            fresh_name = make_name_unique(fresh_name, context_1)
+
+        if context_2.type_of(fresh_name):
+            fresh_name = make_name_unique(fresh_name, context_2)
+
+    return original_name, fresh_name
+
+
+def combine_recursive(rec_ctx_1: TypingContext, rec_ctx_2: TypingContext):
+    if isinstance(rec_ctx_2, EmptyContext):
+        return rec_ctx_1, []
+
+    # Start with the recursive call, since we want to rename the variables in the copy context
+    rec_ctx_1, renamings = combine_recursive(rec_ctx_1, rec_ctx_2.prev)
+
+    if isinstance(rec_ctx_2, VariableBinder):
+        original_name, fresh_name = fresh_in_both(rec_ctx_2, rec_ctx_1)
+
+        rec_ctx_2.name = fresh_name
+        rec_ctx_2.type.refinement = substitution_in_liquid(rec_ctx_2.type.refinement, LiquidVar(fresh_name), original_name)
+
+        for name, renaming in renamings:
[…]

### 2023-03-28 code/LayeredTypes/examples/python_impls/use_case_2.py

@@ -1,13 +1,37 @@
 from random import randrange
 from builtins import len, print
 
-def set_list_element(ls, idx, val):
-    ls[idx] = val
-    return ls
+def create_dataset(num_cols, num_rows):
+    dataset = []
+    for i in range(num_rows):
+        row = []
+        for j in range(num_cols):
+            row.append(randrange(100))
+        dataset.append(row)
+    return dataset
 
-def get_element(ls, idx):
-    return ls[idx]
+def add_rows(dataset, num_rows):
+    for i in range(num_rows):
+        row = []
+        for j in range(len(dataset[0])):
+            row.append(randrange(100))
+        dataset.append(row)
+    return dataset
 
-def get_list():
-    # Return list with 12 random elements
-    return [randrange( 100) for _ in range(12)]
+def add_cols(dataset, num_cols):
+    for i in range(len(dataset)):
+        for j in range(num_cols):
+            dataset[i].append(randrange(100))
+    return dataset
+
+# Reduce the number of cols in the dataset to num_cols
+def reduce_cols(dataset, num_cols):
+    for i in range(len(dataset)):
+        dataset[i] = dataset[i][:num_cols]
+    return dataset
+
+def sample(dataset, num_rows):
+    return dataset[:num_rows]
+
+def train(dataset):
+    print("Training on dataset of size: ", len(dataset), "x", len(dataset[0]))

### 2023-03-29 tex/chapters/use-cases.tex
* Added description of use-case 2

--- 2023-01-30 paper/use-cases.tex
@@
+%! TEX root = ../acmpaper/paper.tex
 \section{Example Use-Cases}
@@
 \label{sssec:use_case_2}
+
+In our second use-case we use a scenario where we verify different properties of a program based on liquid types.
+Specifically, we show how our approach can be used to improve the verification of orthogonal properties.
+
+\subsubsection{Description}
+\label{sssec:uc2_description}
+
+Liquid Types can be particulary useful when operating with machine learning pipelines: Learning datasets usually contain many dimensions that can often be manipulated independently from one another.
+In its' simplest form, a dataset can be compared to a two-dimensional table where we can add and remove rows or columns independently from one another.
+However some operations may require interaction of these two dimensions.
+For example, in our two dimensional running example we could consider each column a specific feature and each row a specific datapoint.
+If we now want to train a model based on a specific dataset, we might want to ensure that we have enough datapoints in relation to the number of features we observe.
+
+For our example use-case we model some basic operations that might be used in a machine learning environment to create and modify datasets. Specifically, we will support the following operations:
+\begin{itemize}
+	\item \texttt{createDataset(c,r)} -- Creates a dataset with a specific number of rows and columns
+	\item \texttt{add\_rows(d,n)} -- Adds a specific number of rows to a dataset
+	\item \texttt{add\_cols(d,n)} -- Adds a specific number of columns to a dataset
+	\item \texttt{reduce\_cols(d,n)} -- Reduces a datset to a specific number of columns
+	\item \texttt{sample(d,n)} -- Samples a specific number of datapoints from a dataset
+	\item \texttt{train(d)} -- Trains a model based on a specified dataset
+\end{itemize}
+
+From the operations above, it is already clear that these operations will affect rows and columns of a dataset.
+Sometimes independently from one another, but some operations will also require or provide information about both of them.
+We will describe these changes in dimensions using liquid types.
+Under the assumption that the in-built function \texttt{n\_rows(d)} and \texttt{n\_cols(d)} return the number of rows and columns of a dataset respectively, these function have the following liquid signatures:
+
+\begin{lstlisting}[numbers=none,tabsize=8,caption={Liquid Signatures of functions in Use-Case 2}]
+createDataset :: { c:Int | (c >= 0) }
+	-> { r:Int | (r >= 0) }
+	-> { d:DataSet | (n_cols(d) == c) && (n_rows(d) == r) }
+
+add_rows :: { d:DataSet | n_cols(d) > 0 }
+	-> {n:Int | n>0} 
+	-> { d2:DataSet | (n_rows(d2) == (n_rows(d) + n)) && (n_cols(d2) == n_cols(d)) }
+
+add_cols :: { d:DataSet | true } 
+	-> { n:Int | n>0 }
+	-> { d2:DataSet | (n_cols(d2) == (n_cols(d) + n)) && (n_rows(d2) == n_rows(d)) }
+
+reduce_cols :: { d:DataSet | true } 
+	-> { c:Int | (c >= 0) && (c < (n_cols(d))) }
+	-> { d2:DataSet | (n_cols(d2) == c) && (n_rows(d2) == n_rows(d)) }
+
+sample :: { d:DataSet | true } 
+	-> { n:Int | (n >= 0) && (n <= (n_rows(d))) } 
+	-> { d2:DataSet | (n_rows(d2) == n) && (n_cols(d2) == n_cols(d)) }
+
+train :: { d:DataSet | (n_rows(d) >= 10 * (n_cols(d))) } -> { m:Model | true }
+\end{lstlisting}
+
+When inspecting the liquid signatures above, we notice that only in |train| the number of rows directly interacts with the number of columns.
+For all other functions these two properties are orthogonal to each other.
+Current liquid type implementations are limited to this combined verification of these properties for all cases.
+However in \Cref{sssec:uc2_comparison} we will showcase the challenges of this approach and how our implementation can lift this limitation.
+
+For our use-case we will look at three individual code snippets: One that only uses operations that manipulate the rows of a dataset, one that only manipulates columns and one that depends on both dimensions by using \texttt{train}.
+The three code snippets are show in \Cref{lst:uc2_code1,lst:uc2_code2,lst:uc2_code3}.
+A more in-depth explanation of the exmaples will follow in \Cref{sssec:uc2_comparison}.
+
+\begin{minipage}{.45\textwidth}
+\begin{lstlisting}[caption={Code example that only manipulates rows},label={lst:uc2_code1}]
+let empty := create_dataset(10,0) in {
+    let d := add_rows(empty, 50) in {
+        sample(d, 10)
+    }
+}
+\end{lstlisting}
+\end{minipage}\hfill%
+\begin{minipage}{.45\textwidth}
+\begin{lstlisting}[caption={Code example that only manipulates columns},label={lst:uc2_code2}]
+let d := create_dataset(10,50) in {
+    let d2 := add_cols(d, 100) in {
+        reduce_cols(d2, 20)
+    }
+}
+\end{lstlisting}	
+\end{minipage}
+
+\begin{center}
+\begin{minipage}{0.5\textwidth}
+\begin{lstlisting}[caption={Caption that uses both rows and columns},label={lst:uc2_code3}]
+let empty_d := create_dataset(10,0) in {
+    let d := add_rows(empty_d, 50) in {
+       let column := reduce_cols(d, 1) in {
+           train(column)
+       }
+
+       let small_d := reduce_cols(d, 5) in {
+           train(small_d)
+       }
+    }
+}		
+\end{lstlisting}
+\end{minipage}
+\end{center}
+\subsubsection{Implementation Details}
+\label{sssec:uc2_implementation}
+
+\subsubsection{Comparison of monolithic to independent layers}
+\label{sssec:uc2_comparison}
+
+\subsubsection{Discussion}
+\label{sssec:uc2_discussion}
+
+\subsubsection{Limitations}
+\label{sssec:uc2_limitations}

### 2023-03-29 tex/chapters/use-cases.tex
* Finished technical description of implementation of use-case 2

@@
 
-\begin{lstlisting}[numbers=none,tabsize=8,caption={Liquid Signatures of functions in Use-Case 2}]
+\begin{lstlisting}[numbers=none,tabsize=8,caption={Liquid Signatures of functions in Use-Case 2},label={lst:uc2_liq_sig}]
 createDataset :: { c:Int | (c >= 0) }
@@
 
+Similar to the layers implemented for Use-Case 1, we implemented the verification layers based on the interpreter classes provided by \Lark.
+However, there is a slight difference: While the layers in our previous use-case had a static, predefined name, our liquid layer uses a more generic approach where the user can choose the layer name.
+This was an explicit design decision as it makes it easy to re-use the same liquid verification implementation to verify different liquid properties independently (See \Cref{sssec:uc2_comparison}).
+
+As a basis for the liquid type checking we use our own fork of the \textsc{Aeon} language \footnote{See \url{https://github.com/LuAbelt/aeon-lt/tree/aeon-lt}}.
+Therefore, we use the same type definitions for liquid types as \textsc{Aeon} does.
+An example of this annotation style (excluding the layer identifier) can be seen in \Cref{lst:uc2_liq_sig}.
+It is possible to annotate both variable identifiers and function calls in our liquid layer.
+In a program, the verifier will perform a liquid type check for two specific scenarios:
+\begin{itemize}
+	\item When assigning an expression to a variable as part of a \texttt{let} statement
+	\item When passing an expression as an argument to a function call
+\end{itemize}
+
+Consider the following example:
+
+\begin{lstlisting}[caption={Simple example code of a liquid layer},label={lst:uc2_simple_example}]
+var :: liquid :: { v:Int | v>0 }
+fun :: liquid :: { arg:Int | arg>5 } -> {ret:Int | ret>=arg }
+
+let var := fun(10) in {
+	var
+}
+\end{lstlisting}
+
+In this example two liquid type checks happen.
+First, it is checked, whether the integer literal \texttt{10} is a subtype of the defined type for the first argument of the function \texttt{fun}, which requires an argument greater than $5$.
+Internally, \textsc{Aeon} performs this typecheck by employing the constraint solver \textsc{Z3}, which effectively checks the following:
+
+$$\forall v:Int | (v==10)\implies(v>5)$$
+
+Since in liquid types, there can be cross-references to previously defined variables, to be able to verify this properly our layer also builds a \textit{typing context}(For brevity simply referred to as context from here on).
+This context contains the specific types for all previously defined and potentially referenced identifiers.
+This context will be extended by each \texttt{let} and each function call that is verified.
+
+One thing to note about the context is, that when our liquid verification layer adds types to it, it will always add the most specific type that is known.
+For example, in the code snippet above, within the body of the \texttt{let} statement, we know that \texttt{var} will always be greater than $10$, even though purely by the liquid signature of \texttt{fun} we would only know that it always has to be greater than $5$.
+However, by adding the specific type of the argument passed to \texttt{fun} to the context we have a more precise information available.
+
+The current context as well as resulting type of a node is also attached as additional meta-information to the AST.
+With this, our liquid layer implementation allows a novel technique that differentiates it from current implementations available: With this we can not only verify liquid properties independently from one another, but also allow later liquid layers to use type information provided by a previous liquid layer to be taken into account when type checking a joint layer.
+One usage of this will be shown in \Cref{sssec:uc2_comparison}.
+To allow this interaction with other liquid layers, the user simply provides a set of layer identifiers that will then be queried for additional type and context information.
+
 \subsubsection{Comparison of monolithic to independent layers}
@@
 \label{sssec:uc2_limitations}
+- Divide into conceptual and technical limitations
+
+- Conceptual:
+	- User needs to properly divide properties
+
+- Technical
+	- Scalability: Context can grow fast (Could be fixed)
+	- Function Defs are not checked against defined return type (Could be fixed)
+	- All identifies/functions needs to be annotated even if layer does not add/need information
+	(Might be fixable?)
+	- Also irrelevant vars are added to context
+	- No "normal" assign

### 2023-03-30 tex/chapters/implementation.tex
Restructured LaTeX directories
- Now represents individual submissions/styles
* Finished description of use-case 2

--- 2023-01-30 paper/implementation.tex
@@
 The language is similar to most simple functional languages.
-Implicitly, our language natively supports \textt{Boolean} and \texttt{Integer} datatypes through the available constants.
+Implicitly, our language natively supports \texttt{Boolean} and \texttt{Integer} datatypes through the available constants.
 However, the language on it's own does not require any type annotations.
@@
 Each layer will be checked individually and independently.
-A more detailled approach of the architecture behind our layered approach will be described in section \ref{ssec:architecture}.
+A more detailled approach of the architecture behind our layered approach will be described in \Cref{ssec:architecture}.
 
-An example how a program written in our language could look like can be seen in \ref{lst:fac_example}.
+An example how a program written in our language could look like can be seen in \Cref{lst:fac_example}.
 The program defines a factorial function that is refined in two layers.

### 2023-03-30 tex/chapters/introduction.tex
* Finished description of use-case 2

@@
 
-One example for such a refinement in LiquidHaskell is shown in \ref{lst:combined_lt}.
+One example for such a refinement in LiquidHaskell is shown in \Cref{lst:combined_lt}.
 There, we define a function that extracts the maximum value of a (descendingly) ordered, non-empty list.
@@
 Effectively one could treat such different requirements as separate \textit{Layers} that can be verified independently.
-Additionally with a proper implementation using such layers may also result in an implementation benefit, that different Layers can be defined individually and combined arbitrarily, as opposed to the example in Listing \ref{lst:combined_lt} where the combination of the two traits had to be explicitly defined as an refinement on it's own.
+Additionally with a proper implementation using such layers may also result in an implementation benefit, that different Layers can be defined individually and combined arbitrarily, as opposed to the example in Listing \Cref{lst:combined_lt} where the combination of the two traits had to be explicitly defined as an refinement on it's own.
 
 One other possible usage of such an iterative approach is that it could allow to more flexibly add different static analysis approaches on top of each other.
-For example \ref{handley2020} have shown how to use liquid types in Haskell to statically perform cost analysis of different resources such as memory consumption or computational complexity.
+For example \cite{liquidate-assets} have shown how to use liquid types in Haskell to statically perform cost analysis of different resources such as memory consumption or computational complexity.
 In their approach they showed that liquid types can be used to reason about their upper and lower bounds.

### 2023-03-30 tex/chapters/use-cases.tex
* Finished description of use-case 2

@@
 
-\subsubsection{Discussion}
+In the previously described examples, all the code presented was valid and consistent with the liquid signatures we described earlier.
+Thus, verifying them all within the same verification pass does not bear any issues.
+An examples of slightly adapted versions of the code are shown in \Cref{lst:uc2_code1_fail,lst:uc2_code2_fail,lst:uc2_code3_fail} that would lead to the verification errors shown in \Cref{lst:uc2_errors_monolithic}.
+For this simple code, even without profound domain knowledge it is relatively easy to spot that the errors are caused by the rows or columns respectively.
+The only error that might be slightly harder to interpret is the error caused by \texttt{train} since it is caused by an interaction of the two properties.
+
+\begin{minipage}{.45\textwidth}
+\begin{lstlisting}[caption={Adapted version of \Cref{lst:uc2_code1}},label={lst:uc2_code1_fail}]
+let empty := create_dataset(10,0) in {
+    let d := add_rows(empty, 50) in {
+        sample(d, 100)
+    }
+}
+\end{lstlisting}
+\end{minipage}\hfill%
+\begin{minipage}{.45\textwidth}
+\begin{lstlisting}[caption={Adapted version of \Cref{lst:uc2_code2}},label={lst:uc2_code2_fail}]
+let d := create_dataset(10,50) in {
+    reduce_cols(d, 20)
+}
+\end{lstlisting}	
+\end{minipage}
+
+\begin{center}
+\begin{minipage}{0.5\textwidth}
+\begin{lstlisting}[caption={Adapted version of \Cref{lst:uc2_code3}},label={lst:uc2_code3_fail}]
+let empty_d := create_dataset(10,0) in {
+    let d := add_rows(empty_d, 50) in {
+       train(d)
+    }
+}
+\end{lstlisting}
+\end{minipage}
+\end{center}
+
+\begin{lstlisting}[caption={Error messages when using one monolithic liquid layer},label={lst:uc2_errors_monolithic},numbers=none]
+-- Error in sample:
+Layer 'liquid' failed with error: Error in function call to sample, argument 2:
+{ v:Int | (v == 100) } is not a subtype of { n:Int | ((n >= 0) && (n <= n_rows(sample_d))) }
+[...]
+
+-- Error in reduce_cols:
+Layer 'liquid' failed with error: Error in function call to reduce_cols, argument 2:
+{ v:Int | (v == 20) } is not a subtype of { c:Int | ((c >= 0) && (c < n_cols(reduce_cols_d))) }
+[...]
+
+-- Error in train:
+Layer 'liquid' failed with error: Error in function call to train, argument 1:
+{ d2:DataSet | ((n_rows(d2) == (n_rows(add_rows_d) + add_rows_n)) && (n_cols(d2) == n_cols(add_rows_d))) } is not a subtype of { d:DataSet | (n_rows(d) >= (10 * n_cols(d))) }
+[...]
+\end{lstlisting}
+
+However, there are also very basic examples, where such a distinction is not easily possible anymore.
+Consider the following example, where we have a dataset stored in a variable on which we want to both ensure a certain number of rows and a number of columns:
+
+\begin{lstlisting}[caption={Small example of a dataset imposing restrictions on both rows and dimensions},label={lst:uc2_fail_monolithic}]
+d :: liquid :: { d:DataSet | (n_rows(d)>100) && (n_cols(d)>10)}
+
+let d := createDataset(10,1000) in {
+	d
+}
+\end{lstlisting}
+
+In this example, the programmer might have overlooked that the amount of columns needs to be greater than $10$.
+The resulting verification error is shown in \Cref{lst:uc2_error_mono}.
+For the developer, it is not directly evident that the error is caused by an issue with the number of rows.
+Only after careful inspection of the exact error and verification context, it is possible for the developer to properly identify this.
+
+\begin{lstlisting}[caption={Verification error produced by \Cref{lst:uc2_fail_monolithic}}]
+Layer 'liquid' failed with error: Error during assigning identifier 'd' in let statement:
+{ d:DataSet | ((n_cols(d) == create_dataset_c) && (n_rows(d) == create_dataset_r)) } is not a subtype of { d:DataSet | ((n_rows(d) > 100) && (n_cols(d) > 10)) }
+In Context:
+create_dataset_r : { v:Int | (v == 1000) }
+create_dataset_c : { v:Int | (v == 10) }
+(line 3)
+\end{lstlisting}
+
+Our implementation allows us to separate these individual checks into different verification passes. For the running example, we will consider three separate layers:
+
+\begin{enumerate}
+	\item \texttt{rows} -- This layer is only verifies all information that is only relevant to the number of rows of a dataset
+	\item \texttt{cols} -- This layer only verifies information only relevant to the number of columns of a dataset
+	\item \texttt{rowcols} -- This is a layer that verifies all \textit{interactions} between rows and columns of a dataset. It depends on the previous two layers and re-uses their contexts
+\end{enumerate}
+
+With this approach, each function and identifier will get three separate layer annotations, one for each respective layer.
+An example of how these annotations look for \texttt{create\_dataset} and \texttt{train} are shown in \Cref{lst:uc2_separate_liq_sigs}\footnote{For brevity, we omitted further examples for the other methods.
+However, they can be found in our GitHub repository}.
+
+\begin{lstlisting}[numbers=none,caption={Liquid signatures for separate layers},label={lst:uc2_separate_liq_sigs},tabsize=4]
+-- Only define requirements and effect on rows
+create_dataset :: rows :: { c:Int | true } -> { r:Int | (r >= 0) } 
+			-> { d:DataSet | (n_rows(d) == r) }
+
+-- Only define requirements and effect on columns
+create_dataset :: cols :: { c:Int | (c >=0) } -> { r:Int | true } 
+			-> { d:DataSet | (n_cols(d) == c) }
+
+-- No interaction between layers
+create_dataset :: rowcols :: { c:Int | true } -> { r:Int | true } 
+			-> { d:DataSet | true }
+
+-- train does not have properties that only affect rows or column independently
+train :: rows :: { d:DataSet | true } -> { m:Model | true }
+train :: cols :: { d:DataSet | true } -> { m:Model | true }
+-- Check the interaction
+train :: rowcols :: { d:DataSet | (n_rows(d) >= 10*(n_cols(d))) } -> { m:Model | true }
+\end{lstlisting}
+
+Similarly, the example from \Cref{lst:uc2_fail_monolithic} slightly changes to the code shown in \Cref{lst:uc2_fail_separate}.
+With this new approach the verifier output changes as shown in \Cref{lst:uc2_error_separate}.
+It now provides with a much more accurate information from which we know, that all properties related to the number of rows could be verified but there was an error during verifying the \texttt{cols} layer.
+Subsequently, the \texttt{rowcols} layer is blocked for verification.
+Additionally we get a more concise error message for the verification error than we had when only using one monolithic layer.
+\begin{lstlisting}[caption={Adapted version of \Cref{lst:uc2_fail_monolithic} that uses separate layers},label={lst:uc2_fail_separate}]
+d :: rows :: { d:DataSet | (n_rows(d)>100)}
+d :: cols :: { d:DataSet | (n_cols(d)>10)}
+d :: liquid :: {d:DataSet | true }
+
+let d := createDataset(10,1000) in {
+	d
+}
+\end{lstlisting}
+\begin{lstlisting}[caption={Verifier output produced by \Cref{lst:uc2_fail_separate}},label=lst:uc2_error_separate,numbers=none]
+Not all layers could be verified.
+The following layers were processed:
+	rows: SUCCESS
+	cols: FAILURE
+	rowcols: BLOCKED
+
+The following layers failed during typechecking:
+	cols:
+	 Layer 'cols' failed with error: 20:1: Error during assigning identifier 'd' in let statement:
+{ d:DataSet | (n_cols(d) == create_dataset_c) } is not a subtype of { d:DataSet | (n_cols(d) > 10) }
+In Context:
+create_dataset_r : { v:Int | (v == 1000) }
+create_dataset_c : { v:Int | (v == 10) }
+\end{lstlisting}
+
+Due to the spearation we now have much more detailled information on what exactly causes the liquid error. For a programmer this can make it much simpler to debug programs and fix liquid errors.
+
+\subsubsection{Discussion and Limitations}
 \label{sssec:uc2_discussion}
 
-\subsubsection{Limitations}
-\label{sssec:uc2_limitations}
-- Divide into conceptual and technical limitations
+In this second use-case, we have successfully shown:
+\begin{itemize}
+	\item That our approach can be used to verify liquid types in general by providing a general liquid verification layer
+	\item Furthermore we provided a small demonstration how the same implementation can be used to separate different concerns of liquid types into separate independent verification passes
+	\item Lastly our implementation enables the re-use of type information provided by a liquid layer previously verified.
+\end{itemize}
 
-- Conceptual:
-	- User needs to properly divide properties
+Especially for the separation of concerns we believe that this presents a novelty in the domain of liquid types as we are not aware of similar work in this field.
+However, we also have to discuss certain limitations that our implementation and approach may exhibit.
+We divide this limitations into conceptual and technical limitations.
 
-- Technical
-	- Scalability: Context can grow fast (Could be fixed)
-	- Function Defs are not checked against defined return type (Could be fixed)
-	- All identifies/functions needs to be annotated even if layer does not add/need information
-	(Might be fixable?)
-	- Also irrelevant vars are added to context
-	- No "normal" assign
+\paragraph{Conceptual Limitations} -- While the separation of concerns for liquid properties presents a powerful mechanism, it also exposes a certain risk.
+When separating the concerns, the developer needs to have a clear understanding if and how properties of the affected functions and variables may affect each other or interact with one another.
+Meaning that the developer performing the separation into the different layers requires a profound domain knowledge of the system to be verified.
+
+If this separation does not performed properly this could lead to undesired side effects.
+One example of of an unproper separation could happen when a developer decides to separate two properties in the assumption that they are independent if in reality they are not.
+In that case the two properties might lead to a valid program when verified independently drom one another, but when verified in conjunction one might realize that there are flaws in the system.
+Another example are properties that contradict each other.
+
+This issue is related to correctly identifying the dependencies and interactions between properties.
+For a sufficient large software systems it might not be trivial to identify the dependencies and interactions between different parts of the system.
+Similar findings have been made in the domain of Software-Product Line Engineering when identifying interactions between features\cite{apel2013}.
+
+\paragraph{Technical Limitations} Our prototype exhibits some technical limitations that might reduce the applicability of our implementation on bigger systems.
+First and foremost, since we limit ourselves to the small example language that we designed ourselves, using it on real-world examples would require and considerable amount of work to develop a proper wrapper with our language and the built-in python integration.
+
+Additionally, our prototype only allows to bind variables for a specific scope via the \texttt{let} statement.
+Reassigning an existing variable to a new value is not possible.
+Additionally, all the identifiers and functions need to be fully annotated even if the specific layer does not require or add any new information about them (As seen in for example \texttt{train} for the layers \texttt{rows} and \texttt{cols} in \Cref{lst:uc2_separate_liq_sigs}).
+However, we believe that this design decision is a more safer one than allowing unannotated identifiers that are simply not verified, as it adds an additional level of explicitness to the verification process.
+However, with many layers this leads to a considerable amount of additional boilerplate code that might be required.
+
+Lastly, our implementation currently also might suffer from a general scalability issue, when using the feature to consider additional contexts during verification.
+During verification the current context is merged with all the dependent contexts.
+This merged context is then also used for the further verification steps.
+However the next time this new context is merged again with the dependent contexts, there is no mechanism in place that checks whether the variables from the additional contexts have already been merged.
+This leads to multiple independent copies of the same variables (With varying names) existing in the context.
+
+This shortcoming of our implementation however could be fixed with little implementation effort.
+One simply needs to keep track which variables have already been added to the context previously and what they might have been renamed to.
+When then merging the contexts a second time, already existing variables would simply be skipped.
+However, due to limited time this feature is not yet present in our prototype.

### 2023-03-30 tex/chapters/conclusion.tex
* Finished evaluation and conclusion for acm paper

\section{Conclusion}
\label{sec:conclusion}

In this work, we evaluated the prossibility of a novel verification approach called that can combine different dependent and independent type system in one verification framework. We have presented \textsc{LayeredTypes} -- A framework that realizes this verification approach on a simple functional language.

Using two distinct use-cases, we have shown how this approach can be used for various scenarios, including an exploration to the improvement of current liquid type implementations. Our implementation provides a scientific demonstrator for a layered and incremental verification approach. We believe this novel verification approach could exhibit great potential when building complex software systems. We hope that this work, alongside with our implementation, can provide valuable insights on the usefulness and potential challenges of such an approach.

### 2023-03-30 tex/chapters/evaluation.tex
* Finished evaluation and conclusion for acm paper

\section{Evaluation}
\label{sec:evaluation}

In this section we will critically to evaluate our implemented approach and in which capacity we where able to answer our research questions. In RQ1, we aksed whether it is possible for us to define a layered type system that can be used to verify different properties independently from one another.

We believe that our approach has successfully shown that you can build an implementation that flexibly allows to verify independent properties of different type systems independently from each other. In particular, use case 1 has shown an applied example of this layered type system. However, in this work we do not present a formalized description of such a layered type system. This might be particulary relevant when building and formalizing different type layers that depend and interact with each other.

Additionally, the system we built is rather limited; While our language supports all basic constructs of a programming language which allows it to also build complex functionality, we are limited to this simple syntax. In order to apply our layered verification approach to already exisiting software a non-insignificant integration effort might be required. Therefore our implementation should rather be viewed as a stand-alone prototype that merely evaluates the theoretical possibilities and potential limits of such a system.

For RQ2, which focusses on how and if this approach can be used to improve current liquid type implementations, we have achieved mixed results: One the one hand, we have successfully shown in use-case 2 that we can built an implementation that can achieve this on a small scale. However, the exact implementation of this has proven rather complicated. Due to limited time we were not able to evaluate the implementation on a more complex example which might potentially unveil further implementation issues and bugs.

Additionally, our implementation faces similar scalability challenges as the ones we also face for RQ1, when trying to take this approach to real-world software. Lastly, our research question specifically asked whether our approach improves challenges that are present in current Liquid Type implementations. We believe that our approach can improve a few challenges from liquid types, mainly identifying the source of a liquid type errors by cleanly dividing the responsibilities into separate verification passes. However we also note that this division itself might be non-trivial and required advanced domain knowledge.

Lastly, our layered approach does not improve the interpretability of the liquid error messages itself. When considering the example of \Cref{lst:uc2_code3_fail}, when verified with separated layers, the error message still gets very verbose and hard to interpret as shown in \Cref{lst:uc2_code3_error}. One reason for this is certainly the issue already mentioned in \Cref{sssec:uc2_discussion}, that our context currently grows faster than it needs. We therefore think that further work needs to be done to improve liquid error messages itself.

\begin{lstlisting}[caption={Error message for \Cref{lst:uc2_code3_fail}},label={lst:uc2_code3_error}]
Layer 'rowcols' failed with error: 24:5: Error in function call to train, argument 1:
{ d:DataSet | ((((true && (n_cols(d) == create_dataset_c)) && (n_rows(d) == create_dataset_r)) && (n_cols(d) == create_dataset_c)) && (n_rows(d) == create_dataset_r)) } is not a subtype of { d:DataSet | (n_rows(d) >= (10 * n_cols(d))) }
In Context:
[... 12 lines of context omitted ...]
\end{lstlisting}
